### 2024-10-30 Emily_exploring_data.Rmd

    Working with Git Project:
    1. Make sure project is set to D445_Project in RStudio
    2. In RStudio Console:
        getwd() = Verify working directory is set to this project
        setwd() = Can change working directory to this path
        list.files() = Can verify the files are available
    3. In RStudio Terminal:
        cd <destination> = Change directory
        cd - = Go back to previous directory
        ls = Lists contents of current directory
        git init = Initialize git connection
        git status = Check what status needs to be changed before comitting
          - git  add <file>
          - git rm <file>
          - git restore <file>
        git commit -m "" = Commit changes from terminal with message
    4. Might need to open GitHub Desktop and verify changes. Then, click "push origin".

```{r}
#############
# Libraries #
#############

```

Creating dfs for the price, number of daily bike trips, and the number of daily scooter trips taken over our entire data period.

```{r}
data_dir <- 'FOCO Spin Data'

df_daily_scooter_trips <- data.frame(read.csv(file.path(data_dir,"scooters 2021-07-26--2024-10-17-analyze-trips_by_date.csv")))
df_daily_bike_trips <- data.frame(read.csv(file.path(data_dir,"bikes 2021-07-26--2024-10-17-analyze-trips_by_date.csv")))
df_hourly_price_rate <- data.frame(read.csv(file.path(data_dir,"spin-prices.csv")))

# head(df_daily_scooter_trips)
# head(df_daily_bike_trips)
# head(df_hourly_price_rate)
```

### 2024-11-12 Code Files/FoCo_HeatMapping.Rmd

This is where I will be exploring the geological start and end locations for the spin cycle datasets provided for our research project. The goals for this portion of the project are:

1.  Use latitude and longitude data to develop a heat map for spin scooter start and end locations.

```{r}
#libraries
library(tidyverse)
library(tidymodels)
library(sf)
library(ggplot2)
library(ggmap)
library(RColorBrewer)
library(patchwork)
library(here)
```

1. Explore the data.

a)  Load the starting and ending latitude and longitude coordinates into respective data-frames.

```{r}
data_dir <- 'FOCO Spin Data'
start_scoot <- data.frame(read.csv(file.path(data_dir,'fort-collins-curb-events-2021-07-26-2024-10-15-trip_starts.csv')))
end_scoot <- data.frame(read.csv(file.path(data_dir,'fort-collins-curb-events-2021-07-26-2024-10-15-trip_ends.csv')))

head(start_scoot)
head(end_scoot)
```

```{r}
range(start_scoot$latitude)
range(start_scoot$longitude)

range(end_scoot$latitude)
range(end_scoot$longitude)

end_scoot |> arrange(desc(longitude))
```

b)  can we create a map for our data?

```{r}
register_stadiamaps('dd6906c8-64ba-46a6-966a-4a3556ae9e34',write = TRUE)
```

```{r}
foco_map_bounds <- c(left = -105.15, bottom = 40.47, right = -104.97, top = 40.65)

foco_map <- get_stadiamap(foco_map_bounds, zoom = 13, maptype = 'stamen_toner_lite')

foco_coord_map <- ggmap(foco_map, extent = 'device', legend = 'none')
```

2. can we generate a heatmap for our data?

```{r}
start_map <- foco_coord_map + geom_point(data = start_scoot, aes(x = longitude, y = latitude), alpha = 0.15, size = 0.8, color = 'orange') + theme(panel.grid = element_blank())

end_map <- foco_coord_map + stat_density2d(data= end_scoot,  aes(x=longitude, y=latitude, fill=..level.., alpha=..level..), geom="polygon", contour = TRUE, bins = 15, size = 0.2) + scale_fill_gradientn(colors = rev(brewer.pal(9, 'Spectral'))) +  scale_alpha(range = c(0.2, 0.7)) + geom_point(data = end_scoot, aes(x = longitude, y = latitude), alpha = 0.2, size = 0.5)

start_dens_heatmap <- foco_coord_map + stat_density2d(data= start_scoot,  aes(x=longitude, y=latitude, fill=..level..), geom="polygon", contour = TRUE, bins = 20, h = c(0.012, 0.012)) + scale_fill_gradientn(colors = rev(brewer.pal(9, 'PuOr')), name = 'Density') + theme(panel.grid = element_blank())

start_dens_heatmap + start_map + plot_layout(ncol = 2, guides = 'collect')
```
d) this map provides a basic layout of fort collins. perhaps there is a way to enhance the mapping of our data a little more. (outline key features, form boundaries, draw lines at major roads, etc.)

e) This is a good start to my heatmap but the data feels too separated from each other. There is probably a way to base the density bandwidth on the points in our model. We need our model to base density off of the start points nearest to it instead of the overall density of the data. Right now the density model is saying "I'm in a popular location because I live in Fort Collins." we need each point in the data to say instead "I'm in a popular location because theres a lot of other scooters who started/ended their trip nearby me."

  After doing some research on the topic I'm led to believe that incorporating a KNN model to our data could be very useful here...

```{r}
start_split <- initial_split()

training_start_data <-
test_start_data <-

knn1_spec <- nearest_neighbor(mode = "classification", neighbors = 1)

knn1_spec |>
  fit(usage ~ latitude + longitude, data = training_start_data) -> knn.fit

knn.test_res <- knn.fit |> augment(new_data = test_start_data)

knn.test_res |> 
  conf_mat(truth = Direction, estimate = .pred_class)

knn.test_res |> 
  accuracy(truth = Direction, estimate = .pred_class)
```
references:

building a heatmap using ggmap, stadiamaps, & RColor: <https://axelhodler.medium.com/creating-a-heat-map-from-coordinates-using-r-780db4901075>

nearest neighbor mapping: <https://www.espatial.com/blog/nearest-neighbor-map>

### 2025-11-17 Exploratory_Analysis_Josh_Tobey.Rmd
Test push into the project repository from Josh Tobey.

---
title: "Exploratory_Analysis_Josh_Tobey"
author: "Joshua Tobey"
date: "2025-11-17"
output: html_document
---
```{r}

```

### 2025-11-18 Exploratory_Analysis_Josh_Tobey.Rmd
Second attempt at test push
Second Attempt

@@
 ```{r}
-
+print("Push test for exploratory analysis")
+install.packages(c("usethis", "gitcreds"))
 ```

### 2025-11-21 Data Cleaning.Rmd
Initial Commit

---
title: "Data Cleaning"
author: "Dawson Carney"
date: "2025-11-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(dplyr)
library(tidyr)

# Match dates in alkalinity and hardness data
alk <- read.csv("RWAR_alkalinity.csv")
hardness <- read.csv("RWAR_hardness.csv")

# split DateStamp into day and time
alk_new <- separate(alk, col = DateStamp, into = c("Date", "Time"), sep = " ") |>
  select(-c("Time")) |>
  rename(alkalinity = Alkalinity..mg.L)
hardness_new <- separate(hardness, col = DateStamp, into = c("Date", "Time"), sep = " ") |>
  select(-c("Time")) |>
  rename(hardness = Hardness..mg.L)

# merge by Date
merged <- merge(alk_new, hardness_new, by="Date")

merged

```

### 2025-12-01 Paper Draft.Rmd
Initial commit.

---
title: "Predictive Modeling of Drinking Water Treatment Process Chemical Doses"
author: "Dawson Carney, Reece Carmody, Ethan Schilling, Joshua Tobey"
date: "5 December 2025"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

### 2025-12-01 Paper Draft.Rmd
Completed draft of introduction. Started Data Overview. Created overall tentative outline.

@@
 ## Introduction
+
+### Background and Motivation
+
+[Background on Drinking Water Treatment/Process Overview]
+
+Many different factors indicate effective chemical dose performance. Examples could include:
+
+* chemical information about the water such as turbidity, charge, or pH
+* information on the size of the particles being formed in the flocculation process (found via running tests on water samples)
+* observation of filter performance at the end of the process
+* total chemical byproducts produced
+* output water quality
+
+The chemical processes that determine treatment process performance are complex
+and highly interconnected, therefore all of these factors and more are important
+in making effective decisions about chemical dosing.
+
+***The role of an operator in deciding water treatment plant chemical doses is to
+look at a wide array of these factors and make decisions about how to adjust
+chemical doses.***
+
+Treatment plant conditions can change rapidly, so having tools
+that aid in selecting chemical doses in response to changing water quality is 
+crucial for operators. These chemicals are one of the primary costs of the treatment 
+process, costing millions of dollars per year for a mid-size treatment plant. 
+The chemical byproducts produced by excessive coagulant doses also have environmental
+impacts. Therefore, this is both a financial problem impacting taxpayers, and an
+environmental one.
+
+### Project Purpose
+***The goal of this project is to produce a model that serves the role of a 
+treatment plant operator.*** This model would provide recommended chemical doses,
+based on input water quality characteristics, and how operators have dosed 
+chemicals in the past. This could prove a useful tool to operators as a "starting
+point" in chemical dosing. They could see a change in water quality, retrieve
+the suggested chemical dose from the model, test this dose, and adjust from there
+based on other information.
+
+## Data Overview and Cleaning
+
+The data used for this effort is a timeseries dataset from a Colorado water treatment
+plant covering XX years. This treatment plant takes its water from a reservoir, which
+tends to be a more stable water source than sources such as rivers or industrial supplies.
+
+The available data are as follows, with missing date ranges noted:
+
+* Raw Water Data
+    + pH
+    + Temperature
+    + Turbidity
+    + Suspended Grain Size Information
+    + Alkalinity
+    + Hardness
+
+* Chemical Dosing Data
+    + Coagulant Dose
+    + Cationic Polymer Dose
+
+## Exploratory Analysis
+
+## Predictor Selection
+
+## Modeling
+
+### Linear Regression
+
+### [Other Models]
+
+### Tree-Based Models
+The benefit of a tree-based model in this application is that it is easily interpretable.
+It could serve as a decision-making tool for a treatment plant operator:
+
+* The operator sees a change in raw water conditions
+* Obtains a recommended dose based on the decision tree
+* Easily interprets the "why" of the dosing recommendation by following the logic of the decision tree
+* Understands which variables were most important in deciding on a dose

### 2025-12-01 Data Cleaning.Rmd
Added data processing for alkalinity and hardness
Completed data processing.

@@
 
-## R Markdown
+## Alkalinity and Hardness Data
 
@@
 
-merged
+summary(merged) # Both alkalinity and hardness have zero-values (no data)
+
+merged_clean1 <- merged[(merged$alkalinity != 0) & (merged$hardness != 0),] # remove
+
+summary(merged_clean1) # no zero values, still a very large alkalinity value
+
+tail(sort(merged_clean1$alkalinity)) # two significant outliers (equipment likely)
+
+merged_clean2 <- merged[merged$alkalinity < 876, ] # remove outliers
+
+summary(merged_clean2) # successfully removed zero values and outliers
 

### 2025-12-01 Exploratory Analysis.Rmd
Basic intro plots & finding outliers

---
title: "Exploratory Analysis"
author: "Reece Carmody"
date: "2025-11-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)

data = read.csv("RWAR_turb_cond_counts_pH_temp.csv")

data$Timestamp = mdy_hm(data$Timestamp)

# Trying to determine what rows have outliers.
# The goal is to go by column, remove outliers & average that day's data, then ultimately combine every column after it's cleaned to create our fully cleaned data frame.
# Then combine with alkalinity and hardness & coagulant doses data??

ggplot(data, aes(x = Timestamp, y = Turbidity..NTU)) +
  geom_point()

data[data$Turbidity..NTU > 10, ] # row 2701 and 5446

ggplot(data, aes(x = Timestamp, y = Conductivity..mS.cm.1)) +
  geom_point()

data[data$Conductivity..mS.cm.1 > 15, ] # row 5270

# Not sure what to do with particle count (PC) data just yet

ggplot(data, aes(x = Timestamp, y = pH..SU)) +
  geom_point()

ggplot(data, aes(x = Timestamp, y = Temperature..deg.C)) +
  geom_point()

```

### 2025-12-02 Exploratory Analysis Ethan Schilling.Rmd
Added my exploratory analysis file and plotted a few variables over time to see how they changed

---
title: "Exploratory Analysis Ethan Schilling"
output: html_document
---

```{r}
library(tidyverse)
library(knitr)
library(tidymodels)
library(dplyr)
```

```{r}
TDS_unfilt <- read_csv("RWAR_TDS.csv")
TOC_unfilt <- read_csv("RWAR_TOC.csv")

df_TDS <- TDS_unfilt %>%
  rename(date = 1, value = 2) %>%
  mutate(date = as.Date(date, format = "%d-%b-%y"))

ggplot(df_TDS, aes(x = date, y = value)) +
  geom_line() +
  labs(x = "Date", y = "TDS Value", title = "TDS over Time")

df_TOC <- TOC_unfilt %>% 
  rename(date = 1, value = 2) %>%
  mutate(date = as.Date(date, format = "%d-%b-%y"))

ggplot(df_TOC, aes(x = date, y = value)) +
  geom_line() +
  labs(x = "Date", y = "TOC Value", title = "TOC over Time")

```

### 2025-12-02 Data Cleaning.Rmd
Added code for cleaning temperature, pH, conductivity, and turbidity data.

@@
 ```
+
+```{r}
+# output merged data to CSV (run after processing)
+
+library(readr)
+write_csv(merged_clean2, "alkalinity_hardness_cleaned.csv")
+
+```
+
+## Raw Water Data
+
+```{r}
+
+library(lubridate)
+library(ggplot2)
+library(zoo)
+library(purrr)
+
+## Import data, convert timestamp
+rwar <- read.csv("RWAR_turb_cond_counts_pH_temp.csv")
+rwar2 <- rwar |> mutate(DateTime = mdy_hm(rwar$Timestamp))
+summary(rwar2)
+
+## Temperature Data
+# Plot
+ggplot(rwar2) + geom_point(aes(x=DateTime, y=Temperature..deg.C))
+# Find region of approx. zero temperature (outliers), set to NA
+# Interpolate between nearest non-NA temperatures where temp is NA
+rwar2_temp <- rwar2 |>
+  select(c("DateTime", Temperature..deg.C)) |>
+  mutate(temp_cleaned = ifelse(Temperature..deg.C < 1, NA, Temperature..deg.C)) |>
+  mutate(temp_interp = na.approx(temp_cleaned, na.rm = FALSE))
+# Average by day
+rwar2_temp_daily <- rwar2_temp |>
+  mutate(Day=date(DateTime)) |>
+  group_by(Day) |>
+  summarise(mean_Temp = mean(temp_interp))
+ggplot(rwar2_temp_daily) + geom_point(aes(x=Day, y=mean_Temp))
+
+## pH Data
+# Plot
+ggplot(rwar2) + geom_point(aes(x=DateTime, y=pH..SU))
+# Does not seem like there are huge outliers, so average by day
+rwar2_pH_daily <- rwar2 |>
+  select(c("DateTime", pH..SU)) |>
+  mutate(Day=date(DateTime)) |>
+  group_by(Day) |>
+  summarise(mean_pH = mean(pH..SU))
+ggplot(rwar2_pH_daily) + geom_point(aes(x=Day, y=mean_pH))
+
+## Conductivity Data
+# Plot
+ggplot(rwar2) + geom_point(aes(x=DateTime, y=Conductivity..mS.cm.1))
+# Remove obvious outliers (only a few measurements; does not remove full days)
+rwar2_cond <- rwar2 |> 
+  select(c("DateTime", Conductivity..mS.cm.1)) |>
+  filter(Conductivity..mS.cm.1 < 2 & Conductivity..mS.cm.1 > 0.2)
+ggplot(rwar2_cond) + geom_point(aes(x=DateTime, y=Conductivity..mS.cm.1))
+# Average by Day
+rwar2_cond_daily <- rwar2_cond |>
+  mutate(Day=date(DateTime)) |>
+  group_by(Day) |>
+  summarise(mean_Conductivity = mean(Conductivity..mS.cm.1))
+ggplot(rwar2_cond_daily) + geom_point(aes(x=Day, y=mean_Conductivity))
+
+## Turbidity Data
+# Plot
+ggplot(rwar2) + geom_point(aes(x=DateTime, y=Turbidity..NTU))
+# Remove obvious outliers (only a few measurements; does not remove full days)
+# Main obvious outliers are in mid 2019 and 2020, and are all above 7 NTU
+rwar2_turb <- rwar2 |> 
+  select(c("DateTime", Turbidity..NTU)) |>
+  filter(Turbidity..NTU < 7)
+ggplot(rwar2_turb) + geom_point(aes(x=DateTime, y=Turbidity..NTU))
+# Average by Day
+rwar2_turb_daily <- rwar2_turb |>
+  mutate(Day=date(DateTime)) |>
+  group_by(Day) |>
+  summarise(mean_Turbidity = mean(Turbidity..NTU))
+ggplot(rwar2_turb_daily) + geom_point(aes(x=Day, y=mean_Turbidity))
+
+## Merge daily temp, pH, conductivity, turbidity
+
+list_daily_data <- list(rwar2_temp_daily,
+                        rwar2_pH_daily, 
+                        rwar2_cond_daily,
+                        rwar2_turb_daily)
+merged_daily_data <- list_daily_data |>
+  reduce(full_join, by = "Day") |>
+  rename(Timestamp = Day)
+
+```
+
+```{r}
+## Export CSV of cleaned raw water data
+
+write_csv(merged_daily_data, )
+```

### 2025-12-04 Presentation DSCI445.Rmd
Draft of the Presentation

---
title: "DSCI 445 Presentation"
author: "Ethan Schilling"
date: "2025-12-04"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Slide 1 
```{r}
library(tidyverse)
library(knitr)
library(tidymodels)
library(dplyr)
```

```{r}
TDS_unfilt <- read_csv("RWAR_TDS.csv")
TOC_unfilt <- read_csv("RWAR_TOC.csv")

df_TDS <- TDS_unfilt %>%
  rename(date = 1, value = 2) %>%
  mutate(date = as.Date(date, format = "%d-%b-%y"))

ggplot(df_TDS, aes(x = date, y = value)) +
  geom_line() +
  labs(x = "Date", y = "TDS Value", title = "TDS over Time")

df_TOC <- TOC_unfilt %>% 
  rename(date = 1, value = 2) %>%
  mutate(date = as.Date(date, format = "%d-%b-%y"))

ggplot(df_TOC, aes(x = date, y = value)) +
  geom_line() +
  labs(x = "Date", y = "TOC Value", title = "TOC over Time")
```

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

### 2025-12-04 Exploratory Analysis.Rmd
Basic intro plots & finding outliers
Tried clustering

@@
 ```
+
+```{r}
+
+df1 = read.csv("RWAR_temp_pH_cond_turb_cleaned.csv")
+df2 = read.csv("alkalinity_hardness_cleaned.csv")
+df3 = read.csv("CoagulantDoses_2017-20.csv")
+
+df1 = df1 %>%
+  mutate(Date = ymd(Timestamp))
+df2 = df2 %>%
+  mutate(Date = mdy(Date))
+df3 = df3 %>%
+  mutate(Date = mdy(Date))
+
+# This merge is only keeping the rows that are in both data frames
+# Not using df3 because I don't want our target variables to be included in the clustering
+merged = df1 %>% inner_join(df2, by = "Date")
+merged$Date = NULL # We already have the Timestamp column
+
+cluster_data = merged %>%
+  select(where(is.numeric)) %>%
+  drop_na()
+
+cluster_data = scale(cluster_data) # Standardize the data
+
+cluster_fit = kmeans(cluster_data, centers = 5, nstart = 25)
+
+merged$cluster = cluster_fit$cluster
+
+# Interpretation
+
+table(merged$cluster)
+cluster_fit$centers
+
+ggplot(merged, aes(x = cluster_data[ , "mean_Temp"], y = cluster_data[ , "mean_pH"], # you can compare any two columns
+                   color = factor(cluster))) +
+  geom_point() +
+  labs(title = "K-Means Clustering", color = "Cluster") +
+  theme_bw()
+
+```

### 2025-12-04 Presentation DSCI445.Rmd
Created some plots and the presentation file

@@
 title: "DSCI 445 Presentation"
-author: "Ethan Schilling"
+author: "Ethan Schilling, Joshua Tobey, Dawson Carney, Reece Carmody"
 date: "2025-12-04"
@@
 ## Slide 1
-```{r}
-library(tidyverse)
-library(knitr)
-library(tidymodels)
-library(dplyr)
-```
 
 ```{r}
-TDS_unfilt <- read_csv("RWAR_TDS.csv")
-TOC_unfilt <- read_csv("RWAR_TOC.csv")
-
-df_TDS <- TDS_unfilt %>%
-  rename(date = 1, value = 2) %>%
-  mutate(date = as.Date(date, format = "%d-%b-%y"))
-
-ggplot(df_TDS, aes(x = date, y = value)) +
-  geom_line() +
-  labs(x = "Date", y = "TDS Value", title = "TDS over Time")
-
-df_TOC <- TOC_unfilt %>% 
-  rename(date = 1, value = 2) %>%
-  mutate(date = as.Date(date, format = "%d-%b-%y"))
-
-ggplot(df_TOC, aes(x = date, y = value)) +
-  geom_line() +
-  labs(x = "Date", y = "TOC Value", title = "TOC over Time")
+readRDS("TOC_plot.rds")
+readRDS("TDS_plot.rds")
 ```
-
-## Slide with Bullets
-
-- Bullet 1
-- Bullet 2
-- Bullet 3
-
-## Slide with R Output
-
-```{r cars, echo = TRUE}
-summary(cars)
-```
-
-## Slide with Plot
-
-```{r pressure}
-plot(pressure)
-```

### 2025-12-04 Exploratory Analysis.Rmd
Attempting clustering (merged data, no lags)

@@
 # Not using df3 because I don't want our target variables to be included in the clustering
-merged = df1 %>% inner_join(df2, by = "Date")
-merged$Date = NULL # We already have the Timestamp column
+merged = df1 %>% 
+  inner_join(df2, by = "Date")
+merged$Timestamp = NULL # We'll keep the Date column
 
@@
 
-ggplot(merged, aes(x = cluster_data[ , "mean_Temp"], y = cluster_data[ , "mean_pH"], # you can compare any two columns
+cluster_plot = merged %>% 
+  inner_join(df3, by = "Date")
+
+# You can change the names to see different variables and their clusters
+ggplot(merged, aes(x = cluster_plot[ , "mean_Turbidity"], y = cluster_plot[ , "Ferric.chloride..mg.L"], # you can compare any two columns
                    color = factor(cluster))) +

### 2025-12-04 Exploratory_Analysis_Josh_Tobey.Rmd
WIP before rebase

--- 2025-11-17 Exploratory_Analysis_Josh_Tobey.Rmd
@@
 ```{r}
+install.packages(c("usethis", "gitcreds"))
+```
+
+```{r}
+library(tidyverse)
+library(tidymodels)
+library(dplyr)
+library(knitr)
 
 ```
+
+```{r}
+exploratory_df <- read_csv("RWAR_DOSING_clean.csv")
+colnames(exploratory_df)
+```
+
+```{r}
+ggplot(exploratory_df) +
+  geom_point(aes(x = `Turbidity, NTU`,
+                 y = `Temperature, deg C`,
+                 color = `Temperature, deg C`)) +
+  scale_color_gradient(low = 'yellow', high = 'green')
+```
+
+```{r}
+head(exploratory_df)
+
+```
+```{r}
+exploratory_df$Timestamp = mdy_hm(exploratory_df$Timestamp)
+ggplot(exploratory_df, aes(x = Timestamp,
+                           y = `pH, SU`,
+                           color = `pH, SU`)) +
+  geom_line() + 
+  scale_color_gradient(low = 'blue', high = 'red')
+
+```
+
+```{r}
+unique(exploratory_df$`Conductivity, mS cm-1`)
+```
+```{r}
+ggplot(exploratory_df) +
+  geom_jitter(aes(x = `Turbidity, NTU`,
+                 y = `Conductivity, mS cm-1`,
+                 color = `Conductivity, mS cm-1`), width = 1, height = 3) +
+  scale_color_gradient(low = 'skyblue', high = 'orange')
+
+```
+Nothing of real interest here.
+```{r}
+coag_dose_df <- read_csv("CoagulantDoses_2017-20.csv")
+head(coag_dose_df)
+
+```
+```{r}
+#exploratory_df <- exploratory_df %>%
+  #mutate(`dosing (exploratory)` = coag_dose_df$`Cationic Polimer, mg/L` )
+length(exploratory_df) == length(coag_dose_df)
+```
+```{r}
+
+coag_dose_df$Date <- mdy(coag_dose_df$Date)
+ggplot(coag_dose_df) +
+  geom_line(aes(x = Date,
+                y = `Cationic Polimer, mg/L`))
+
+```
+```{r}
+alk_df <- read_csv("alkalinity_hardness_cleaned.csv")
+head(alk_df)
+
+```
+```{r}
+alk_df$Date <- mdy(alk_df$Date)
+ggplot(alk_df) +
+  geom_line(aes(x = Date,
+                y = alkalinity),
+            color = 'orange') +
+  geom_line(aes(x = Date,
+                y = hardness),
+            color = 'blue')
+
+```
+```{r}
+generic_mlr <- linear_reg() |>
+  set_mode("regression") |>
+  set_engine("lm")
+
+mlr_recipe <- recipe(alkalinity ~ hardness, data = alk_df)
+mlr_wf <- workflow() |> 
+  add_model(generic_mlr) |> 
+  add_recipe(mlr_recipe)
+fit(mlr_wf, data = alk_df)
+```
+
+```{r}
+time_lag_1 <- 
+time_lag_2 <- 
+time_lag_3 <-
+
+weekly_log_spec <- logistic_reg()
+weekly_log_fit <- weekly_log_spec |>
+  fit(Direction ~ Lag1 + 
+      Lag2, 
+      family = "binomial", 
+      data = Weekly)
+weekly_log_fit |>
+  pluck("fit") |>
+  summary()
+
+```
+
+```{r}
+cross_validation <- initial_split(exploratory_df, prop = 0.6)
+exploratory_train <- training(cross_validation)
+exploratory_test <- testing(cross_validation)
+#OPTION 1:
+linear_recipe <- recipe(`pH SU` ~ `Temperature deg C`, data = exploratory_df)
+quad_recipe <- linear_recipe |> step_mutate(`Temperature deg C_2` = `Temperature deg C`^2)
+cubic_recipe <- quad_recipe |> step_mutate(`Temperature deg C_3` = `Temperature deg C`^3)
+quart_recipe <- cubic_recipe |> step_mutate(`Temperature deg C_4` = `Temperature deg C`^4)
+#OPTION 2:
+linear_recipe <- recipe(`pH SU` ~ ., data = exploratory_df)
+quad_recipe <- linear_recipe |> step_poly(all_numeric_predictors(), degree = 2)
+cubic_recipe <- quad_recipe |> step_poly(all_numeric_predictors(), degree = 3)
+quart_recipe <- cubic_recipe |> step_poly(all_numeric_predictors(), degree = 4)
+
+loocv_df <- loo_cv(dfi)
+test_errors <- data.frame()
+
+retrieve_test_error <- function(least_sqrs_model, testing_set) {
+  augment(least_sqrs_model, new_data = testing_set) |>
+    mutate(resid2 = .resid^2) |> 
+    summarise(rmse = sqrt(mean(resid2))) |>
+    pull(rmse)
+}
+
+for (i in seq_len(nrow(loocv_df))) {
+  train_data = training(loocv_df[i,]$splits[[1]])
+  test_data = testing(loocv_df[i,]$splits[[1]])
+
+  Y1 <- workflow() |> add_model(lstSqrs_spec) |> add_recipe(linear_recipe) |> fit(train_data)
+  Y2 <- workflow() |> add_model(lstSqrs_spec) |> add_recipe(quad_recipe) |> fit(train_data)
+  Y3 <- workflow() |> add_model(lstSqrs_spec) |> add_recipe(cubic_recipe) |> fit(train_data)
+  Y4 <- workflow() |> add_model(lstSqrs_spec) |> add_recipe(quart_recipe) |> fit(train_data)
+
+  data.frame(model = c("linear", "quadratic", "cubic", "quartic"),
+             rmse = c(retrieve_test_error(Y1, test_data), 
+                   retrieve_test_error(Y2, test_data), 
+                   retrieve_test_error(Y3, test_data), 
+                   retrieve_test_error(Y4, test_data)),
+             i = i) |>
+    rbind(test_errors) -> test_errors
+
+}
+test_errors |> 
+  group_by(model) |>
+  summarise(rmse = sqrt(mean(rmse))) |>
+  kable()
+
+```

### 2025-12-04 Data Cleaning.Rmd

@@
 
+```{r}
+library(dplyr)
+library(tidyr)
+library(readr)
+library(lubridate)
+library(ggplot2)
+library(zoo)
+library(purrr)
+
+```
+
 ## Alkalinity and Hardness Data
@@
 ```{r}
-library(dplyr)
-library(tidyr)
 
@@
 
-library(readr)
 write_csv(merged_clean2, "alkalinity_hardness_cleaned.csv")
@@
 ```{r}
-
-library(lubridate)
-library(ggplot2)
-library(zoo)
-library(purrr)
 
@@
 ## Export CSV of cleaned raw water data
+# run once, only after processing
+write_csv(merged_daily_data, "RWAR_temp_pH_cond_turb_cleaned.csv")
 
-write_csv(merged_daily_data, )
 ```
+
+## Coagulant and Polymer Doses
+
+```{r}
+
+doses <- read.csv("CoagulantDoses_2017-20.csv")
+doses <- doses |> 
+  mutate(Timestamp = mdy(doses$Date)) |>
+  select(-c("Date"))
+
+ggplot(doses) +
+  geom_point(aes(x=Timestamp, y=Ferric.chloride..mg.L), color="red") +
+  geom_point(aes(x=Timestamp, y=Cationic.Polimer..mg.L), color="blue") +
+  xlab("Date")+
+  ylab("Dosing (mg/L)")+
+  ggtitle("Doses over Time (FeCl red, Cationic Polymer blue)")
+
+RWAR_dosing_cleaned <- merge(doses, merged_daily_data, by="Timestamp")
+
+```
+
+```{r}
+## Export to CSV
+# only execute after run processing steps above
+write_csv(RWAR_dosing_cleaned, "RWAR_dosing_cleaned.csv")
+
+```
+
+```{r}
+water_lm <- lm(Ferric.chloride..mg.L ~ ., 
+               data=RWAR_dosing_cleaned |> select(-c(Timestamp, Cationic.Polimer..mg.L)))
+summary(water_lm)
+
+```
+
+```{r}
+data <- read.csv("RWAR_dosing_cleaned.csv")
+data |>
+  mutate(Temp_lag1 = lag(mean_Temp, 1), 
+         Temp_lag2 = lag(mean_Temp, 2),
+         pH_lag1 = lag(mean_pH, 1),
+         pH_lag2 = lag(mean_pH, 2),
+         Conductivity_lag1 = lag(conductivity, 1),
+         Conductivity_lag2 = lag(conductivity, 2))
+
+```

### 2025-12-04 Exploratory Analysis Reece C.Rmd
Working on graphs

--- 2025-12-04 Exploratory Analysis.Rmd
@@
 ---
-title: "Exploratory Analysis"
+title: "Exploratory Analysis Reece C"
 author: "Reece Carmody"
@@
 
+# Merging data
+
 df1 = read.csv("RWAR_temp_pH_cond_turb_cleaned.csv")
@@
 
+# Clustering (probably not as useful with time series data...)
+
 cluster_fit = kmeans(cluster_data, centers = 5, nstart = 25)
@@
 
-cluster_plot = merged %>% 
+full_merge = merged %>% 
   inner_join(df3, by = "Date")
@@
 # You can change the names to see different variables and their clusters
-ggplot(merged, aes(x = cluster_plot[ , "mean_Turbidity"], y = cluster_plot[ , "Ferric.chloride..mg.L"], # you can compare any two columns
+ggplot(full_merge, aes(x = full_merge[ , "mean_Turbidity"],
+                       y = full_merge[ , "Ferric.chloride..mg.L"], # you can compare any two columns
                    color = factor(cluster))) +
@@
 ```
+
+```{r}
+
+# Exploratory plots
+
+long_df = full_merge %>%
+  pivot_longer(cols = c("mean_Temp", "mean_pH", "mean_Conductivity", "mean_Turbidity",
+                        "alkalinity", "hardness", "Ferric.chloride..mg.L", "Cationic.Polimer..mg.L"),
+               names_to = "metric",
+               values_to = "value")
+
+ggplot(long_df, aes(x = Date, y = value))
+
+ggplot(full_merge, aes(x = Date, y = mean_pH)) +
+  geom_smooth() + 
+  labs(title = "title") +
+  theme_bw()
+
+```

### 2025-12-04 Paper Draft.Rmd
Added section on classification algorithms.

@@
 
-[Background on Drinking Water Treatment/Process Overview]
+The drinking water treatment process takes water from a river, lake, reservoir, or other source, and purifies it to have it reach drinking water standards.
+
+There are four primary steps of the water treatment process:  
+1. Coagulation  
+2. Flocculation  
+3. Filtration  
+4. Disinfection
 
@@
 
+### Data Overview
+
 The data used for this effort is a timeseries dataset from a Colorado water treatment
@@
     + Cationic Polymer Dose
+
+### Data Cleaning
 
@@
 * Understands which variables were most important in deciding on a dose
+
+### Classification of Dose Change
+
+We added two variables called `dose_change_FeCl` and `dose_change_Poly` that indicate
+whether the chemical dose was increased, decreased, or stayed the same. Additionally, we
+added variables indicating change in temperature, pH, turbidity, and conductivity from
+the previous day.
+
+This will allow classification methods to be applied. Additionally, if effective,
+this could provide a more helpful tool for operators. Instead of attempting to 
+predict a specific dose, it could give a general direction for an operator to 
+attempt a dose change. This is how plants work in practice: an operator uses 
+some tool (a water test, a reading from a device, etc.), and uses that to make 
+a dose change. They then check important treatment metrics and make any needed
+further adjustments.
+
+```{r}
+
+```

### 2025-12-04 Exploratory Analysis Reece C.Rmd

@@
 ---
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
 
@@
 
-ggplot(long_df, aes(x = Date, y = value))
+ggplot(long_df, aes(x = Date, y = value)) +
+  geom_point(size = 0.1) +
+  facet_wrap(~ metric, scales = "free_y")
 
-ggplot(full_merge, aes(x = Date, y = mean_pH)) +
-  geom_smooth() + 
-  labs(title = "title") +
-  theme_bw()
+#ggplot(full_merge, aes(x = Date, y = mean_pH)) +
+#  geom_smooth() + 
+#  labs(title = "title") +
+#  theme_bw()
 

### 2025-12-04 Exploratory Analysis Ethan Schilling.Rmd
my analysis

@@
 ```{r}
+
 TDS_unfilt <- read_csv("RWAR_TDS.csv")
@@
 
-ggplot(df_TDS, aes(x = date, y = value)) +
+TDS_plot <- ggplot(df_TDS, aes(x = date, y = value)) +
   geom_line() +
   labs(x = "Date", y = "TDS Value", title = "TDS over Time")
+
+saveRDS(TDS_plot,"TDS_plot.rds")
 
@@
 
-ggplot(df_TOC, aes(x = date, y = value)) +
+TOC_plot <- ggplot(df_TOC, aes(x = date, y = value)) +
   geom_line() +
@@
 
+saveRDS(TOC_plot,"TOC_plot.rds")
+
 ```

### 2025-12-04 HomeworkSet_10_Joshua_Tobey.Rmd

---
title: "HomeworkSet_10"
author: "Joshua Tobey"
date: "2025-12-04"
output: html_document
---
```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
library(ggplot2)

set.seed(445)
```
PROBELM 1) We will explore the maximal margin classifier on a toy data set.
a) 
Refer to the sketch in the pdf.

b) Equation for Optimal hyperplane, which I found in part d.  See the sketch in the pdf.

$f(x^*) =  \beta_0 + \beta_1x_1 + \beta_2x_2$
$>>1 - 2\beta_1 + 2\beta_2 = 0$

c) Describe the classification rule for the maximal marginal classifier. It should be along the lines of “Classify to Red if 𝛽0+𝛽1𝑋1+𝛽2𝑋2>0, and classify as Blue otherwise. Provide the values of 𝛽0,𝛽1,𝛽2.

\[
\beta_0 + \beta_1x_1 + \beta_2x_2 
\begin{cases}
> 0 & \text{Classify Red} \\
< 0 & \text{Classify Blue} \\
= 0 & \text{On hyperplane}
\end{cases}
\]

Values of $\beta_0, \beta_1 & \beta_2$: Refer to the pdf of sketch which has the algebra work on it.
It uses the 3 support vectors (2,1), (2,2) and (4,3) as well as their equations.

$\beta_0 = 1$
$\beta_1 = -2$
$\beta_2 = 2$

d) On your sketch, indicate the margin for the maximal margin classifier.

Refer back to pdf of sketch and algebra work.  Margins are in orange on the sketched graph.

Given hyperplane $ = 1 - 2\beta_1 + 2\beta_2 = 0$.  The margins are subject to:
\[
$1 - 2\beta_1 + 2\beta_2 = 1$
$1 - 2\beta_1 + 2\beta_2 = -1$
\]

e) Indicate the support vectors for the maximal margin classifier.

Testing which points give an equation equal to 1 for red points and -1 for blue points.
Red support vectors:
$(2,2) = 1 - 2(2) + 2(2) = 1$
$(4,4) = 1 - 2(4) + 2(4) = 1$
Blue support vectors:
$(2,1) = 1 - 2(2) + 2(1) = -1$
$(4,3) = 1 - 2(4) + 2(3) = -1$

f) Argue that a slight movement of the seventh observation would not affect the maximal margin hyperplane.

Seeing that the seventh is a blue categorized point, it lies far below the hyper plane, so slight adjustments won't cause the point to interfere with the support vectors that are much closer to the hyperplane.

g) Draw an additional observation on the plot so that the two classes are no longer separable by a hyperplane.

Refer to the pdf of the sketched graph.  The added point will be circled in green.

PROBLEM 2) We have seen that we can fit an SVM with a non-linear kernel in order to perform classification using a non-linear decision boundary. We will now see that we can also obtain a non-linear decision boundary by performing logistic regression using non-linear transformations of the features.

a) Generate a data set with 𝑛=500 and 𝑝=2, such that the observations belong to two classes with a quadratic decision boundary.

```{r}
n <- 500
x1 <- runif(n) - 0.5
x2 <- runif(n) - 0.5
y = as.numeric(x1^2 - x2^2 > 0)

```
b) Plot the observations, colored according to their class labels.
```{r}
plot(x1, x2, col = ifelse(y == 1, "red", "blue"))

```
c) Fit a logistic regression model to the data using 𝑋1 and 𝑋2 as predictors.
```{r}
library(parsnip)
log_df <- data.frame(x1 = x1, x2 = x2, response = factor(y))
log_spec <- logistic_reg()
log_fit <- log_spec |>
  fit(response ~ x1 + x2, data = log_df)
log_fit |>
  pluck("fit") |>
  summary()

```
I googled the error:

Error in `check_outcome()`:
! For a classification model, the outcome should be a <factor>, not a double vector.
Run `rlang::last_trace()` to see where the error occurred.

This led me to using the factor() function on the y in creating a dataframe to fit the logistic model.

d) Apply this model to the training data in order to obtain a predicted class label for each training observation. Plot the observations, colored according to the predicted class labels. What shape is the decision boundary?

```{r}

```

### 2025-12-04 Data Cleaning.Rmd
Completed code for adding Lag variables to dataset.

@@
 data <- read.csv("RWAR_dosing_cleaned.csv")
-data |>
+lagged_data <- data |>
   mutate(Temp_lag1 = lag(mean_Temp, 1), 
@@
          pH_lag2 = lag(mean_pH, 2),
-         Conductivity_lag1 = lag(conductivity, 1),
-         Conductivity_lag2 = lag(conductivity, 2))
+         Cond_lag1 = lag(mean_Conductivity, 1),
+         Cond_lag2 = lag(mean_Conductivity, 2),
+         Turb_lag1 = lag(mean_Turbidity, 1),
+         Turb_lag2 = lag(mean_Turbidity, 2))
+# Temp
+lagged_data$Temp_lag1[1] <- lagged_data$Temp_lag1[2]
+lagged_data$Temp_lag2[1:2] <- lagged_data$Temp_lag2[3]
+# pH
+lagged_data$pH_lag1[1] <- lagged_data$pH_lag1[2]
+lagged_data$pH_lag2[1:2] <- lagged_data$pH_lag2[3]
+# Conductivity
+lagged_data$Cond_lag1[1] <- lagged_data$Cond_lag1[2]
+lagged_data$Cond_lag2[1:2] <- lagged_data$Cond_lag2[3]
+# Turbidity
+lagged_data$Turb_lag1[1] <- lagged_data$Turb_lag1[2]
+lagged_data$Turb_lag2[1:2] <- lagged_data$Turb_lag2[3]
+
+lagged_data
+lm(Ferric.chloride..mg.L ~ ., data=lagged_data |> select(-c(Timestamp)))
 

### 2025-12-05 Exploratory_Analysis_Josh_Tobey.Rmd
New plots and some preliminary model fitting

@@
 library(knitr)
+library(lubridate)
 
@@
 ```{r}
-exploratory_df <- read_csv("RWAR_DOSING_clean.csv")
+exploratory_df <- read_csv("RWAR_dosing_cleaned.csv")
 colnames(exploratory_df)
@@
 ggplot(exploratory_df) +
-  geom_point(aes(x = `Turbidity, NTU`,
-                 y = `Temperature, deg C`,
-                 color = `Temperature, deg C`)) +
+  geom_point(aes(x = mean_Turbidity,
+                 y = mean_Temp,
+                 color = mean_Temp)) +
   scale_color_gradient(low = 'yellow', high = 'green')
@@
 ```{r}
-exploratory_df$Timestamp = mdy_hm(exploratory_df$Timestamp)
+summary(exploratory_df)
+
+```
+
+```{r}
+exploratory_df$Timestamp = ymd(exploratory_df$Timestamp)
+exploratory_df <- exploratory_df[order(exploratory_df$Timestamp), ]
 ggplot(exploratory_df, aes(x = Timestamp,
-                           y = `pH, SU`,
-                           color = `pH, SU`)) +
+                           y = mean_pH,
+                           color = mean_pH)) +
   geom_line() + 
@@
 ```
+```{r}
+exploratory_df$Timestamp <- ymd(exploratory_df$Timestamp)
+exploratory_df <- exploratory_df[order(exploratory_df$Timestamp), ]
+ggplot(exploratory_df, aes(x = Timestamp,
+                           y = mean_Temp,
+                           color = mean_Temp)) +
+  geom_line() + 
+  scale_color_gradient(low = 'yellow', high = 'green')
+
+```
+```{r}
+library(patchwork)
+plot1 <- ggplot(exploratory_df, aes(x = Timestamp,
+                           y = mean_pH,
+                           color = mean_pH)) +
+  geom_line() + 
+  scale_color_gradient(low = 'blue', high = 'red') +
+  theme_minimal() +
+  ggtitle("Average pH over Time")
+
+plot2 <- ggplot(exploratory_df, aes(x = Timestamp,
+                           y = Cationic.Polimer..mg.L,
+                           color = Cationic.Polimer..mg.L)) +
+  geom_line() + 
+  scale_color_gradient(low = 'yellow', high = 'green') +
+  theme_minimal() +
+  ggtitle("Polymer Coagulant over Time")
+
+plot1/plot2
+```
 
 ```{r}
-unique(exploratory_df$`Conductivity, mS cm-1`)
-```
-```{r}
 ggplot(exploratory_df) +
-  geom_jitter(aes(x = `Turbidity, NTU`,
-                 y = `Conductivity, mS cm-1`,
-                 color = `Conductivity, mS cm-1`), width = 1, height = 3) +
+  geom_jitter(aes(x = mean_Turbidity,
+                 y = mean_Conductivity,
+                 color = mean_Conductivity), width = 1, height = 3) +
   scale_color_gradient(low = 'skyblue', high = 'orange')
@@
 ```
+
+coagual

### 2025-12-05 Data Cleaning.Rmd
Re-made the joining of dosign and rwar datasets after merge issues.

--- 2025-12-02 Data Cleaning.Rmd
@@
 
-write_csv(merged_daily_data, )
+#write_csv(merged_daily_data, )
 ```
+
+```{r}
+# Merge dosing data to this data. Have significant overlap
+dosing <- read.csv("CoagulantDoses_2017-20.csv") |>
+  rename(Timestamp = Date) |> 
+  mutate(Timestamp = mdy(dosing$Timestamp))
+
+RWAR_dosing_cleaned <- inner_join(dosing, merged_daily_data)
+summary(RWAR_dosing_cleaned) #can export to csv with same name
+
+```
+
+```{r}
+water_lm <- lm(Ferric.chloride..mg.L ~ ., 
+               data=RWAR_dosing_cleaned |> select(-c(Timestamp, Cationic.Polimer..mg.L)))
+summary(water_lm)
+
+```
+
+```{r}
+data <- read.csv("RWAR_dosing_cleaned.csv")
+lagged_data <- data |>
+  mutate(Temp_lag1 = lag(mean_Temp, 1), 
+         Temp_lag2 = lag(mean_Temp, 2),
+         pH_lag1 = lag(mean_pH, 1),
+         pH_lag2 = lag(mean_pH, 2),
+         Cond_lag1 = lag(mean_Conductivity, 1),
+         Cond_lag2 = lag(mean_Conductivity, 2),
+         Turb_lag1 = lag(mean_Turbidity, 1),
+         Turb_lag2 = lag(mean_Turbidity, 2))
+# Temp
+lagged_data$Temp_lag1[1] <- lagged_data$Temp_lag1[2]
+lagged_data$Temp_lag2[1:2] <- lagged_data$Temp_lag2[3]
+# pH
+lagged_data$pH_lag1[1] <- lagged_data$pH_lag1[2]
+lagged_data$pH_lag2[1:2] <- lagged_data$pH_lag2[3]
+# Conductivity
+lagged_data$Cond_lag1[1] <- lagged_data$Cond_lag1[2]
+lagged_data$Cond_lag2[1:2] <- lagged_data$Cond_lag2[3]
+# Turbidity
+lagged_data$Turb_lag1[1] <- lagged_data$Turb_lag1[2]
+lagged_data$Turb_lag2[1:2] <- lagged_data$Turb_lag2[3]
+
+lm_fit <- lm(Ferric.chloride..mg.L ~ ., data=lagged_data |> select(-c(Timestamp)))
+summary(lm_fit)
+
+```

### 2025-12-08 Paper Draft.Rmd

@@
 
+```{r, include=F}
+library(dplyr)
+library(tidyr)
+library(readr)
+library(lubridate)
+library(ggplot2)
+library(zoo)
+library(purrr)
+library(knitr)
+library (patchwork)
+```
+
 ## Introduction
@@
 There are four primary steps of the water treatment process:  
-1. Coagulation  
-2. Flocculation  
-3. Filtration  
-4. Disinfection
+1. Coagulation - Chemicals ("coagulants") are added to raw water to help contaminants
+group together into "floc" particles  
+2. Flocculation - The water is then slowly mixed to allow these floc particles to grow  
+3. Filtration - These particles are filtered out  
+4. Disinfection - Water is disinfected to get rid of biological contaminants
+
+This project will focus on Step 1. These chemical doses are one of the most important
+features of the treatment process that an operator can change to improve performance.
 
@@
 The data used for this effort is a timeseries dataset from a Colorado water treatment
-plant covering XX years. This treatment plant takes its water from a reservoir, which
-tends to be a more stable water source than sources such as rivers or industrial supplies.
+plant covering 3 years, from 2018-2020. This treatment plant takes its water from a reservoir, which tends to be a more stable water source than sources such as rivers or industrial supplies.
 
-The available data are as follows, with missing date ranges noted:
+The available data are as follows:
 
@@
     + pH
-    + Temperature
+    + Temperature (of the water)
     + Turbidity
@@
 * Chemical Dosing Data
-    + Coagulant Dose
-    + Cationic Polymer Dose
+    + Coagulant Dose - primary additive (allows for floc particle formation)
+    + Cationic Polymer Dose - secondary additive (boosts size of floc particles)
 
@@
 
+Basic data cleaning was conducted based on visual inspection. Most of the raw water 
+data we have available was on a 4-hour time increment. So, we decided to combine and 
+average by day for our predictions. This is because the dosing data is only 
+available as daily averages, so we needed to match our time resolution.
+
+Before this averaging, we removed or modified values that were clearly outliers, so
+the averages would not be skewed by incorrect data. Reasons for these outliers are
+most likely equipment malfunction.
+
+* pH: No significant outliers that we could see, so we simply day-averaged.  
+* Temperature: There was one region of approximately zero temperature (unrealistic
+for the reservoir). We got rid of these values and interpolated to the nearest non-zero temperature.
+* Conductivity: Values less than 0.2 and greater than 2 were removed, since they 
+are unreasonable based on this dataset and the "usual" values.
+* Turbidity: A few exceptionally high values (relative to the usual data values)
+were removed before averaging.
+* Grain Size Information: This data was highly variable, and simply looks like "noise",
+so did not seem helpful for prediction. We thus did not clean it.
+* Alkalinity and Hardness: These data have different time availability than the other
+datasets but will be inspected. Removed zero-values.
+
+Final Combination: We merged the raw water data (excluding alkalinity/hardness) with the dosing data. We only lost about 10 records in this process, where there was not available 
+dosing data for the raw water measurements.
+
 ## Exploratory Analysis
 
-## Predictor Selection
+Some of this cleaned data is presented below.
+
+### Plots of Key Treatment Plant Variables
+
+Alkalinity and hardness tend to stay fairly steady over time.
+
+```{r,echo=F}
+alk_df <- read_csv("alkalinity_hardness_cleaned.csv",show_col_types = FALSE)
+alk_df$Date <- mdy(alk_df$Date)
+
+plot_alk <- ggplot(alk_df, aes(x = Date,
+                           y = alkalinity,
+                           color = alkalinity)) +
+  geom_point() + 
+  scale_color_gradient(low = 'blue', high = 'red') +
+  theme_minimal() +
+  ggtitle("Alkalinity")
+
+plot_hardness <- ggplot(alk_df, aes(x = Date,
+                           y = hardness,
+                           color = hardness)) +
+  geom_point() + 
+  scale_color_gradient(low = 'blue', high = 'red') +
+  theme_minimal() +
+  ggtitle("Hardness")
+
+plot_alk/plot_hardness
+```
+See below for dosing compared to key raw water characteristics.
+
+```{r,echo=F}
+
+rwar_dosing <- read.csv("RWAR_dosing_cleaned.csv")
+rwar_dosing$Timestamp <- ymd(rwar_dosing$Timestamp)
+
+plot_pH <- ggplot(rwar_dosing, aes(x = Timestamp,
+                           y = mean_pH,
+                           color = mean_pH)) +
+  geom_line() + 
+  scale_color_gradient(low = 'blue', high = 'red') +
+  theme_minimal() +
+  ggtitle("Average pH")
+
+plot_polymer <- ggplot(rwar_dosing, aes(x = Timestamp,
+                           y = Cationic.Polimer..mg.L,
+                           color = Cationic.Polimer..mg.L)) +
+  geom_line() + 
+  scale_color_gradient(low = 'yellow', high = 'green') +
+  theme_minimal() +
+  ggtitle("Polymer Coagulant")
+
+plot_FeCl <- ggplot(rwar_dosing, aes(x = Timestamp,
+                           y = Ferric.chloride..mg.L,
+                           color = Ferric.chloride..mg.L)) +
+  geom_line() + 
+  scale_color_gradient(low = 'blue', high = 'orange') +
+  theme_minimal() +
+  ggtitle("FeCl Coagulant")
+
+plot_temp <- ggplot(rwar_dosing, aes(x = Timestamp,
+                           y = mean_Temp,
+                           color = mean_Temp)) +
+  geom_line() + 
+  scale_color_gradient(low = 'blue', high = 'red') +
+  theme_minimal() +
+  ggtitle("Water Temperature")
+
+plot_cond <- ggplot(rwar_dosing, aes(x = Timestamp,
+                           y = mean_Conductivity,
+                           color = mean_Conductivity)) +
+  geom_line() + 
+  scale_color_gradient(low = 'purple', high = 'yellow') +
+  theme_minimal() +
+  ggtitle("Water Conductivity")
+
+plot_turb <- ggplot(rwar_dosing, aes(x = Timestamp,
+                           y = mean_Turbidity,
+                           color = mean_Turbidity)) +
+  geom_line() + 
+  scale_color_gradient(low = 'black', high = 'orange') +
+  theme_minimal() +
+  ggtitle("Water Turbidity")
+
+plot_polymer/plot_FeCl
+plot_temp / plot_pH
+plot_cond / plot_turb
+```
+
+### Predictor Selection
+
+Overall, it seems like temperature and pH have the strongest 
+relationships to chemical dosing. This is consistent with feedback from treatment
+plant operators. It is not immediately clear which relationships could
+make effective predictors. Turbidity is highly variable and likely does 
+not have a strong relationship. Conductivity is similar.
+
+### Addition of Lagged Variables
+
+To capture time-dependence of this dataset, we will introduce lagged variables. 
+For each of the main water parameters, we will introduce two lagged variables, giving
+the water characteristics for the day before and two days before.
+```{r, echo=F}
+data <- read.csv("RWAR_dosing_cleaned.csv")
+lagged_data <- data |>
+  mutate(Temp_lag1 = lag(mean_Temp, 1), 
+         Temp_lag2 = lag(mean_Temp, 2),
+         pH_lag1 = lag(mean_pH, 1),
+         pH_lag2 = lag(mean_pH, 2),
+         Cond_lag1 = lag(mean_Conductivity, 1),
+         Cond_lag2 = lag(mean_Conductivity, 2),
+         Turb_lag1 = lag(mean_Turbidity, 1),
+         Turb_lag2 = lag(mean_Turbidity, 2))
+# Temp
+lagged_data$Temp_lag1[1] <- lagged_data$Temp_lag1[2]
+lagged_data$Temp_lag2[1:2] <- lagged_data$Temp_lag2[3]
+# pH
+lagged_data$pH_lag1[1] <- lagged_data$pH_lag1[2]
+lagged_data$pH_lag2[1:2] <- lagged_data$pH_lag2[3]
+# Conductivity
+lagged_data$Cond_lag1[1] <- lagged_data$Cond_lag1[2]
+lagged_data$Cond_lag2[1:2] <- lagged_data$Cond_lag2[3]
+# Turbidity
+lagged_data$Turb_lag1[1] <- lagged_data$Turb_lag1[2]
+lagged_data$Turb_lag2[1:2] <- lagged_data$Turb_lag2[3]
+```
 
@@
 ### Linear Regression
+
+```{r}
+
+lm_fit <- lm(Ferric.chloride..mg.L ~ ., data=lagged_data |> select(-c(Timestamp, Cationic.Polimer..mg.L)))
+summary(lm_fit)
+
+```
+ Interesting initial result: the pH and temperature from two days before are
+ the only predictors with low p-values. This makes sense based on the "lag time"
+ from water quality change to dosing change.
+
+ We will now try models incorporating only some of the predictors.
 

### 2025-12-08 Data Cleaning.Rmd
Updated alkalinity/hardness cleaning to fix error that did not remove outliers.
Edited lag addition. Added a dose change column for each of coagulant doses to run predictive model on dose change.

@@
 
-merged_clean2 <- merged[merged$alkalinity < 876, ] # remove outliers
+merged_clean2 <- merged_clean1[merged_clean1$alkalinity < 120, ] # remove outliers
 
@@
 data <- read.csv("RWAR_dosing_cleaned.csv")
-lagged_data <- data |>
+rwar_dosing_lag <- data |>
   mutate(Temp_lag1 = lag(mean_Temp, 1), 
@@
          Turb_lag1 = lag(mean_Turbidity, 1),
-         Turb_lag2 = lag(mean_Turbidity, 2))
-# Temp
-lagged_data$Temp_lag1[1] <- lagged_data$Temp_lag1[2]
-lagged_data$Temp_lag2[1:2] <- lagged_data$Temp_lag2[3]
-# pH
-lagged_data$pH_lag1[1] <- lagged_data$pH_lag1[2]
-lagged_data$pH_lag2[1:2] <- lagged_data$pH_lag2[3]
-# Conductivity
-lagged_data$Cond_lag1[1] <- lagged_data$Cond_lag1[2]
-lagged_data$Cond_lag2[1:2] <- lagged_data$Cond_lag2[3]
-# Turbidity
-lagged_data$Turb_lag1[1] <- lagged_data$Turb_lag1[2]
-lagged_data$Turb_lag2[1:2] <- lagged_data$Turb_lag2[3]
+         Turb_lag2 = lag(mean_Turbidity, 2),
+         FeCl_change = Ferric.chloride..mg.L - lag(Ferric.chloride..mg.L),
+         Polymer_change = Cationic.Polimer..mg.L - lag(Cationic.Polimer..mg.L))
 
-lm_fit <- lm(Ferric.chloride..mg.L ~ ., data=lagged_data |> select(-c(Timestamp)))
+# Get rid of first two observations (have NA values now)
+rwar_dosing_lag1 <- rwar_dosing_lag[3:nrow(lagged_data),]
+
+class_df <- rwar_dosing_lag1
+
+threshold_smallchange <- 0.05
+
+# normal data
+class_df$FeCl_change <- ifelse(
+    test = abs(class_df$FeCl_change) < threshold_smallchange,
+    yes = "same",
+    no = ifelse(
+        test = class_df$FeCl_change < 0,
+        yes = "dec",
+        no = "inc"
+    )
+)
+
+# polymer data
+class_df$Polymer_change <- ifelse(
+    test = abs(class_df$Polymer_change) < threshold_smallchange,
+    yes = "same",
+    no = ifelse(
+        test = class_df$Polymer_change < 0,
+        yes = "dec",
+        no = "inc"
+    )
+)
+
+write_csv(rwar_dosing_lag1, )
+# # Temp
+# lagged_data$Temp_lag1[1] <- lagged_data$Temp_lag1[2]
+# lagged_data$Temp_lag2[1:2] <- lagged_data$Temp_lag2[3]
+# # pH
+# lagged_data$pH_lag1[1] <- lagged_data$pH_lag1[2]
+# lagged_data$pH_lag2[1:2] <- lagged_data$pH_lag2[3]
+# # Conductivity
+# lagged_data$Cond_lag1[1] <- lagged_data$Cond_lag1[2]
+# lagged_data$Cond_lag2[1:2] <- lagged_data$Cond_lag2[3]
+# # Turbidity
+# lagged_data$Turb_lag1[1] <- lagged_data$Turb_lag1[2]
+# lagged_data$Turb_lag2[1:2] <- lagged_data$Turb_lag2[3]
+
+lm_fit <- lm(Ferric.chloride..mg.L ~ ., data=lagged_data1 |> select(-c(Timestamp)))
 summary(lm_fit)
@@
 ```
+
+```{r}
+
+### Add inc/dec/same for dose
+
+class_df <- lagged_data1
+
+# normal data
+class_df$FeCl_change <- ifelse(
+    test = abs(class_df$FeCl_change) < 0.01,
+    yes = "same",
+    no = ifelse(
+        test = class_df$FeCl_change < 0,
+        yes = "dec",
+        no = "inc"
+    )
+)
+
+# polymer data
+class_df$Polymer_change <- ifelse(
+    test = abs(class_df$Polymer_change) < 0.01,
+    yes = "same",
+    no = ifelse(
+        test = class_df$Polymer_change < 0,
+        yes = "dec",
+        no = "inc"
+    )
+)
+
+class_df
+```

### 2025-12-08 Data Cleaning.Rmd
Adjusted data, exported CSV of lagged data and dose change indicator.

@@
 
-threshold_smallchange <- 0.05
+threshold_smallchange <- 0.01
 
@@
 
-write_csv(rwar_dosing_lag1, )
+write_csv(class_df, "RWAR_dosing_lag_incdec.csv")
+
 # # Temp
@@
 ```
-
-```{r}
-
-### Add inc/dec/same for dose
-
-class_df <- lagged_data1
-
-# normal data
-class_df$FeCl_change <- ifelse(
-    test = abs(class_df$FeCl_change) < 0.01,
-    yes = "same",
-    no = ifelse(
-        test = class_df$FeCl_change < 0,
-        yes = "dec",
-        no = "inc"
-    )
-)
-
-# polymer data
-class_df$Polymer_change <- ifelse(
-    test = abs(class_df$Polymer_change) < 0.01,
-    yes = "same",
-    no = ifelse(
-        test = class_df$Polymer_change < 0,
-        yes = "dec",
-        no = "inc"
-    )
-)
-
-class_df
-```

### 2025-12-08 Data Cleaning.Rmd
Preliminary assessment for classification algorithm of dose change.

@@
 
-write_csv(class_df, "RWAR_dosing_lag_incdec.csv")
-
 # # Temp
@@
 ```
+
+```{r}
+write_csv(class_df, "RWAR_dosing_lag_incdec.csv")
+```
+
+### Classification
+
+```{r}
+
+df_lag_incdec <- read.csv("RWAR_dosing_lag_incdec.csv")
+
+df_lag_incdec$FeCl_change <- as.factor(df_lag_incdec$FeCl_change)
+df_lag_incdec$Polymer_change <- as.factor(df_lag_incdec$Polymer_change)
+
+summary(df_lag_incdec)
+
+df_lag_incdec <- df_lag_incdec |> 
+  mutate(Timestamp = ymd(df_lag_incdec$Timestamp))
+
+df_lag_incdec |>
+  select(-c(Timestamp)) |>
+  group_by(FeCl_change) |>
+  summarise(mean_Temp = mean(mean_Temp),
+            mean_pH = mean(mean_pH),
+            mean_cond = mean(mean_Conductivity),
+            mean_turbidity = mean(mean_Turbidity))
+
+# counts for FeCl change
+summary(df_lag_incdec$FeCl_change)
+# counts for polymer change
+summary(df_lag_incdec$Polymer_change)
+
+```

### 2025-12-08 Data Cleaning.Rmd

@@
 
-threshold_smallchange <- 0.01
+threshold_smallchange <- 0.05
 
@@
 
+write_csv(rwar_dosing_lag1, )
+
 # # Temp
@@
 ```{r}
-write_csv(class_df, "RWAR_dosing_lag_incdec.csv")
+
+### Add inc/dec/same for dose
+
+class_df <- lagged_data1
+
+# normal data
+class_df$FeCl_change <- ifelse(
+    test = abs(class_df$FeCl_change) < 0.01,
+    yes = "same",
+    no = ifelse(
+        test = class_df$FeCl_change < 0,
+        yes = "dec",
+        no = "inc"
+    )
+)
+
+# polymer data
+class_df$Polymer_change <- ifelse(
+    test = abs(class_df$Polymer_change) < 0.01,
+    yes = "same",
+    no = ifelse(
+        test = class_df$Polymer_change < 0,
+        yes = "dec",
+        no = "inc"
+    )
+)
+
+class_df
 ```

### 2025-12-08 Exploratory Analysis Ethan Schilling.Rmd

@@
 ```
+
+```{r}
+library(ggplot2)
+library(tidyr)
+library(dplyr)
+
+dosing_data <- read_csv("RWAR_dosing_cleaned.csv")
+
+dosing_long <- dosing_data %>%
+  pivot_longer(
+    cols = starts_with("mean_"),
+    names_to = "Characteristics",
+    values_to = "Value"
+  )
+
+ggplot(dosing_long, aes(x = Timestamp, y = Value, color = Ferric.chloride..mg.L)) +
+  geom_point(size = 0.7) +
+  facet_wrap(~Characteristics, scales = "free_y") +
+  scale_color_viridis_c(option = "plasma") +
+  theme_minimal() +
+  labs(
+    x = "Date",
+    y = "Value",
+    color = "Ferric Chloride (mg/L)",
+    title = "Water Quality Characteristics Over Time by Dosing"
+  )
+```

### 2025-12-08 Exploratory Analysis.Rmd

--- 2025-12-01 Exploratory Analysis.Rmd
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
 ```{r}
-library(dplyr)
-library(tidyr)
-library(lubridate)
-library(ggplot2)
-
-data = read.csv("RWAR_turb_cond_counts_pH_temp.csv")
-
-data$Timestamp = mdy_hm(data$Timestamp)
-
-# Trying to determine what rows have outliers.
-# The goal is to go by column, remove outliers & average that day's data, then ultimately combine every column after it's cleaned to create our fully cleaned data frame.
-# Then combine with alkalinity and hardness & coagulant doses data??
-
-ggplot(data, aes(x = Timestamp, y = Turbidity..NTU)) +
-  geom_point()
-
-data[data$Turbidity..NTU > 10, ] # row 2701 and 5446
-
-ggplot(data, aes(x = Timestamp, y = Conductivity..mS.cm.1)) +
-  geom_point()
-
-data[data$Conductivity..mS.cm.1 > 15, ] # row 5270
-
-# Not sure what to do with particle count (PC) data just yet
-
-ggplot(data, aes(x = Timestamp, y = pH..SU)) +
-  geom_point()
-
-ggplot(data, aes(x = Timestamp, y = Temperature..deg.C)) +
-  geom_point()
+print("GitHub test")
 

### 2025-12-08 Exploratory Analysis Reece C.Rmd
Exploratory graphs & clustering
why are these files untracked?

@@
 
+```
+
+```{r}
 cluster_data = merged %>%
@@
 
+names(full_merge) = c("Temperature", "pH", "Conductivity", "Turbidity", "Date", "Alkalinity",
+                      "Hardness", "Cluster", "Ferric Chloride, mg/L", "Cationic Polimer, mg/L")
+
 long_df = full_merge %>%
-  pivot_longer(cols = c("mean_Temp", "mean_pH", "mean_Conductivity", "mean_Turbidity",
-                        "alkalinity", "hardness", "Ferric.chloride..mg.L", "Cationic.Polimer..mg.L"),
+  pivot_longer(cols = c("Temperature", "pH", "Conductivity", "Turbidity",
+                      "Alkalinity", "Hardness", "Ferric Chloride, mg/L", "Cationic Polimer, mg/L"),
                names_to = "metric",

### 2025-12-08 Exploratory_Analysis_Josh_Tobey.Rmd
Working on GAM Fit.

@@
 
-coagual
+```{r}
+ggplot(exploratory_df) +
+  geom_point(aes(x = mean_pH,
+                 y = mean_Temp))
+```
+```{r}
+RWAR_dosing_lag_incdec <- read_csv("RWAR_dosing_lag_incdec.csv")
+head(RWAR_dosing_lag_incdec)
+
+```
+```{r}
+RWAR_dosing_lag_incdec <- RWAR_dosing_lag_incdec %>%
+  select(-c(FeCl_change, Polymer_change, Timestamp))
+head(RWAR_dosing_lag_incdec)
+```
+
+```{r}
+library(mgcv)
+
+lm_spec <- linear_reg()
+
+gam_formula <- Ferric.chloride..mg.L ~ s(Cationic.Polimer..mg.L) + s(mean_Temp) + s(mean_pH) + s(mean_Conductivity) + s(mean_Turbidity) + s(Temp_lag1) + s(Temp_lag2) + s(pH_lag1) + s(pH_lag2) + s(Cond_lag1) + s(Cond_lag2) + s(Turb_lag1) + s(Turb_lag2)
+
+gam_fit <- gam(gam_formula, data = RWAR_dosing_lag_incdec)
+gam_rec <- recipe(Ferric.chloride..mg.L ~ ., data = RWAR_dosing_lag_incdec) %>%
+  step_ns(Cationic.Polimer..mg.L, deg_free = 10) %>%
+  step_ns(mean_Temp, deg_free = 10) %>%
+  step_ns(mean_pH, deg_free = 10) %>%
+  step_ns(mean_Conductivity, deg_free = 10) %>%
+  step_ns(mean_Turbidity, deg_free = 10) %>%
+  step_ns(Temp_lag1, deg_free = 10) %>%
+  step_ns(Temp_lag2, deg_free = 10) %>%
+  step_ns(pH_lag1, deg_free = 10) %>%
+  step_ns(pH_lag2, deg_free = 10) %>%
+  step_ns(Cond_lag1, deg_free = 10) %>%
+  step_ns(Cond_lag2, deg_free = 10) %>%
+  step_ns(Turb_lag1, deg_free = 10) %>%
+  step_ns(Turb_lag2, deg_free = 10)
+
+gam_wf <- workflow() |>
+  add_model(lm_spec) |>
+  add_recipe(gam_rec)
+
+gam_fit <- gam_wf |>
+  fit(data = RWAR_dosing_lag_incdec)
+
+gam_fit |> tidy()
+gam_fit |> glance()
+
+dat_grid <- tibble(expand.grid(Cationic.Polimer..mg.L = seq(min(RWAR_dosing_lag_incdec$Cationic.Polimer..mg.L), 
+                               max(RWAR_dosing_lag_incdec$Cationic.Polimer..mg.L)),
+                               mean_Temp = seq(min(RWAR_dosing_lag_incdec$mean_Temp), max(RWAR_dosing_lag_incdec$mean_Temp)),
+                               mean_pH = seq(min(RWAR_dosing_lag_incdec$mean_pH), max(RWAR_dosing_lag_incdec$mean_pH)),
+                               mean_Conductivity = seq(min(RWAR_dosing_lag_incdec$mean_Conductivity), max(RWAR_dosing_lag_incdec$mean_Conductivity)),
+                               mean_Turbidity = seq(min(RWAR_dosing_lag_incdec$mean_Turbidity), max(RWAR_dosing_lag_incdec$mean_Turbidity)),
+                               Temp_lag1 = seq(min(RWAR_dosing_lag_incdec$Temp_lag1), max(RWAR_dosing_lag_incdec$Temp_lag1)),
+                               Temp_lag2 = seq(min(RWAR_dosing_lag_incdec$Temp_lag2), max(RWAR_dosing_lag_incdec$Temp_lag2)),
+                               pH_lag1 = seq(min(RWAR_dosing_lag_incdec$pH_lag1), max(RWAR_dosing_lag_incdec$pH_lag1)),
+                               pH_lag2 = seq(min(RWAR_dosing_lag_incdec$pH_lag2), max(RWAR_dosing_lag_incdec$pH_lag2)),
+                               Cond_lag1 = seq(min(RWAR_dosing_lag_incdec$Cond_lag1), max(RWAR_dosing_lag_incdec$Cond_lag1)),
+                               Cond_lag2 = seq(min(RWAR_dosing_lag_incdec$Cond_lag2), max(RWAR_dosing_lag_incdec$Cond_lag2)),
+                               Turb_lag1 = seq(min(RWAR_dosing_lag_incdec$Turb_lag1), max(RWAR_dosing_lag_incdec$Turb_lag1)),
+                               Turb_lag2 = seq(min(RWAR_dosing_lag_incdec$Turb_lag2), max(RWAR_dosing_lag_incdec$Turb_lag2))
+                               ))
+bind_cols(
+  augment(gam_fit, new_data = dat_grid),
+  predict(gam_fit, new_data = dat_grid, type = "conf_int")
+) -> dose_pred_gam
+
+for(pred in c("age", "year", "education")) {
+  pred_dat <- dose_pred_gam[, c(pred, ".pred", ".pred_lower", ".pred_upper")] |>
+    group_by_at(pred) |>
+    summarise(.pred = mean(.pred), .pred_lower = mean(.pred_lower), .pred_upper = mean(.pred_upper))
+  dose_dat <- RWAR_dosing_lag_incdec[, c(pred, "Ferric.chloride..mg.L")]
+
+  if(is.numeric(dose_dat |> pull(pred))) {
+    p <- dose_dat |>
+      ggplot() +
+      geom_point(aes_string(pred, "wage"), alpha = .3) +
+      geom_line(aes_string(pred, ".pred"), data = pred_dat, colour = "darkgreen") +
+      geom_line(aes_string(pred, ".pred_lower"), data = pred_dat, colour = "darkgreen", lty = 2) +
+      geom_line(aes_string(pred, ".pred_upper"), data = pred_dat, colour = "darkgreen", lty = 2)
+  } else {
+    p <- dose_dat |>
+      ggplot() +
+      geom_jitter(aes_string(pred, "wage"), alpha = .3) +
+      geom_point(aes_string(pred, ".pred"), data = pred_dat, colour = "darkgreen") +
+      geom_errorbar(aes_string(pred, ymin = ".pred_lower", ymax = ".pred_upper"), data = pred_dat, colour = "darkgreen")
+  }
+  print(p)
+}
+
+```

### 2025-12-08 Exploratory_Analysis_Josh_Tobey.Rmd

@@
 ```
-```{r}
-summary(exploratory_df)
-
-```
 
@@
 ```
-```{r}
-exploratory_df$Timestamp <- ymd(exploratory_df$Timestamp)
-exploratory_df <- exploratory_df[order(exploratory_df$Timestamp), ]
-ggplot(exploratory_df, aes(x = Timestamp,
-                           y = mean_Temp,
-                           color = mean_Temp)) +
-  geom_line() + 
-  scale_color_gradient(low = 'yellow', high = 'green')
 
-```
 ```{r}

### 2025-12-08 Data Cleaning.Rmd
Added notes on classification models to try.
Added code to pull dose from previous year.

@@
 
+# do regression tree
+# LDA/QDA
+# random forest
+
+# Look at performance of a model that simply pulls previous year dose
+
+closest_prev_dose <- df_lag_incdec |>
+  rowwise() |>
+  mutate(
+    target_time = Timestamp - years(1),
+    prev_year_dose = df_lag_incdec$value[which.min(abs(df_lag_incdec$Timestamp - target_time))]
+  ) |>
+  ungroup()
+
+closest_prev_dose
+
 ```

### 2025-12-09 Data Cleaning.Rmd
Added a single line to test pulling.
Attempted to pull previous year doses as a predictor.

--- 2025-12-08 Data Cleaning.Rmd
@@
 df_lag_incdec <- df_lag_incdec |> 
-  mutate(Timestamp = ymd(df_lag_incdec$Timestamp))
+  mutate(Timestamp = ymd(Timestamp))
 
@@
 ```
+
+```{r}
+# Look at performance of a model that simply pulls previous year dose
+
+dosing <- read_csv("CoagulantDoses_2017-20.csv")
+dosing <- dosing |> 
+  mutate(Timestamp = mdy(dosing$Date)) |>
+  select(-c(Date))
+
+df_lag_incdec_prevdose <- df_lag_incdec |>
+  rowwise() |>
+  mutate(
+    target_time = Timestamp - years(1),
+    FeCl_prev = dosing$Ferric.chloride..mg.L[which.min(abs(dosing$Timestamp - target_time))],
+    Polymer_prev = dosing$`Cationic Polimer, mg/L`[which.min(abs(dosing$Timestamp - target_time))]) |>
+  ungroup()
+
+```
+
+```{r}
+df_lag_incdec |>
+  rename(FeCl_dose = Ferric.chloride..mg.L,
+         Polymer_dose = Cationic.Polimer..mg.L)
+
+write.csv(df_lag_incdec, "RWAR_dosing_lag_incdec.csv")
+
+```

### 2025-12-09 Data Cleaning.Rmd
Changed column names for FeCl/polymer doses

@@
 # Look at performance of a model that simply pulls previous year dose
+# not working
 
@@
 ```{r}
-df_lag_incdec |>
+new_df <- df_lag_incdec |>
   rename(FeCl_dose = Ferric.chloride..mg.L,
@@
 
-write.csv(df_lag_incdec, "RWAR_dosing_lag_incdec.csv")
+write.csv(new_df, "RWAR_dosing_lag_incdec.csv")
 

### 2025-12-09 Data Cleaning.Rmd
Started into Random Forest model for classification

@@
 
-df_lag_incdec <- read.csv("RWAR_dosing_lag_incdec.csv")
-
+# Pull in data
+df_lag_incdec <- read.csv("RWAR_dosing_lag_incdec.csv") |> 
+  mutate(Timestamp = ymd(Timestamp))
 df_lag_incdec$FeCl_change <- as.factor(df_lag_incdec$FeCl_change)
@@
 
-summary(df_lag_incdec)
-
-df_lag_incdec <- df_lag_incdec |> 
-  mutate(Timestamp = ymd(Timestamp))
-
+# Look at average water characteristics for each FeCl dose change group
 df_lag_incdec |>
@@
   group_by(FeCl_change) |>
-  summarise(mean_Temp = mean(mean_Temp),
+  summarise(count = n(),
+            mean_Temp = mean(mean_Temp),
+            mean_pH = mean(mean_pH),
+            mean_cond = mean(mean_Conductivity),
+            mean_turbidity = mean(mean_Turbidity))
+
+# Look at average water characteristics for each polymer dose change group
+df_lag_incdec |>
+  select(-c(Timestamp)) |>
+  group_by(Polymer_change) |>
+  summarise(count = n(),
+            mean_Temp = mean(mean_Temp),
             mean_pH = mean(mean_pH),
@@
 
+```
+
+```{r}
+
+## TRAIN/TEST DATA
+train_df <- df_lag_incdec |> filter(year(Timestamp) <= 2019)
+test_df  <- df_lag_incdec |> filter(year(Timestamp) == 2020)
+
+## RANDOM FOREST
+
+rf_spec <- rand_forest(mtry = sqrt(.cols())) |>
+  set_engine("randomForest", importance = TRUE) |>
+  set_mode("classification")
+
+rf_fit <- rf_spec |>
+  fit(high_sales ~ ., data = df_train)
+
+vip(rf_fit) ## still Price and Shelf location are the most important
+
+rf_fit |>
+  augment(new_data = df_test) |>
+  accuracy(truth = high_sales, estimate = .pred_class)
+
+# 5. Compare the OOB confusion matrix to your test confusion matrix. [**Hint:** The `confusion` element of the model output fit is OOB.]
+
+## OOB:
+rf_fit |>
+  extract_fit_engine() %>%
+  .$confusion
+
+## test confusion
+rf_fit |>
+  augment(new_data = df_test) |>
+  conf_mat(truth = high_sales, estimate = .pred_class)
 ```
@@
 ```
-
-```{r}
-new_df <- df_lag_incdec |>
-  rename(FeCl_dose = Ferric.chloride..mg.L,
-         Polymer_dose = Cationic.Polimer..mg.L)
-
-write.csv(new_df, "RWAR_dosing_lag_incdec.csv")
-
-```

### 2025-12-09 Paper Draft.Rmd
Redid variable names for plotting section

@@
 
-rwar_dosing <- read.csv("RWAR_dosing_cleaned.csv")
-rwar_dosing$Timestamp <- ymd(rwar_dosing$Timestamp)
+df_lag_incdec <- read.csv("RWAR_dosing_lag_incdec.csv") |>
+  mutate(Timestamp = ymd(Timestamp))
 
-plot_pH <- ggplot(rwar_dosing, aes(x = Timestamp,
+plot_pH <- ggplot(df_lag_incdec, aes(x = Timestamp,
                            y = mean_pH,
@@
 
-plot_polymer <- ggplot(rwar_dosing, aes(x = Timestamp,
-                           y = Cationic.Polimer..mg.L,
-                           color = Cationic.Polimer..mg.L)) +
+plot_polymer <- ggplot(df_lag_incdec, aes(x = Timestamp,
+                           y = Polymer_dose,
+                           color = Polymer_dose)) +
   geom_line() + 
@@
 
-plot_FeCl <- ggplot(rwar_dosing, aes(x = Timestamp,
-                           y = Ferric.chloride..mg.L,
-                           color = Ferric.chloride..mg.L)) +
+plot_FeCl <- ggplot(df_lag_incdec, aes(x = Timestamp,
+                           y = FeCl_dose,
+                           color = FeCl_dose)) +
   geom_line() + 
@@
 
-plot_temp <- ggplot(rwar_dosing, aes(x = Timestamp,
+plot_temp <- ggplot(df_lag_incdec, aes(x = Timestamp,
                            y = mean_Temp,
@@
 
-plot_cond <- ggplot(rwar_dosing, aes(x = Timestamp,
+plot_cond <- ggplot(df_lag_incdec, aes(x = Timestamp,
                            y = mean_Conductivity,
@@
 
-plot_turb <- ggplot(rwar_dosing, aes(x = Timestamp,
+plot_turb <- ggplot(df_lag_incdec, aes(x = Timestamp,
                            y = mean_Turbidity,

### 2025-12-09 Data Cleaning.Rmd
Completed preliminary classification algorithm analysis via random forest, LDA, QDA.

@@
 knitr::opts_chunk$set(echo = TRUE)
+```
+
+```{r, include=F}
+library(ISLR) ## data package
+library(tidyverse) ## data manipulation
+library(tidymodels) ## tidy modeling
+library(knitr) ## tables
+library(rpart.plot) ## tree diagrams
+library(vip) ## plotting variable importance
+library(dplyr)
+library(tidyr)
+library(readr)
+library(lubridate)
+library(ggplot2)
+library(zoo)
+library(purrr)
+library(knitr)
+library(patchwork) ## plot arrangement
+library(parsnip)
+library(discrim)
 ```
@@
 
+## Set up a column of inc/dec for 
 class_df <- rwar_dosing_lag1
-
-threshold_smallchange <- 0.01
+threshold_smallchange <- 0.01 # cutoff for dose changes that are approx. zero
 
@@
 )
-
-# # Temp
-# lagged_data$Temp_lag1[1] <- lagged_data$Temp_lag1[2]
-# lagged_data$Temp_lag2[1:2] <- lagged_data$Temp_lag2[3]
-# # pH
-# lagged_data$pH_lag1[1] <- lagged_data$pH_lag1[2]
-# lagged_data$pH_lag2[1:2] <- lagged_data$pH_lag2[3]
-# # Conductivity
-# lagged_data$Cond_lag1[1] <- lagged_data$Cond_lag1[2]
-# lagged_data$Cond_lag2[1:2] <- lagged_data$Cond_lag2[3]
-# # Turbidity
-# lagged_data$Turb_lag1[1] <- lagged_data$Turb_lag1[2]
-# lagged_data$Turb_lag2[1:2] <- lagged_data$Turb_lag2[3]
-
-lm_fit <- lm(Ferric.chloride..mg.L ~ ., data=lagged_data1 |> select(-c(Timestamp)))
-summary(lm_fit)
 
@@
 ```{r}
+## SET SEED
+set.seed(445)
 
-## TRAIN/TEST DATA
-train_df <- df_lag_incdec |> filter(year(Timestamp) <= 2019)
-test_df  <- df_lag_incdec |> filter(year(Timestamp) == 2020)
+## TRAIN/TEST DATA (2018-2019 and 2020)
+df_full <- df_lag_incdec |> select(-c(FeCl_dose, Polymer_dose))
+df_train <- df_full |> filter(year(Timestamp) <= 2019)
+df_test  <- df_full |> filter(year(Timestamp) == 2020)
+```
+
+```{r}
+set.seed(445)
 
 ## RANDOM FOREST
-
 rf_spec <- rand_forest(mtry = sqrt(.cols())) |>
@@
 
-rf_fit <- rf_spec |>
-  fit(high_sales ~ ., data = df_train)
+# Different model fits
+rf_fit1 <- rf_spec |>
+  fit(FeCl_change ~ ., data = df_train |> 
+        select(-c(Polymer_change, Timestamp)))
+rf_fit2 <- rf_spec |>
+  fit(FeCl_change ~ mean_Temp + mean_pH + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2, 
+      data = df_train)
+rf_fit3 <- rf_spec |>
+  fit(FeCl_change ~ mean_pH + mean_Temp, 
+      data = df_train)
 
-vip(rf_fit) ## still Price and Shelf location are the most important
+# Model fit to take performance metrics of:
+rf_fit <- rf_fit3
 
+# Most important predictors
+vip(rf_fit)
+
+## Accuracy
 rf_fit |>
   augment(new_data = df_test) |>
-  accuracy(truth = high_sales, estimate = .pred_class)
-
-# 5. Compare the OOB confusion matrix to your test confusion matrix. [**Hint:** The `confusion` element of the model output fit is OOB.]
-
-## OOB:
-rf_fit |>
-  extract_fit_engine() %>%
-  .$confusion
-
-## test confusion
+  accuracy(truth = FeCl_change, estimate = .pred_class)
+## Test confusion matrix
 rf_fit |>
   augment(new_data = df_test) |>
-  conf_mat(truth = high_sales, estimate = .pred_class)
+  conf_mat(truth = FeCl_change, estimate = .pred_class)
+
+## Notes
+# 34% accuracy for all predictors
+# 47% accuracy for just temperature and pH
+
+```
+
+```{r}
+## set seed
+set.seed(445)
+
+## LDA
+lda_spec <- discrim_linear()
+lda_fit1 <- lda_spec |>
+  fit(FeCl_change ~ ., data = df_train |> 
+        select(-c(Polymer_change, Timestamp)))
+lda_fit2 <- lda_spec |>
+  fit(FeCl_change ~ mean_pH + mean_Temp, data = df_train)
+
+# select LDA to take metrics of
+lda_fit <- lda_fit2
+
+# Test confusion matrix
+lda_fit |> 
+  augment(new_data = df_test) |>
+  conf_mat(truth = FeCl_change, estimate = .pred_class)
+
+# Test accuracy
+lda_fit |>
+  augment(new_data = df_test) |>
+  accuracy(truth = FeCl_change, estimate = .pred_class)
+
+## Notes
+# 39% test accuracy on all predictors
+# 41% test accuracy on temp and pH
+
+## QDA
+qda_spec <- discrim_quad()
+qda_fit1 <- qda_spec |>
+  fit(FeCl_change ~ ., data = df_train |> 
+        select(-c(Polymer_change, Timestamp)))
+qda_fit2 <- qda_spec |>
+  fit(FeCl_change ~ mean_pH + mean_Temp, data = df_train)
+
+# select LDA to take metrics of
+qda_fit <- qda_fit2
+
+# Test confusion matrix
+qda_fit |> 
+  augment(new_data = df_test) |>
+  conf_mat(truth = FeCl_change, estimate = .pred_class)
+
+# Test accuracy
+qda_fit |>
+  augment(new_data = df_test) |>
+  accuracy(truth = FeCl_change, estimate = .pred_class)
+
+## Notes
+# 39.5% test accuracy on all predictors
+# 43.7% test accuracy on temp and pH
+
 ```

### 2025-12-09 Tree Models Reece C.Rmd
Decision tree & random forest

---
title: "Tree Models"
output: pdf_document
---

```{r}
set.seed(445)

library(dplyr)
library(tidyverse)
library(tidymodels)

# Decision Tree

# For now just predicting numerical amounts, not classifying
data = read.csv("RWAR_dosing_lag_incdec.csv")
data = as.data.frame(data)
data$FeCl_change = NULL
data$Polymer_change = NULL

# Set up training & testing data

# Can change these if you want to modify the tree/data
tree_depth = 5 # tree depth
min_n = 50 # minimum observations needed to further split
train_proportion = 0.8 # split of data

n = nrow(data)
indices = sample(1:n, size = floor(train_proportion * n))

train_data = data[indices, ]
test_data  = data[-indices, ]

# Create the tree

recipe = recipe(FeCl_dose ~ ., data = train_data) %>% # Change target variable here
  step_rm(Timestamp) %>%
  step_rm(Polymer_dose) # Remove the other target variable here

tree_spec = decision_tree(tree_depth = tree_depth, min_n = min_n) %>%
  set_engine("rpart") %>%
  set_mode("regression")

workflow = workflow() %>%
  add_recipe(recipe) %>%
  add_model(tree_spec)

fit_tree = workflow %>%
  fit(data = train_data)

# Assess accuracy

predictions = fit_tree %>%
  predict(test_data) %>%
  bind_cols(test_data)

actual = predictions$`FeCl_dose` # Change target variable here
predicted = predictions$.pred

rmse = sqrt(mean((actual - predicted)^2))
print(paste("RMSE: ", rmse))

r_squared = cor(actual, predicted)^2
print(paste("R Squared: ", r_squared))

# Plot the tree

library(rpart.plot)

fit_tree %>%
  extract_fit_engine() %>%
  rpart.plot(roundint = FALSE)

```

```{r}
set.seed(445)

library(dplyr)
library(tidyverse)
library(tidymodels)

# Random Forest

# For now just predicting numerical amounts, not classifying
data = read.csv("RWAR_dosing_lag_incdec.csv")
data = as.data.frame(data)
data$FeCl_change = NULL
data$Polymer_change = NULL

# Set up training & testing data

# Can change these if you want to modify the tree/data
trees = 500 # number of trees in the forest
min_n = 50 # minimum observations needed to further split
train_proportion = 0.8 # split of data

n = nrow(data)
indices = sample(1:n, size = floor(train_proportion * n))

train_data = data[indices, ]
test_data  = data[-indices, ]

# Create the tree

recipe = recipe(FeCl_dose ~ ., data = train_data) %>% # Change target variable here
  step_rm(Timestamp) %>%
  step_rm(Polymer_dose) # Remove the other target variable here

rf_spec = rand_forest(trees = trees, min_n = min_n) %>%
  set_engine("ranger") %>%
  set_mode("regression")

workflow = workflow() %>%
  add_recipe(recipe) %>%
  add_model(rf_spec)

fit_rf = workflow %>%
  fit(data = train_data)

# Assess accuracy

predictions = fit_rf %>%
  predict(test_data) %>%
  bind_cols(test_data)

actual = predictions$`FeCl_dose` # Change target variable here
predicted = predictions$.pred

rmse = sqrt(mean((actual - predicted)^2))
print(paste("RMSE: ", rmse))

r_squared = cor(actual, predicted)^2
print(paste("R Squared: ", r_squared))

```

### 2025-12-09 Exploratory_Analysis_Josh_Tobey.Rmd
Updated version of GAM so far for each predictor.

--- 2025-12-05 Exploratory_Analysis_Josh_Tobey.Rmd
@@
 
-coagual
+```{r}
+ggplot(exploratory_df) +
+  geom_point(aes(x = mean_pH,
+                 y = mean_Temp))
+```
+```{r}
+RWAR_dosing_lag_incdec <- read_csv("RWAR_dosing_lag_incdec.csv")
+head(RWAR_dosing_lag_incdec)
+
+```
+```{r}
+RWAR_dosing_lag_incdec <- RWAR_dosing_lag_incdec %>%
+  select(-c(FeCl_change, Polymer_change, Timestamp))
+head(RWAR_dosing_lag_incdec)
+```
+
+```{r}
+library(mgcv)
+
+lm_spec <- linear_reg()
+
+gam_formula <- Ferric.chloride..mg.L ~ s(Cationic.Polimer..mg.L) + s(mean_Temp) + s(mean_pH) + s(mean_Conductivity) + s(mean_Turbidity) + s(Temp_lag1) + s(Temp_lag2) + s(pH_lag1) + s(pH_lag2) + s(Cond_lag1) + s(Cond_lag2) + s(Turb_lag1) + s(Turb_lag2)
+
+gam_fit <- gam(gam_formula, data = RWAR_dosing_lag_incdec)
+gam_rec <- recipe(Ferric.chloride..mg.L ~ ., data = RWAR_dosing_lag_incdec) %>%
+  step_ns(Cationic.Polimer..mg.L, deg_free = 10) %>%
+  step_ns(mean_Temp, deg_free = 10) %>%
+  step_ns(mean_pH, deg_free = 10) %>%
+  step_ns(mean_Conductivity, deg_free = 10) %>%
+  step_ns(mean_Turbidity, deg_free = 10) %>%
+  step_ns(Temp_lag1, deg_free = 10) %>%
+  step_ns(Temp_lag2, deg_free = 10) %>%
+  step_ns(pH_lag1, deg_free = 10) %>%
+  step_ns(pH_lag2, deg_free = 10) %>%
+  step_ns(Cond_lag1, deg_free = 10) %>%
+  step_ns(Cond_lag2, deg_free = 10) %>%
+  step_ns(Turb_lag1, deg_free = 10) %>%
+  step_ns(Turb_lag2, deg_free = 10)
+
+gam_wf <- workflow() |>
+  add_model(lm_spec) |>
+  add_recipe(gam_rec)
+
+gam_fit <- gam_wf |>
+  fit(data = RWAR_dosing_lag_incdec)
+
+gam_fit |> tidy()
+gam_fit |> glance()
+
+dat_grid_dose <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Polymer_dose = seq(min(RWAR_dosing_lag_incdec$Polymer_dose),
+                                      max(RWAR_dosing_lag_incdec$Polymer_dose),
+                                      length.out = 100))
+dat_grid_temp <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(mean_Temp = seq(min(RWAR_dosing_lag_incdec$mean_Temp), 
+                         max(RWAR_dosing_lag_incdec$mean_Temp),
+                                      length.out = 100))
+dat_grid_pH <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(mean_pH = seq(min(RWAR_dosing_lag_incdec$mean_pH), max(RWAR_dosing_lag_incdec$mean_pH),
+                                      length.out = 100))
+
+dat_grid_cond <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(mean_Conductivity = seq(min(RWAR_dosing_lag_incdec$mean_Conductivity), max(RWAR_dosing_lag_incdec$mean_Conductivity),
+                                      length.out = 100))
+dat_grid_turb <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(mean_Turbidity = seq(min(RWAR_dosing_lag_incdec$mean_Turbidity), max(RWAR_dosing_lag_incdec$mean_Turbidity),
+                                      length.out = 100))
+dat_grid_temp_lag1 <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Temp_lag1 = seq(min(RWAR_dosing_lag_incdec$Temp_lag1), max(RWAR_dosing_lag_incdec$Temp_lag1),
+                                      length.out = 100))
+dat_grid_temp_lag2 <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Temp_lag2 = seq(min(RWAR_dosing_lag_incdec$Temp_lag2), max(RWAR_dosing_lag_incdec$Temp_lag2),
+                                      length.out = 100))
+dat_grid_pH_lag1 <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(pH_lag1 = seq(min(RWAR_dosing_lag_incdec$pH_lag1), max(RWAR_dosing_lag_incdec$pH_lag1),
+                                      length.out = 100))
+dat_grid_pH_lag2 <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(pH_lag2 = seq(min(RWAR_dosing_lag_incdec$pH_lag2), max(RWAR_dosing_lag_incdec$pH_lag2),
+                                      length.out = 100))
+dat_grid_cond_lag1 <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Cond_lag1 = seq(min(RWAR_dosing_lag_incdec$Cond_lag1), max(RWAR_dosing_lag_incdec$Cond_lag1),
+                                      length.out = 100))
+dat_grid_cond_lag2 <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Cond_lag2 = seq(min(RWAR_dosing_lag_incdec$Cond_lag2), max(RWAR_dosing_lag_incdec$Cond_lag2),
+                                      length.out = 100))
+dat_grid_turb_lag1 <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Turb_lag1 = seq(min(RWAR_dosing_lag_incdec$Turb_lag1), max(RWAR_dosing_lag_incdec$Turb_lag1),
+                                      length.out = 100))
+dat_grid_turb_lag2 <- RWAR_dosing_lag_incdec |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Turb_lag2 = seq(min(RWAR_dosing_lag_incdec$Turb_lag2), max(RWAR_dosing_lag_incdec$Turb_lag2),
+                                      length.out = 100))
+
+dose_pred <- augment(gam_fit, new_data = dat_grid_dose)
+temp_pred <- augment(gam_fit, new_data = dat_grid_temp)
+pH_pred <- augment(gam_fit, new_data = dat_grid_pH)
+cond_pred <- augment(gam_fit, new_data = dat_grid_cond)
+turb_pred <- augment(gam_fit, new_data = dat_grid_turb)
+tempLag1_pred <- augment(gam_fit, new_data = dat_grid_temp_lag1)
+tempLag2_pred <- augment(gam_fit, new_data = dat_grid_temp_lag2)
+pHLag1_pred <- augment(gam_fit, new_data = dat_grid_pH_lag1)
+pHLag2_pred <- augment(gam_fit, new_data = dat_grid_pH_lag2)
+condLag1_pred <- augment(gam_fit, new_data = dat_grid_cond_lag1)
+condLag2_pred <- augment(gam_fit, new_data = dat_grid_cond_lag2)
+turbLag1_pred <- augment(gam_fit, new_data = dat_grid_turb_lag1)
+turbLag2_pred <- augment(gam_fit, new_data = dat_grid_turb_lag2)
+
+```
+
+```{r}
+for(pred in c("age", "year", "education")) {
+  pred_dat <- dose_pred_gam[, c(pred, ".pred", ".pred_lower", ".pred_upper")] |>
+    group_by_at(pred) |>
+    summarise(.pred = mean(.pred), .pred_lower = mean(.pred_lower), .pred_upper = mean(.pred_upper))
+  dose_dat <- RWAR_dosing_lag_incdec[, c(pred, "Ferric.chloride..mg.L")]
+
+  if(is.numeric(dose_dat |> pull(pred))) {
+    p <- dose_dat |>
+      ggplot() +
+      geom_point(aes_string(pred, "wage"), alpha = .3) +
+      geom_line(aes_string(pred, ".pred"), data = pred_dat, colour = "darkgreen") +
+      geom_line(aes_string(pred, ".pred_lower"), data = pred_dat, colour = "darkgreen", lty = 2) +
+      geom_line(aes_string(pred, ".pred_upper"), data = pred_dat, colour = "darkgreen", lty = 2)
+  } else {
+    p <- dose_dat |>
+      ggplot() +
+      geom_jitter(aes_string(pred, "wage"), alpha = .3) +
+      geom_point(aes_string(pred, ".pred"), data = pred_dat, colour = "darkgreen") +
+      geom_errorbar(aes_string(pred, ymin = ".pred_lower", ymax = ".pred_upper"), data = pred_dat, colour = "darkgreen")
+  }
+  print(p)
+}
+```

### 2025-12-09 Data Cleaning.Rmd
Added multinomial classification.

@@
 library(discrim)
+library(nnet)
+library(broom)
+library(dplyr)
+library(kableExtra)
+
 ```
@@
 ```{r}
+
+## MULTINOMIAL REGRESSION
+multi_model <- multinom_reg() |>
+  set_engine("nnet") |>
+  set_mode("classification")
+
+multi_recipe1 <- recipe(FeCl_change ~ mean_pH + mean_Temp, data = df_train) |>
+  step_normalize(all_numeric_predictors()) |>
+  step_dummy(all_nominal_predictors())
+
+multi_recipe <- multi_recipe1
+
+multi_workflow <- workflow() |>
+  add_recipe(multi_recipe) |>
+  add_model(multi_model)
+
+multi_fit <- fit(multi_workflow, data = df_train)
+
+df_pred_class <- predict(multi_fit, new_data = df_test, type = "class") |>
+  bind_cols(df_test |> select(FeCl_change))
+df_pred_probs <- predict(multi_fit, new_data = df_test, type = "prob") |>
+  bind_cols(df_test |> select(FeCl_change))
+
+accuracy_result <- metrics(df_pred_class, truth = FeCl_change, estimate = .pred_class)
+confusion_matrix <- conf_mat(df_pred_class, truth = FeCl_change, estimate = .pred_class)
+
+# results
+print(accuracy_result)
+print(confusion_matrix)
+
+library(broom)
+library(dplyr)
+library(knitr)    # For kable()
+library(kableExtra) # For extra formatting
+
+# 1. Extract the model fit object
+fit_engine <- multi_fit |> extract_fit_engine()
+
+# 2. Tidy the results using the broom package
+# The tidy function converts the coefficient matrix into a clean data frame.
+tidy_coeffs <- tidy(fit_engine)
+
+# 3. Format the table using kable/kableExtra for display
+tidy_coeffs |>
+  # Round all numeric columns for cleaner display
+  mutate(across(where(is.numeric), ~ round(., 4))) |>
+
+  # Rename columns for clarity (optional, but helpful)
+  rename(
+    Category = y.level,
+    Predictor = term,
+    Coefficient = estimate,
+    Std_Error = std.error,
+    Z_Value = statistic,
+    P_Value = p.value
+  ) |>
+
+  # Create a nicely formatted Markdown table
+  knitr::kable("markdown")
+  # OR use kableExtra for more styling options (e.g., HTML/PDF output)
+  # kableExtra::kbl(format = "html") |>
+  # kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
+multi_fit |> extract_fit_engine() |> tibble()
+```
+
+```{r}
 # Look at performance of a model that simply pulls previous year dose

### 2025-12-09 Data Cleaning.Rmd
updated multinomial regression model

@@
 
-multi_recipe1 <- recipe(FeCl_change ~ mean_pH + mean_Temp, data = df_train) |>
+multi_recipe1 <- recipe(FeCl_change ~ mean_Temp + Temp_lag1 + Temp_lag2 + mean_pH + pH_lag1 + pH_lag2 + mean_Conductivity + Cond_lag1 + Cond_lag2, data = df_train) |>
   step_normalize(all_numeric_predictors()) |>
@@
 
-accuracy_result <- metrics(df_pred_class, truth = FeCl_change, estimate = .pred_class)
+accuracy_result <- metrics(df_pred_class, truth = FeCl_change, estimate = .pred_class) |> 
+  filter(.metric == "accuracy")
 confusion_matrix <- conf_mat(df_pred_class, truth = FeCl_change, estimate = .pred_class)
@@
 
-library(broom)
-library(dplyr)
-library(knitr)    # For kable()
-library(kableExtra) # For extra formatting
+multi_fit |> extract_fit_engine()
 
-# 1. Extract the model fit object
-fit_engine <- multi_fit |> extract_fit_engine()
+## NOTES
+# turbidity is highly variable and does not help
 
-# 2. Tidy the results using the broom package
-# The tidy function converts the coefficient matrix into a clean data frame.
-tidy_coeffs <- tidy(fit_engine)
-
-# 3. Format the table using kable/kableExtra for display
-tidy_coeffs |>
-  # Round all numeric columns for cleaner display
-  mutate(across(where(is.numeric), ~ round(., 4))) |>
-
-  # Rename columns for clarity (optional, but helpful)
-  rename(
-    Category = y.level,
-    Predictor = term,
-    Coefficient = estimate,
-    Std_Error = std.error,
-    Z_Value = statistic,
-    P_Value = p.value
-  ) |>
-
-  # Create a nicely formatted Markdown table
-  knitr::kable("markdown")
-  # OR use kableExtra for more styling options (e.g., HTML/PDF output)
-  # kableExtra::kbl(format = "html") |>
-  # kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
-multi_fit |> extract_fit_engine() |> tibble()
 ```

### 2025-12-09 Exploratory Analysis Ethan SchillingREAL.Rmd
Hopefully it works

---
title: "Exploratory Analysis Ethan Schilling"
output: html_document
---

```{r}
library(tidyverse)
library(knitr)
library(tidymodels)
library(dplyr)
library(Metrics)
```

```{r}

TDS_unfilt <- read_csv("~/DSCI445/project-4/RWAR_TDS.csv")
TOC_unfilt <- read_csv("~/DSCI445/project-4/RWAR_TOC.csv")
DOC_unfilt <- read_csv("~/DSCI445/project-4/RWAR_DOC.csv")

df_TDS <- TDS_unfilt %>%
  rename(date = 1, value = 2) %>%
  mutate(date = as.Date(date, format = "%d-%b-%y"))

TDS_plot <- ggplot(df_TDS, aes(x = date, y = value)) +
  geom_line() +
  labs(x = "Date", y = "Total Dissolved Solids mg/L", title = "Total Dissolved Solids over Time")

df_TOC <- TOC_unfilt %>% 
  rename(date = 1, value = 2) %>%
  mutate(date = as.Date(date, format = "%d-%b-%y"))

TOC_plot <- ggplot(df_TOC, aes(x = date, y = value)) +
  geom_line() +
  labs(x = "Date", y = "Total Organinc Carbon mg/L", title = "Total Organic Carbon over Time")

df_DOC <- DOC_unfilt %>%
  rename(date = 1, value = 2) %>%
  mutate(date = as.Date(date, format = "%d-%b-%y"))

DOC_plot <- ggplot(df_DOC, aes(x = date, y = value)) +
  geom_line() +
  labs(x = "Date", y = "Dissolved Organic Carbon mg/L", title = "Dissolved Organic Carbon over Time")

saveRDS(TDS_plot,"TDS_plot.rds")
saveRDS(TOC_plot,"TOC_plot.rds")
saveRDS(DOC_plot,"DOC_plot.rds")

TDS_plot
TOC_plot
DOC_plot

getwd()
```

```{r}
library(lubridate)
library(dplyr)
library(ggplot2)
library(tidyr)
library(Metrics)

dosing_data <- read.csv("~/DSCI445/project-4/RWAR_dosing_cleaned.csv")

dosing_long <- dosing_data %>%
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "Characteristics",
    values_to = "Value"
  )

ggplot(dosing_long, aes(x = Timestamp, y = Value, color = Ferric.chloride..mg.L)) +
  geom_point(size = 0.7) +
  facet_wrap(~Characteristics, scales = "free_y") +
  scale_color_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(
    x = "Date",
    y = "Value",
    color = "Ferric Chloride (mg/L)",
    title = "Water Quality Characteristics Over Time by Dosing"
  )

RWAR_dosing <- read.csv("~/DSCI445/project-4/RWAR_dosing_lag_incdec.csv")

RWAR_dosing

RWAR_dosing <- RWAR_dosing %>% 
  mutate(Timestamp = ymd(Timestamp),
         Year = year(Timestamp))

RWAR_dosing

train <- RWAR_dosing %>% filter(Year %in% c(2018, 2019))
test  <- RWAR_dosing %>% filter(Year == 2020)

# 1 Interaction Term
model1 <- lm(FeCl_dose ~ mean_Temp * mean_Turbidity + mean_pH + Temp_lag1 + Turb_lag1, 
             data = train)
summary(model1)

test$pred_model1 <- predict(model1, newdata = test)

rmse(test$FeCl_dose, test$pred_model1)
mae(test$FeCl_dose, test$pred_model1)

# 2 Interaction Terms
model2 <- lm(FeCl_dose ~ mean_Temp * mean_pH +mean_Conductivity * mean_Turbidity + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2, data = train)

test$pred_model2 <- predict(model2, newdata = test)

rmse(test$FeCl_dose, test$pred_model2)
mae(test$FeCl_dose, test$pred_model2)

# Step Up

response_var <- "FeCl_dose" 
predictors <- c("mean_Temp", "mean_pH", "mean_Conductivity",
                "mean_Turbidity", "Temp_lag1", "Temp_lag2",
                "pH_lag1", "pH_lag2", "Cond_lag1", "Cond_lag2",
                "Turb_lag1", "Turb_lag2")

```

### 2025-12-09 Exploratory Analysis Ethan SchillingREAL.Rmd
WORKRKKKKRKR
added some more models with RMSE and MAE

@@
 # 1 Interaction Term
-model1 <- lm(FeCl_dose ~ mean_Temp * mean_Turbidity + mean_pH + Temp_lag1 + Turb_lag1, 
+
+model1 <- lm(FeCl_dose ~ mean_Temp * mean_pH + mean_Turbidity + Temp_lag1 + Turb_lag1, 
              data = train)
@@
 
-rmse(test$FeCl_dose, test$pred_model1)
-mae(test$FeCl_dose, test$pred_model1)
+cat("Model 1 RMSE:", rmse(test$FeCl_dose, test$pred_model1), "\n")
+cat("Model 1 MAE :", mae(test$FeCl_dose, test$pred_model1), "\n")
 
 # 2 Interaction Terms
-model2 <- lm(FeCl_dose ~ mean_Temp * mean_pH +mean_Conductivity * mean_Turbidity + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2, data = train)
+
+model2 <- lm(FeCl_dose ~ mean_Temp * mean_pH + mean_Conductivity * mean_Turbidity + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2, data = train)
 
@@
 
-rmse(test$FeCl_dose, test$pred_model2)
-mae(test$FeCl_dose, test$pred_model2)
+cat("Model 2 RMSE:", rmse(test$FeCl_dose, test$pred_model2), "\n")
+cat("Model 2 MAE :", mae(test$FeCl_dose, test$pred_model2), "\n")
 
-# Step Up
+model3 <- lm(FeCl_dose ~ mean_Temp * mean_Conductivity + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2, data = train)
 
-response_var <- "FeCl_dose" 
-predictors <- c("mean_Temp", "mean_pH", "mean_Conductivity",
-                "mean_Turbidity", "Temp_lag1", "Temp_lag2",
-                "pH_lag1", "pH_lag2", "Cond_lag1", "Cond_lag2",
-                "Turb_lag1", "Turb_lag2")
+test$pred_model3 <- predict(model3, newdata = test)
+
+cat("Model 3 RMSE:", rmse(test$FeCl_dose, test$pred_model3), "\n")
+cat("Model 3 MAE :", mae(test$FeCl_dose, test$pred_model3), "\n")
+
+# 2 Interactions: Cond and PH and Temp and Cond
+
+model4 <- lm(FeCl_dose ~ mean_Conductivity * mean_pH + mean_Temp * mean_Conductivity + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2, data = train)
+
+test$pred_model4 <- predict(model4, newdata = test)
+
+cat("Model 4 RMSE:", rmse(test$FeCl_dose, test$pred_model4), "\n")
+cat("Model 4 MAE :", mae(test$FeCl_dose, test$pred_model4), "\n")
 

### 2025-12-09 Paper Draft.Rmd
Added dose change graph.

@@
 
+```{r}
+
+df_lag_incdec_long <- df_lag_incdec |>
+  pivot_longer(
+    cols = starts_with("mean_"),
+    names_to = "Characteristics",
+    values_to = "Value"
+  )
+
+ggplot(df_lag_incdec_long, aes(x = Timestamp, y = Value, color = FeCl_dose)) +
+  geom_point(size = 0.7) +
+  facet_wrap(~Characteristics, scales = "free_y") +
+  scale_color_viridis_c(option = "plasma") +
+  theme_minimal() +
+  labs(
+    x = "Date",
+    y = "Value",
+    color = "Ferric Chloride (mg/L)",
+    title = "Water Quality Characteristics Over Time Colored by Dosing"
+  )
+
+ggplot(df_lag_incdec_long, aes(x = Timestamp, y = Value, color = FeCl_change)) +
+  geom_point(size = 0.7) +
+  facet_wrap(~Characteristics, scales = "free_y") +
+  theme_minimal() +
+  labs(
+    x = "Date",
+    y = "Value",
+    color = "FeCl_change",
+    shape = "FeCl_change",
+    title = "Water Quality Characteristics Colored by Dose Change"
+  )
+
+```
+
 ### Predictor Selection

### 2025-12-09 Exploratory_Analysis_Josh_Tobey.Rmd
Preliminary fitting of GAM.

@@
 ```{r}
-time_lag_1 <- 
-time_lag_2 <- 
-time_lag_3 <-
-
 weekly_log_spec <- logistic_reg()
@@
 library(mgcv)
+library(parsnip)
 
-lm_spec <- linear_reg()
-
-gam_formula <- Ferric.chloride..mg.L ~ s(Cationic.Polimer..mg.L) + s(mean_Temp) + s(mean_pH) + s(mean_Conductivity) + s(mean_Turbidity) + s(Temp_lag1) + s(Temp_lag2) + s(pH_lag1) + s(pH_lag2) + s(Cond_lag1) + s(Cond_lag2) + s(Turb_lag1) + s(Turb_lag2)
-
-gam_fit <- gam(gam_formula, data = RWAR_dosing_lag_incdec)
-gam_rec <- recipe(Ferric.chloride..mg.L ~ ., data = RWAR_dosing_lag_incdec) %>%
-  step_ns(Cationic.Polimer..mg.L, deg_free = 10) %>%
-  step_ns(mean_Temp, deg_free = 10) %>%
-  step_ns(mean_pH, deg_free = 10) %>%
-  step_ns(mean_Conductivity, deg_free = 10) %>%
-  step_ns(mean_Turbidity, deg_free = 10) %>%
-  step_ns(Temp_lag1, deg_free = 10) %>%
-  step_ns(Temp_lag2, deg_free = 10) %>%
-  step_ns(pH_lag1, deg_free = 10) %>%
-  step_ns(pH_lag2, deg_free = 10) %>%
-  step_ns(Cond_lag1, deg_free = 10) %>%
-  step_ns(Cond_lag2, deg_free = 10) %>%
-  step_ns(Turb_lag1, deg_free = 10) %>%
-  step_ns(Turb_lag2, deg_free = 10)
+gam_spec <- gen_additive_mod() |> 
+  set_engine("mgcv") |> 
+  set_mode("regression")
+gam_rec <- recipe(FeCl_dose ~ Polymer_dose + mean_Temp + mean_pH + mean_Conductivity +
+                 mean_Turbidity + Temp_lag1 + Temp_lag2 +
+                 pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2 +
+                 Turb_lag1 + Turb_lag2, data = RWAR_dosing_lag_incdec)
 
 gam_wf <- workflow() |>
-  add_model(lm_spec) |>
+  add_model(gam_spec) |>
   add_recipe(gam_rec)
 
-gam_fit <- gam_wf |>
-  fit(data = RWAR_dosing_lag_incdec)
+gam_fit <- gam(
+  FeCl_dose ~ s(Polymer_dose) + s(mean_Temp) + s(mean_pH) + s(mean_Conductivity) +
+               s(mean_Turbidity) + s(Temp_lag1) + s(Temp_lag2) +
+               s(pH_lag1) + s(pH_lag2) + s(Cond_lag1) + s(Cond_lag2) +
+               s(Turb_lag1) + s(Turb_lag2),
+  data = RWAR_dosing_lag_incdec
+)
 
@@
 gam_fit |> glance()
+```
+```{r}
 
@@
 
-dose_pred <- augment(gam_fit, new_data = dat_grid_dose)
-temp_pred <- augment(gam_fit, new_data = dat_grid_temp)
+pred1 <- predict(gam_fit, newdata = dat_grid_dose, type = "response", se.fit = TRUE)
+dose_pred <- dat_grid_dose |> mutate(
+  .fitted = pred1$fit,
+  .lower = pred1$fit - 2 * pred1$se.fit,
+  .upper = pred1$fit + 2 * pred1$se.fit
+  )
+pred2 <- predict(gam_fit, newdata = dat_grid_temp, type = "response", se.fit = TRUE)
+temp_pred <- dat_grid_temp |> mutate(
+  .fitted = pred2$fit,
+  .lower = pred2$fit - 2 * pred2$se.fit,
+  .upper = pred2$fit + 2 * pred2$se.fit
+)
 pH_pred <- augment(gam_fit, new_data = dat_grid_pH)
@@
 ```{r}
-for(pred in c("age", "year", "education")) {
-  pred_dat <- dose_pred_gam[, c(pred, ".pred", ".pred_lower", ".pred_upper")] |>
-    group_by_at(pred) |>
-    summarise(.pred = mean(.pred), .pred_lower = mean(.pred_lower), .pred_upper = mean(.pred_upper))
-  dose_dat <- RWAR_dosing_lag_incdec[, c(pred, "Ferric.chloride..mg.L")]
-
-  if(is.numeric(dose_dat |> pull(pred))) {
-    p <- dose_dat |>
-      ggplot() +
-      geom_point(aes_string(pred, "wage"), alpha = .3) +
-      geom_line(aes_string(pred, ".pred"), data = pred_dat, colour = "darkgreen") +
-      geom_line(aes_string(pred, ".pred_lower"), data = pred_dat, colour = "darkgreen", lty = 2) +
-      geom_line(aes_string(pred, ".pred_upper"), data = pred_dat, colour = "darkgreen", lty = 2)
-  } else {
-    p <- dose_dat |>
-      ggplot() +
-      geom_jitter(aes_string(pred, "wage"), alpha = .3) +
-      geom_point(aes_string(pred, ".pred"), data = pred_dat, colour = "darkgreen") +
-      geom_errorbar(aes_string(pred, ymin = ".pred_lower", ymax = ".pred_upper"), data = pred_dat, colour = "darkgreen")
-  }
-  print(p)
-}
+ggplot(dose_pred, aes(Polymer_dose, .fitted)) +
+  geom_line() + 
+  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.2)
 ```

### 2025-12-10 Data Cleaning.Rmd
Added code to fit a model using just time of year.

@@
 ```{r}
+library(tidymodels)
+library(dplyr)
+library(lubridate)
+
+df_modeling <- df_lag_incdec |> 
+  mutate(day = yday(Timestamp))
+
+lm_model_spec <- linear_reg() |>
+  set_engine("lm") |>
+  set_mode("regression")
+
+lm_recipe <- 
+  recipe(FeCl_dose ~ day, data = df_modeling) |>
+  step_poly(day, degree = 10)
+
+lm_workflow <- workflow() |>
+  add_model(lm_model_spec) |>
+  add_recipe(lm_recipe)
+
+lm_fit_result <- fit(lm_workflow, data = df_modeling)
+
+predictions <- predict(lm_fit_result, new_data = df_modeling) |>
+  bind_cols(df_modeling |> select(FeCl_dose))
+
+rmse_result <- predictions |>
+  rmse(truth = FeCl_dose, estimate = .pred)
+
+mae_result <- predictions |>
+  mae(truth = FeCl_dose, estimate = .pred)
+
+print(rmse_result)
+print(mae_result)
+
+```
+
+```{r}
 # Look at performance of a model that simply pulls previous year dose

### 2025-12-10 Exploratory Analysis Ethan SchillingREAL.Rmd

@@
 library(Metrics)
+library(kableExtra)
 ```
@@
 
-saveRDS(TDS_plot,"TDS_plot.rds")
-saveRDS(TOC_plot,"TOC_plot.rds")
-saveRDS(DOC_plot,"DOC_plot.rds")
+#saveRDS(TDS_plot,"TDS_plot.rds")
+#saveRDS(TOC_plot,"TOC_plot.rds")
+#saveRDS(DOC_plot,"DOC_plot.rds")
 
@@
 
-getwd()
 ```
@@
 
+# All predictors
+
+model0 <- lm(FeCl_dose ~ mean_Temp + mean_pH + mean_Conductivity + mean_Turbidity + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2 + Turb_lag1 + Turb_lag2, data = train)
+
+test$pred_model0 <- predict(model0, newdata = test)
+
+cat("Model 0 RMSE:", rmse(test$FeCl_dose, test$pred_model0), "\n")
+cat("Model 0 MAE :", mae(test$FeCl_dose, test$pred_model0), "\n")
+
 # 1 Interaction Term
@@
 
+model_descriptions <- c(
+  "Main effects only",
+  "Temp × pH interaction",
+  "Temp × pH + Cond × Turb interactions",
+  "Temp × Cond interaction",
+  "Cond × pH + Temp × Cond interactions"
+)
+
+results <- data.frame(
+  Model = paste0("Model ", 0:4),
+  RMSE = c(
+    rmse(test$FeCl_dose, test$pred_model0),
+    rmse(test$FeCl_dose, test$pred_model1),
+    rmse(test$FeCl_dose, test$pred_model2),
+    rmse(test$FeCl_dose, test$pred_model3),
+    rmse(test$FeCl_dose, test$pred_model4)
+  ),
+  MAE = c(
+    mae(test$FeCl_dose, test$pred_model0),
+    mae(test$FeCl_dose, test$pred_model1),
+    mae(test$FeCl_dose, test$pred_model2),
+    mae(test$FeCl_dose, test$pred_model3),
+    mae(test$FeCl_dose, test$pred_model4)
+  ),
+  Description = model_descriptions
+)
+
+results <- results %>%
+  mutate(
+    RMSE = round(RMSE, 3),
+    MAE  = round(MAE, 3)
+  ) %>%
+  arrange(RMSE)
+
+best_rmse <- min(results$RMSE)
+best_mae  <- min(results$MAE)
+
+results %>%
+  kable("html", caption = "Model Performance Summary ") %>%
+  kable_styling(
+    bootstrap_options = c("striped", "hover", "condensed"),
+    full_width = FALSE,
+    position = "center"
+  ) %>%
+  column_spec(1, bold = TRUE) %>%
+  column_spec(4, width = "20em") %>%
+  row_spec(which(results$RMSE == best_rmse), background = "#d4edda") %>%  
+  row_spec(which(results$MAE == best_mae), background = "#cfe2f3")
+
 ```

### 2025-12-10 Presentation DSCI445.Rmd
Updates to the Presentation

@@
 author: "Ethan Schilling, Joshua Tobey, Dawson Carney, Reece Carmody"
-date: "2025-12-04"
-output: beamer_presentation
+date: "2025-12-11"
+output: 
+  beamer_presentation:
+    theme: "Warsaw"
+    colortheme: "default"
 ---
@@
 
-## Slide 1
+```{r, include=F}
+library(dplyr)
+library(tidyr)
+library(readr)
+library(lubridate)
+library(ggplot2)
+library(zoo)
+library(purrr)
+library(knitr)
+library(patchwork)
+library(tidyverse)
+library(tidymodels)
+library(rpart.plot)
+library(kableExtra)
+library(vip)
+```
+
+# Example Slide Title
+
+Example text.
+
+- Bullet points
+- Bullet points
+
+## New section of text
+
+More example text
 
 ```{r}
-readRDS("TOC_plot.rds")
-readRDS("TDS_plot.rds")
+#readRDS("TOC_plot.rds")
+#readRDS("TDS_plot.rds")
 ```
+
+# Background and Motivation
+
+The drinking water treatment process takes water from a river, lake, reservoir, or other source, and purifies it to have it reach drinking water standards.
+There are four primary steps of the water treatment process:
+
+1. Coagulation - Chemicals ("coagulants") are added to raw water to help contaminants
+group together into "floc" particles  
+2. Flocculation - The water is then slowly mixed to allow these floc particles to grow  
+3. Filtration - These particles are filtered out  
+4. Disinfection - Water is disinfected to get rid of biological contaminants
+
+This project will focus on Step 1. These chemical doses are one of the most important
+features of the treatment process that an operator can change to improve performance.
+
+# Background and Motivation
+
+**The role of an operator in deciding water treatment plant chemical doses is to
+look at a wide array of factors and make decisions about how to adjust
+chemical doses.**
+
+- Treatment plant conditions change rapidly
+- A predictive tool to help select chemical doses is beneficial both financially and environmentally
+
+```{r}
+# Add explanation picture/plot to help people visualize here??
+```
+
+# Data Overview
+
+Three years of data from 2018-2020, provided by a water treatment plant in Aurora, Colorado.
+
+- Raw Water Data
+   - pH
+   - Temperature (of the water)
+   - Turbidity
+   - Conductivity
+   - Suspended Grain Size
+   - Alkalinity
+   - Hardness
+
+- Chemical Dosing Data
+   - Ferric Chloride Dose: primary additive
+   - Cationic Polymer Dose: secondary additive
+
+# Data Cleaning
+
+- Features were averaged to match the dosing data's daily increments 
+- Merged various datasets into a final cleaned dataset
+- Various outliers (likely caused by equipment malfunction) were either removed or interpolated
+- Suspended Grain Size was not used as it was highly variable
+- Alkalinity and Hardness was not used because of missing data chunks and differing time ranges
+
+# Exploratory Analysis
+
+```{r, fig.width=7, fig.height=2}
+df_lag_incdec <- read.csv("RWAR_dosing_lag_incdec.csv") |>
+  mutate(Timestamp = ymd(Timestamp))
+
+plot_polymer <- ggplot(df_lag_incdec, aes(x = Timestamp,
+                           y = Polymer_dose,
+                           color = Polymer_dose)) +
+  geom_line() + 
+  scale_color_gradient(low = 'yellow', high = 'green') +
+  theme_minimal() +
+  ggtitle("Polymer Coagulant")
+
+print(plot_polymer)
+
+plot_FeCl <- ggplot(df_lag_incdec, aes(x = Timestamp,
+                           y = FeCl_dose,
+                           color = FeCl_dose)) +
+  geom_line() + 
+  scale_color_gradient(low = 'blue', high = 'orange') +
+  theme_minimal() +
+  ggtitle("FeCl Coagulant")
+
+print(plot_FeCl)
+```
+
+# Exploratory Analysis
+
+```{r, fig.width=7, fig.height=2}
+
+plot_pH <- ggplot(df_lag_incdec, aes(x = Timestamp,
+                           y = mean_pH,
+                           color = mean_pH)) +
+  geom_line() + 
+  scale_color_gradient(low = 'blue', high = 'red') +
+  theme_minimal() +
+  ggtitle("Average pH")
+
+print(plot_pH)
+
+plot_temp <- ggplot(df_lag_incdec, aes(x = Timestamp,
+                           y = mean_Temp,
+                           color = mean_Temp)) +
+  geom_line() + 
+  scale_color_gradient(low = 'blue', high = 'red') +
+  theme_minimal() +
+  ggtitle("Water Temperature")
+
+print(plot_temp)
+
+```
+
+# Exploratory Analysis
+
+```{r, fig.width=7, fig.height=2}
+
+plot_cond <- ggplot(df_lag_incdec, aes(x = Timestamp,
+                           y = mean_Conductivity,
+                           color = mean_Conductivity)) +
+  geom_line() + 
+  scale_color_gradient(low = 'purple', high = 'yellow') +
+  theme_minimal() +
+  ggtitle("Water Conductivity")
+
+print(plot_cond)
+
+plot_turb <- ggplot(df_lag_incdec, aes(x = Timestamp,
+                           y = mean_Turbidity,
+                           color = mean_Turbidity)) +
+  geom_line() + 
+  scale_color_gradient(low = 'black', high = 'orange') +
+  theme_minimal() +
+  ggtitle("Water Turbidity")
+
+print(plot_turb)
+
+```
+
+# Predictor Selection
+
+Generally, temperature and pH seem to have the strongest relationship. This coincides with feedback from treatment plant operators. For further analysis, we will use:
+
+- pH
+- Temperature
+- Conductivity
+- Turbidity
+
+# Introducing Lagged Variables
+
+Similar to the Stock dataset we've seen in labs and homeworks, our time series dataset needs lagged variables.
+
+- For each of the main four parameters:
+   - Introduce two lagged variables
+   - First represents the day before
+   - Second represents two days before
+
+# Coagulant Dose Prediction
+
+**Add slides for various numerical prediction models**
+
+# Coagulant Dose Prediction: Decision Tree
+
+```{r, fig.width=7, fig.height=3.8}
+set.seed(445)
+
+# Decision Tree
+
+# For now just predicting numerical amounts, not classifying
+data = read.csv("RWAR_dosing_lag_incdec.csv")
+data = as.data.frame(data)
+data$FeCl_change = NULL
+data$Polymer_change = NULL
+
+# Set up training & testing data
+
+# Can change these if you want to modify the tree/data
+tree_depth = 5 # tree depth
+min_n = 50 # minimum observations needed to further split
+
+train_data = data %>%
+  filter(year(Timestamp) <= 2019)
+test_data = data %>%
+  filter(year(Timestamp) >= 2020)
+
+# Create the tree
+
+recipe = recipe(FeCl_dose ~ ., data = train_data) %>% # Change target variable here
+  step_rm(Timestamp) %>%
+  step_rm(Polymer_dose) # Remove the other target variable here
+
+tree_spec = decision_tree(tree_depth = tree_depth, min_n = min_n) %>%
+  set_engine("rpart") %>%
+  set_mode("regression")
+
+workflow = workflow() %>%
+  add_recipe(recipe) %>%
+  add_model(tree_spec)
+
+fit_tree = workflow %>%
+  fit(data = train_data)
+
+# Plot the tree
+
+fit_tree %>%
+  extract_fit_engine() %>%
+  rpart.plot(roundint = FALSE)
+
+# Assess accuracy
+
+predictions = fit_tree %>%
+  predict(test_data) %>%
+  bind_cols(test_data)
+
+actual = predictions$`FeCl_dose` # Change target variable here
+predicted = predictions$.pred
+
+rmse = sqrt(mean((actual - predicted)^2))
+r_squared = cor(actual, predicted)^2
+mae = mean(abs(actual-predicted))
+
+metrics = data.frame(
+  Metric = c("RMSE", "R-Squared", "MAE"),
+  Value  = c(rmse, r_squared, mae)
+)
+
+kable(metrics, booktabs = TRUE, digits = 3) %>%
+  kable_styling(position = "center", font_size = 12)
+
+```
+
+# Coagulant Dose Prediction: Random Forest
+
+**Note - make a cleaner table!**
+
+```{r, fig.width=7, fig.height=3.8}
+set.seed(445)
+
+# Random Forest
+
+# For now just predicting numerical amounts, not classifying
+data = read.csv("RWAR_dosing_lag_incdec.csv")
+data = as.data.frame(data)
+data$FeCl_change = NULL
+data$Polymer_change = NULL
+
+# Set up training & testing data
+
+# Can change these if you want to modify the tree/data
+trees = 500 # number of trees in the forest
+min_n = 50 # minimum observations needed to further split
+
+train_data = data %>%
+  filter(year(Timestamp) <= 2019)
+test_data = data %>%
+  filter(year(Timestamp) >= 2020)
+
+# Create the tree
+
+recipe = recipe(FeCl_dose ~ ., data = train_data) %>% # Change target variable here
+  step_rm(Timestamp) %>%
+  step_rm(Polymer_dose) # Remove the other target variable here
+
+rf_spec = rand_forest(trees = trees, min_n = min_n) %>%
+  set_engine("ranger", importance = "impurity") %>%
+  set_mode("regression")
+
+workflow = workflow() %>%
+  add_recipe(recipe) %>%
+  add_model(rf_spec)
+
+fit_rf = workflow %>%
+  fit(data = train_data)
+
+# Assess accuracy
+
+predictions = fit_rf %>%
+  predict(test_data) %>%
+  bind_cols(test_data)
+
+actual = predictions$`FeCl_dose` # Change target variable here
+predicted = predictions$.pred
+
+# Metrics and plots
+
+model_metrics = predictions %>%
+  metrics(truth = FeCl_dose, estimate = .pred) # Change target variable here
+
+print(model_metrics)
+
+vip(fit_rf, bar = TRUE) +
+  theme_minimal(base_size = 14) +
+  labs(title = "Top Feature Importances")
+
+```
+
+# Coagulant Change Prediction
+
+**Add slides for various classification models**

### 2025-12-10 Tree Models Reece C.Rmd

@@
 library(tidymodels)
+library(rpart.plot)
+library(knitr)
+library(kableExtra)
 
@@
 min_n = 50 # minimum observations needed to further split
-train_proportion = 0.8 # split of data
 
-n = nrow(data)
-indices = sample(1:n, size = floor(train_proportion * n))
-
-train_data = data[indices, ]
-test_data  = data[-indices, ]
+train_data = data %>%
+  filter(year(Timestamp) <= 2019)
+test_data = data %>%
+  filter(year(Timestamp) >= 2020)
 
@@
 rmse = sqrt(mean((actual - predicted)^2))
-print(paste("RMSE: ", rmse))
+r_squared = cor(actual, predicted)^2
+mae = mean(abs(actual-predicted))
 
-r_squared = cor(actual, predicted)^2
-print(paste("R Squared: ", r_squared))
+metrics = data.frame(
+  Metric = c("RMSE", "R-Squared", "MAE"),
+  Value  = c(rmse, r_squared, mae)
+)
+
+kable(metrics, booktabs = TRUE, digits = 3) %>%
+  kable_styling(position = "center", font_size = 12)
 
 # Plot the tree
-
-library(rpart.plot)
 
@@
 library(tidymodels)
+#library(yardstick)
+library(vip)
 
@@
 rf_spec = rand_forest(trees = trees, min_n = min_n) %>%
-  set_engine("ranger") %>%
+  set_engine("ranger", importance = "impurity") %>%
   set_mode("regression")
@@
 
-rmse = sqrt(mean((actual - predicted)^2))
-print(paste("RMSE: ", rmse))
+# Metrics and plots
 
-r_squared = cor(actual, predicted)^2
-print(paste("R Squared: ", r_squared))
+model_metrics = predictions %>%
+  metrics(truth = FeCl_dose, estimate = .pred) # Change target variable here
+
+print(model_metrics)
+
+ggplot(predictions, aes(x = FeCl_dose, y = .pred)) +
+  geom_point() +
+  geom_abline(intercept = 0, slope = 1) +
+  labs(title = "Random Forest Accuracy",
+       x = "Actual",
+       y = "Predicted") +
+  theme_bw()
+
+vip(fit_rf)
 

### 2025-12-10 Data Cleaning.Rmd
Edited table display for confusion matrices.

@@
 library(kableExtra)
+library(tibble)
 
@@
 print(accuracy_result)
-print(confusion_matrix)
+print(confusion_matrix) |> kable()
 

### 2025-12-10 Exploratory Analysis Ethan SchillingREAL.Rmd
Added table of results for linear models predicting on FeCl_dose

--- 2025-12-09 Exploratory Analysis Ethan SchillingREAL.Rmd
@@
 library(Metrics)
+library(kableExtra)
+library(gt)
 
@@
 
+# All predictors
+
+model0 <- lm(FeCl_dose ~ mean_Temp + mean_pH + mean_Conductivity + mean_Turbidity + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2 + Turb_lag1 + Turb_lag2, data = train)
+
+test$pred_model0 <- predict(model0, newdata = test)
+
 # 1 Interaction Term
-model1 <- lm(FeCl_dose ~ mean_Temp * mean_Turbidity + mean_pH + Temp_lag1 + Turb_lag1, 
-             data = train)
+
+model1 = lm(FeCl_dose ~ mean_Temp * mean_pH + mean_Turbidity + Temp_lag1  + Temp_lag2 + Turb_lag1 +     Turb_lag2,data = train)
 summary(model1)
@@
 
-rmse(test$FeCl_dose, test$pred_model1)
-mae(test$FeCl_dose, test$pred_model1)
+# 2 Interaction Terms
 
-# 2 Interaction Terms
-model2 <- lm(FeCl_dose ~ mean_Temp * mean_pH +mean_Conductivity * mean_Turbidity + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2, data = train)
+model2 = lm(FeCl_dose ~ mean_Temp * mean_pH + mean_Conductivity * mean_Turbidity +
+                Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2, data = train)
 
@@
 
-rmse(test$FeCl_dose, test$pred_model2)
-mae(test$FeCl_dose, test$pred_model2)
+model3 = lm(FeCl_dose ~ mean_Temp * mean_Conductivity + Temp_lag1 + Temp_lag2 + Cond_lag1 + Cond_lag2,                 data = train)
 
-# Step Up
+test$pred_model3 <- predict(model3, newdata = test)
 
-response_var <- "FeCl_dose" 
-predictors <- c("mean_Temp", "mean_pH", "mean_Conductivity",
-                "mean_Turbidity", "Temp_lag1", "Temp_lag2",
-                "pH_lag1", "pH_lag2", "Cond_lag1", "Cond_lag2",
-                "Turb_lag1", "Turb_lag2")
+# 2 Interactions: Cond and PH and Temp and Cond
+
+model4 = lm(FeCl_dose ~ mean_Conductivity * mean_pH + mean_Temp * mean_Conductivity +
+                Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2,
+              data = train)
+
+test$pred_model4 <- predict(model4, newdata = test)
+
+model5 = lm(FeCl_dose ~ mean_Temp + mean_pH + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2,
+              data = train)
+
+test$pred_model5 <- predict(model5, newdata = test)
+
+models <- list(
+  model0 = lm(FeCl_dose ~ mean_Temp + mean_pH + mean_Conductivity + mean_Turbidity +
+                Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2 +
+                Turb_lag1 + Turb_lag2, data = train),
+
+  model1 = lm(FeCl_dose ~ mean_Temp * mean_pH + mean_Turbidity + Temp_lag1  + Temp_lag2 + Turb_lag1 +
+                Turb_lag2,data = train),
+
+  model2 = lm(FeCl_dose ~ mean_Temp * mean_pH + mean_Conductivity * mean_Turbidity +
+                Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2, data = train),
+
+  model3 = lm(FeCl_dose ~ mean_Temp * mean_Conductivity + Temp_lag1 + Temp_lag2 + Cond_lag1 + Cond_lag2,                 data = train),
+
+  model4 = lm(FeCl_dose ~ mean_Conductivity * mean_pH + mean_Temp * mean_Conductivity +
+                Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2,
+              data = train),
+
+  model5 = lm(FeCl_dose ~ mean_Temp + mean_pH + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2,
+              data = train)
+)
+
+descriptions <- c(
+  "Main effects only",
+  "Temp × pH interaction",
+  "Temp × pH + Cond × Turb interactions",
+  "Temp × Cond interaction",
+  "Cond × pH + Temp × Cond interactions",
+  "Only Temp and pH"
+)
+
+performance <- data.frame(Model = character(),
+                          RMSE = numeric(),
+                          MAE = numeric(),
+                          Description = character(),
+                          stringsAsFactors = FALSE)
+
+for(i in 1:length(models)) {
+  pred <- predict(models[[i]], newdata = test)
+  performance <- rbind(performance,
+                       data.frame(Model = paste0("Model ", i-1),
+                                  RMSE = round(rmse(test$FeCl_dose, pred), 3),
+                                  MAE  = round(mae(test$FeCl_dose, pred), 3),
+                                  Description = descriptions[i],
+                                  stringsAsFactors = FALSE))
+
+}
+  performance <- performance %>% arrange(RMSE)
+
+best_rmse <- min(performance$RMSE)
+
+kable_table <- performance %>%
+  kable("html", caption = "Model Performance Summary") %>%
+  kable_styling(full_width = F, position = "center") %>%
+  row_spec(which(performance$RMSE == best_rmse), background = "#cfe2f3")
+
+kable_table
 

### 2025-12-10 Presentation DSCI445.Rmd
Made lots of changes to presentation.

@@
 ---
-title: "DSCI 445 Presentation"
+title: "Predicting Water Treatment Plant Chemical Doses"
 author: "Ethan Schilling, Joshua Tobey, Dawson Carney, Reece Carmody"
-date: "2025-12-11"
+date: "11 December 2025"
+aspectratio: 169
 output: 
@@
 library(vip)
-```
-
-# Example Slide Title
-
-Example text.
-
-- Bullet points
-- Bullet points
-
-## New section of text
-
-More example text
-
-```{r}
-#readRDS("TOC_plot.rds")
-#readRDS("TDS_plot.rds")
+library(png)
+library(grid)
 ```
@@
 The drinking water treatment process takes water from a river, lake, reservoir, or other source, and purifies it to have it reach drinking water standards.
-There are four primary steps of the water treatment process:
 
-1. Coagulation - Chemicals ("coagulants") are added to raw water to help contaminants
-group together into "floc" particles  
-2. Flocculation - The water is then slowly mixed to allow these floc particles to grow  
-3. Filtration - These particles are filtered out  
-4. Disinfection - Water is disinfected to get rid of biological contaminants
-
-This project will focus on Step 1. These chemical doses are one of the most important
-features of the treatment process that an operator can change to improve performance.
+```{r}
+img_array = readPNG("Images/treatmentplot1.png")
+grid.raster(img_array)
+```
 
@@
 
-**The role of an operator in deciding water treatment plant chemical doses is to
-look at a wide array of factors and make decisions about how to adjust
+**One role of an operator in the water treatment process is to
+look at a wide array of data and make decisions about how to adjust
 chemical doses.**
 
-- Treatment plant conditions change rapidly
-- A predictive tool to help select chemical doses is beneficial both financially and environmentally
+```{r}
+img_array = readPNG("Images/treatmentplot2.png")
+grid.raster(img_array)
+```
+
+# Background and Motivation
+
+**OUR GOAL: Create a model that serves the role of an "operator," taking raw water data and making predictions of chemical doses.**
 
 ```{r}
-# Add explanation picture/plot to help people visualize here??
+img_array = readPNG("Images/treatmentplot3.png")
+grid.raster(img_array)
 ```
@@
 
-Three years of data from 2018-2020, provided by a water treatment plant in Aurora, Colorado.
+Three years of data from 2018-2020, provided by a Colorado water treatment plant.
 
@@
 
-- Features were averaged to match the dosing data's daily increments 
+- Daily averages were taken for raw water data to match the dosing data's daily timestep 
 - Merged various datasets into a final cleaned dataset
-- Various outliers (likely caused by equipment malfunction) were either removed or interpolated
-- Suspended Grain Size was not used as it was highly variable
-- Alkalinity and Hardness was not used because of missing data chunks and differing time ranges
+- Various large outliers (likely caused by equipment malfunction) were removed
+- Temperature outliers were interpolated to nearest adjacent values
+- Suspended grain size was not used as it was highly variable
+- Alkalinity and hardness were not used because of missing data chunks and differing time ranges
 
@@
 
-Generally, temperature and pH seem to have the strongest relationship. This coincides with feedback from treatment plant operators. For further analysis, we will use:
+Temperature and pH seem to have the strongest relationship to dose, consistent with feedback from treatment plant operators.
+Other data were either highly variable or had minimum availability with large gaps.
+We will focus on the following to predict dose:
 
@@
 - Temperature
-- Conductivity
-- Turbidity
+- Conductivity (lots of noise, weaker)
+- Turbidity (lots of noise, weaker)
 
-# Introducing Lagged Variables
+Focus: **FeCl Dose**. Polymer dose is not changed significantly in plants.
 
-Similar to the Stock dataset we've seen in labs and homeworks, our time series dataset needs lagged variables.
+# Added Variables
+
+Similar to the stock dataset we've seen in labs and homeworks, our time series dataset needs lagged variables.
 
@@
 
+We also introduced a categorical predictor indicating whether the dose increased, decreased, or stayed the same compared to the day before.
+
+# Summary of Models
+
+**GOAL: Interpretable and Captures Nonlinear Behavior**
+
+- Training and Testing
+   - Training: 2018-2019
+   - Testing: 2020
+   - Approximately 70/30 split
+- Coagulant Dose Prediction
+   - Linear Regression
+   - Tree-based Models
+   - GAM
+- Coagulant Dose Change Prediction
+   - Multinomial Classification
+   - LDA, QDA
+   - Random Forest
+
 # Coagulant Dose Prediction
 
-**Add slides for various numerical prediction models**
+**GOAL: Predict coagulant doses based on raw water characteristics.**
+
+```{r, fig.width=7, fig.height=3}
+
+df_lag_incdec_long <- df_lag_incdec |>
+  pivot_longer(
+    cols = starts_with("mean_"),
+    names_to = "Characteristics",
+    values_to = "Value"
+  )
+
+ggplot(df_lag_incdec_long, aes(x = Timestamp, y = Value, color = FeCl_dose)) +
+  geom_point(size = 0.7) +
+  facet_wrap(~Characteristics, scales = "free_y") +
+  scale_color_viridis_c(option = "plasma") +
+  theme_minimal() +
+  labs(
+    x = "Date",
+    y = "Value",
+    color = "Ferric Chloride (mg/L)",
+    title = "Raw Water Characteristics Over Time Colored by Dosing"
+  )
+
+```
+
+# Coagulant Dose Prediction: Linear Regression
+
+**Ethan: insert table with models attempted**
 
@@
 
-```{r, fig.width=7, fig.height=3.8}
+```{r, fig.width=6.5, fig.height=3.4}
 set.seed(445)
@@
 
-# Coagulant Change Prediction
+# Coagulant Dose Change Prediction
 
-**Add slides for various classification models**
+**GOAL: Predict whether coagulant dose increased, decreased, or stayed the same based on raw water characteristics.**
+
+```{r, fig.width=7, fig.height=3}
+
+ggplot(df_lag_incdec_long, aes(x = Timestamp, y = Value, color = FeCl_change)) +
+  geom_point(size = 0.7) +
+  facet_wrap(~Characteristics, scales = "free_y") +
+  theme_minimal() +
+  labs(
+    x = "Date",
+    y = "Value",
+    color = "FeCl_change",
+    shape = "FeCl_change",
+    title = "Raw Water Characteristics Colored by Dose Change"
+  )
+
+```
+
+# Coagulant Dose Change Prediction: Modeling
+
+#
+
+# Summary of Dose Prediction Models
+
+**Insert bar graph/brief conclusions**
+
+# Summary of Dose Change Prediction Models
+
+**Insert bar graph/brief conclusions**

### 2025-12-10 Presentation DSCI445.Rmd
Fixed tree slides

@@
 
-```{r, fig.width=7, fig.height=3.8}
+```{r, fig.width=5, fig.height=2.5}
 set.seed(445)
@@
 
-**Note - make a cleaner table!**
-
-```{r, fig.width=6.5, fig.height=3.4}
+```{r, fig.width=5, fig.height=2.1}
 set.seed(445)
@@
 
+# Metrics and plots
+
 model_metrics = predictions %>%
   metrics(truth = FeCl_dose, estimate = .pred) # Change target variable here
+rmse = model_metrics$.estimate[1]
+r_squared = model_metrics$.estimate[2]
+mae = model_metrics$.estimate[3]
 
-print(model_metrics)
+metrics = data.frame(
+  Metric = c("RMSE", "R-Squared", "MAE"),
+  Value  = c(rmse, r_squared, mae)
+)
 
-vip(fit_rf, bar = TRUE) +
-  theme_minimal(base_size = 14) +
-  labs(title = "Top Feature Importances")
+vip(fit_rf, num_features = 20) +
+  geom_col(color = "black", fill = "coral") +
+  labs(
+    title = "Feature Importance (Random Forest)",
+    x = "Importance",
+    y = "Feature"
+  ) +
+  theme_minimal(base_size = 9)
+
+kable(metrics, booktabs = TRUE, digits = 3) %>%
+  kable_styling(position = "center", font_size = 12)
 

### 2025-12-10 Exploratory Analysis Ethan SchillingREAL.Rmd

@@
 library(kableExtra)
+
 ```
@@
                        data.frame(Model = paste0("Model ", i-1),
-                                  RMSE = round(rmse(test$FeCl_dose, pred), 3),
-                                  MAE  = round(mae(test$FeCl_dose, pred), 3),
+                                  RMSE = round(Metrics::rmse(test$FeCl_dose, pred), 3),
+                                  MAE  = round(Metrics::mae(test$FeCl_dose, pred), 3),
                                   Description = descriptions[i],
@@
 
+p3 <- ggplot(test, aes(x = FeCl_dose, y = pred_model3)) +
+  geom_point(color = "blue") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
+  labs(title = "Model 3: Predicted vs Actual FeCl_dose",
+       x = "Actual FeCl_dose",
+       y = "Predicted FeCl_dose") +
+  theme_minimal()
+
+p4 <- ggplot(test, aes(x = FeCl_dose, y = pred_model4)) +
+  geom_point(color = "green") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
+  labs(title = "Model 4: Predicted vs Actual FeCl_dose",
+       x = "Actual FeCl_dose",
+       y = "Predicted FeCl_dose") +
+  theme_minimal()
+
+p3
+p4
+
 ```
+
+# Coagulant Dose Prediction Linear Regression
+
+```{r}
+knitr::include_graphics("Images/linear_model_table.png")
+```
+
+## Regression Models
+
+- Created 6 models with different predictors and combinations of interaction terms.
+
+- Dropped Turbidity as it was not a good predictor for FeCL dosing.
+
+- Difficult to determine what factors of the dataset are driving the predictions.
+
+### Performance
+
+- Performed best with multiple interactions between Conductivity, Temp, and pH
+
+- Still not great predictions many are overfit and underfit.
+
+- Linear Structure doesn't capture the relations in our predictors

### 2025-12-10 Presentation DSCI445.Rmd

@@
 
-**Ethan: insert table with models attempted**
+- Created 6 models with different predictors and combinations of interaction terms.
+
+- Used lag variables for temp, ph, conductivity, turbidity
+
+- Dropped Turbidity as it was not a good predictor for FeCL dosing.
+  - Conductivity was also dropped eventually due to being similar to Turbidity
+
+- Difficult to determine what factors of the dataset are driving the predictions.
+
+- RMSE and MAE seem to be all over the place dependent on which predictors are selected and the interactions.
+
+# RMSE & MAE Table
+
+```{r, out.width="80%", out.height="88%"}
+knitr::include_graphics("Images/linear_model_table.png")
+```
+
+# Performance
+
+- Performed best with multiple interactions between Conductivity, Temp, and pH
+
+- Still not great predictions, many are overfit and underfit.
+
+- Linear Structure doesn't capture the relations in our predictors, and incorrect predictions could lead to disastorus results if too much or too little chemical is used.
 

### 2025-12-10 Presentation DSCI445.Rmd
Edited data cleaning slide, other.

@@
 
-- Daily averages were taken for raw water data to match the dosing data's daily timestep 
-- Merged various datasets into a final cleaned dataset
-- Various large outliers (likely caused by equipment malfunction) were removed
-- Temperature outliers were interpolated to nearest adjacent values
+- Dropped Data  
    - Suspended grain size was not used as it was highly variable
    - Alkalinity and hardness were not used because of missing data chunks and differing time ranges
+
+- Processing
+   - Daily averages were taken for raw water data to match the dosing data's timestep
+   - Significant outliers (likely caused by equipment malfunction) were removed or interpolated
+   - Dosing and raw water data merged
 
@@
 
-```{r, fig.width=7, fig.height=2}
+```{r, fig.width=7, fig.height=4}
 
-plot_pH <- ggplot(df_lag_incdec, aes(x = Timestamp,
-                           y = mean_pH,
-                           color = mean_pH)) +
-  geom_line() + 
-  scale_color_gradient(low = 'blue', high = 'red') +
+df_lag_incdec_long <- df_lag_incdec |>
+  pivot_longer(
+    cols = starts_with("mean_"),
+    names_to = "Characteristics",
+    values_to = "Value"
+  )
+
+ggplot(df_lag_incdec_long, aes(x = Timestamp, y = Value)) +
+  geom_point(size = 0.5, color="blue") +
+  facet_wrap(~Characteristics, scales = "free_y") +
   theme_minimal() +
-  ggtitle("Average pH")
-
-print(plot_pH)
-
-plot_temp <- ggplot(df_lag_incdec, aes(x = Timestamp,
-                           y = mean_Temp,
-                           color = mean_Temp)) +
-  geom_line() + 
-  scale_color_gradient(low = 'blue', high = 'red') +
-  theme_minimal() +
-  ggtitle("Water Temperature")
-
-print(plot_temp)
-
-```
-
-# Exploratory Analysis
-
-```{r, fig.width=7, fig.height=2}
-
-plot_cond <- ggplot(df_lag_incdec, aes(x = Timestamp,
-                           y = mean_Conductivity,
-                           color = mean_Conductivity)) +
-  geom_line() + 
-  scale_color_gradient(low = 'purple', high = 'yellow') +
-  theme_minimal() +
-  ggtitle("Water Conductivity")
-
-print(plot_cond)
-
-plot_turb <- ggplot(df_lag_incdec, aes(x = Timestamp,
-                           y = mean_Turbidity,
-                           color = mean_Turbidity)) +
-  geom_line() + 
-  scale_color_gradient(low = 'black', high = 'orange') +
-  theme_minimal() +
-  ggtitle("Water Turbidity")
-
-print(plot_turb)
+  labs(
+    x = "Date",
+    y = "Value",
+    title = "Raw Water Characteristics"
+  )
 

### 2025-12-10 Exploratory_Analysis_Josh_Tobey.Rmd
Updated file with gam and plot against fitted and actuals.

--- 2025-12-08 Exploratory_Analysis_Josh_Tobey.Rmd
@@
 
-```{r}
-time_lag_1 <- 
-time_lag_2 <- 
-time_lag_3 <-
-
+```{r, eval = FALSE}
 weekly_log_spec <- logistic_reg()
@@
 
-```{r}
+```{r, eval = FALSE}
 cross_validation <- initial_split(exploratory_df, prop = 0.6)
@@
 RWAR_dosing_lag_incdec <- RWAR_dosing_lag_incdec %>%
-  select(-c(FeCl_change, Polymer_change, Timestamp))
+  select(-c(FeCl_change, Polymer_change))
 head(RWAR_dosing_lag_incdec)
@@
 library(mgcv)
+library(parsnip)
+df_full <- RWAR_dosing_lag_incdec
+df_train <- df_full |> filter(year(Timestamp) <= 2019)
+df_test  <- df_full |> filter(year(Timestamp) == 2020)
 
-lm_spec <- linear_reg()
-
-gam_formula <- Ferric.chloride..mg.L ~ s(Cationic.Polimer..mg.L) + s(mean_Temp) + s(mean_pH) + s(mean_Conductivity) + s(mean_Turbidity) + s(Temp_lag1) + s(Temp_lag2) + s(pH_lag1) + s(pH_lag2) + s(Cond_lag1) + s(Cond_lag2) + s(Turb_lag1) + s(Turb_lag2)
-
-gam_fit <- gam(gam_formula, data = RWAR_dosing_lag_incdec)
-gam_rec <- recipe(Ferric.chloride..mg.L ~ ., data = RWAR_dosing_lag_incdec) %>%
-  step_ns(Cationic.Polimer..mg.L, deg_free = 10) %>%
-  step_ns(mean_Temp, deg_free = 10) %>%
-  step_ns(mean_pH, deg_free = 10) %>%
-  step_ns(mean_Conductivity, deg_free = 10) %>%
-  step_ns(mean_Turbidity, deg_free = 10) %>%
-  step_ns(Temp_lag1, deg_free = 10) %>%
-  step_ns(Temp_lag2, deg_free = 10) %>%
-  step_ns(pH_lag1, deg_free = 10) %>%
-  step_ns(pH_lag2, deg_free = 10) %>%
-  step_ns(Cond_lag1, deg_free = 10) %>%
-  step_ns(Cond_lag2, deg_free = 10) %>%
-  step_ns(Turb_lag1, deg_free = 10) %>%
-  step_ns(Turb_lag2, deg_free = 10)
+gam_spec <- gen_additive_mod() |> 
+  set_engine("mgcv") |> 
+  set_mode("regression")
+gam_rec <- recipe(FeCl_dose ~ Polymer_dose + mean_Temp + mean_pH + mean_Conductivity +
+                 mean_Turbidity + Temp_lag1 + Temp_lag2 +
+                 pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2 +
+                 Turb_lag1 + Turb_lag2, data = df_train)
 
 gam_wf <- workflow() |>
-  add_model(lm_spec) |>
+  add_model(gam_spec) |>
   add_recipe(gam_rec)
 
-gam_fit <- gam_wf |>
-  fit(data = RWAR_dosing_lag_incdec)
+gam_fit <- gam(
+  FeCl_dose ~ s(Polymer_dose) + s(mean_Temp) + s(mean_pH) + s(mean_Conductivity) +
+               s(mean_Turbidity) + s(Temp_lag1) + s(Temp_lag2) +
+               s(pH_lag1) + s(pH_lag2) + s(Cond_lag1) + s(Cond_lag2) +
+               s(Turb_lag1) + s(Turb_lag2),
+  data = df_train
+)
 
@@
 gam_fit |> glance()
-
-dat_grid <- tibble(expand.grid(Cationic.Polimer..mg.L = seq(min(RWAR_dosing_lag_incdec$Cationic.Polimer..mg.L), 
-                               max(RWAR_dosing_lag_incdec$Cationic.Polimer..mg.L)),
-                               mean_Temp = seq(min(RWAR_dosing_lag_incdec$mean_Temp), max(RWAR_dosing_lag_incdec$mean_Temp)),
-                               mean_pH = seq(min(RWAR_dosing_lag_incdec$mean_pH), max(RWAR_dosing_lag_incdec$mean_pH)),
-                               mean_Conductivity = seq(min(RWAR_dosing_lag_incdec$mean_Conductivity), max(RWAR_dosing_lag_incdec$mean_Conductivity)),
-                               mean_Turbidity = seq(min(RWAR_dosing_lag_incdec$mean_Turbidity), max(RWAR_dosing_lag_incdec$mean_Turbidity)),
-                               Temp_lag1 = seq(min(RWAR_dosing_lag_incdec$Temp_lag1), max(RWAR_dosing_lag_incdec$Temp_lag1)),
-                               Temp_lag2 = seq(min(RWAR_dosing_lag_incdec$Temp_lag2), max(RWAR_dosing_lag_incdec$Temp_lag2)),
-                               pH_lag1 = seq(min(RWAR_dosing_lag_incdec$pH_lag1), max(RWAR_dosing_lag_incdec$pH_lag1)),
-                               pH_lag2 = seq(min(RWAR_dosing_lag_incdec$pH_lag2), max(RWAR_dosing_lag_incdec$pH_lag2)),
-                               Cond_lag1 = seq(min(RWAR_dosing_lag_incdec$Cond_lag1), max(RWAR_dosing_lag_incdec$Cond_lag1)),
-                               Cond_lag2 = seq(min(RWAR_dosing_lag_incdec$Cond_lag2), max(RWAR_dosing_lag_incdec$Cond_lag2)),
-                               Turb_lag1 = seq(min(RWAR_dosing_lag_incdec$Turb_lag1), max(RWAR_dosing_lag_incdec$Turb_lag1)),
-                               Turb_lag2 = seq(min(RWAR_dosing_lag_incdec$Turb_lag2), max(RWAR_dosing_lag_incdec$Turb_lag2))
-                               ))
-bind_cols(
-  augment(gam_fit, new_data = dat_grid),
-  predict(gam_fit, new_data = dat_grid, type = "conf_int")
-) -> dose_pred_gam
-
-for(pred in c("age", "year", "education")) {
-  pred_dat <- dose_pred_gam[, c(pred, ".pred", ".pred_lower", ".pred_upper")] |>
-    group_by_at(pred) |>
-    summarise(.pred = mean(.pred), .pred_lower = mean(.pred_lower), .pred_upper = mean(.pred_upper))
-  dose_dat <- RWAR_dosing_lag_incdec[, c(pred, "Ferric.chloride..mg.L")]
-
-  if(is.numeric(dose_dat |> pull(pred))) {
-    p <- dose_dat |>
-      ggplot() +
-      geom_point(aes_string(pred, "wage"), alpha = .3) +
-      geom_line(aes_string(pred, ".pred"), data = pred_dat, colour = "darkgreen") +
-      geom_line(aes_string(pred, ".pred_lower"), data = pred_dat, colour = "darkgreen", lty = 2) +
-      geom_line(aes_string(pred, ".pred_upper"), data = pred_dat, colour = "darkgreen", lty = 2)
-  } else {
-    p <- dose_dat |>
-      ggplot() +
-      geom_jitter(aes_string(pred, "wage"), alpha = .3) +
-      geom_point(aes_string(pred, ".pred"), data = pred_dat, colour = "darkgreen") +
-      geom_errorbar(aes_string(pred, ymin = ".pred_lower", ymax = ".pred_upper"), data = pred_dat, colour = "darkgreen")
-  }
-  print(p)
-}
+```
+```{r}
+pred_df <- df_test %>% 
+  mutate(pred = predict(gam_fit, newdata = df_test))
+rmse(pred_df, truth = FeCl_dose, estimate = pred)
+#The test data shows that
 
 ```
+
+```{r}
+
+dat_grid_dose <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Polymer_dose = seq(min(df_train$Polymer_dose),
+                                      max(df_train$Polymer_dose),
+                                      length.out = 100))
+dat_grid_temp <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(mean_Temp = seq(min(df_train$mean_Temp), 
+                         max(df_train$mean_Temp),
+                                      length.out = 100))
+dat_grid_pH <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(mean_pH = seq(min(df_train$mean_pH), max(df_train$mean_pH),
+                                      length.out = 100))
+dat_grid_cond <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(mean_Conductivity = seq(min(df_train$mean_Conductivity), max(df_train$mean_Conductivity),
+                                      length.out = 100))
+dat_grid_turb <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(mean_Turbidity = seq(min(df_train$mean_Turbidity), max(df_train$mean_Turbidity),
+                                      length.out = 100))
+dat_grid_temp_lag1 <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Temp_lag1 = seq(min(df_train$Temp_lag1), max(df_train$Temp_lag1),
+                                      length.out = 100))
+dat_grid_temp_lag2 <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Temp_lag2 = seq(min(df_train$Temp_lag2), max(df_train$Temp_lag2),
+                                      length.out = 100))
+dat_grid_pH_lag1 <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(pH_lag1 = seq(min(df_train$pH_lag1), max(df_train$pH_lag1),
+                                      length.out = 100))
+dat_grid_pH_lag2 <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(pH_lag2 = seq(min(df_train$pH_lag2), max(df_train$pH_lag2),
+                                      length.out = 100))
+dat_grid_cond_lag1 <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Cond_lag1 = seq(min(df_train$Cond_lag1), max(df_train$Cond_lag1),
+                                      length.out = 100))
+dat_grid_cond_lag2 <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Cond_lag2 = seq(min(df_train$Cond_lag2), max(df_train$Cond_lag2),
+                                      length.out = 100))
+dat_grid_turb_lag1 <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Turb_lag1 = seq(min(df_train$Turb_lag1), max(df_train$Turb_lag1),
+                                      length.out = 100))
+dat_grid_turb_lag2 <- df_train |> 
+  summarize(
+    across(everything(), mean, na.rm = TRUE)
+  ) |> 
+  slice(rep(1, 100)) |> 
+  mutate(Turb_lag2 = seq(min(df_train$Turb_lag2), max(df_train$Turb_lag2),
+                                      length.out = 100))
+
+pred1 <- predict(gam_fit, newdata = dat_grid_dose, type = "response", se.fit = TRUE)
+dose_pred <- dat_grid_dose |> mutate(
+  .fitted = pred1$fit,
+  .lower = pred1$fit - 2 * pred1$se.fit,
+  .upper = pred1$fit + 2 * pred1$se.fit
+  )
+pred2 <- predict(gam_fit, newdata = dat_grid_temp, type = "response", se.fit = TRUE)
+temp_pred <- dat_grid_temp |> mutate(
+  .fitted = pred2$fit,
+  .lower = pred2$fit - 2 * pred2$se.fit,
+  .upper = pred2$fit + 2 * pred2$se.fit
+)
+pred3 <- predict(gam_fit, newdata = dat_grid_pH, type = "response", se.fit = TRUE)
+pH_pred <- dat_grid_pH |> mutate(
+  .fitted = pred3$fit,
+  .lower = pred3$fit - 2 * pred3$se.fit,
+  .upper = pred3$fit + 2 * pred3$se.fit
+)
+pred4 <- predict(gam_fit, newdata = dat_grid_cond, type = "response", se.fit = TRUE)
+cond_pred <- dat_grid_cond |> mutate(
+  .fitted = pred4$fit,
+  .lower = pred4$fit - 2 * pred4$se.fit,
+  .upper = pred4$fit + 2 * pred4$se.fit
+)
+pred5 <- predict(gam_fit, newdata = dat_grid_turb, type = "response", se.fit = TRUE)
+turb_pred <- dat_grid_turb |> mutate(
+  .fitted = pred5$fit,
+  .lower = pred5$fit - 2 * pred5$se.fit,
+  .upper = pred5$fit + 2 * pred5$se.fit
+)
+pred6 <- predict(gam_fit, newdata = dat_grid_temp_lag1, type = "response", se.fit = TRUE)
+tempLag1_pred <- dat_grid_temp_lag1 |> mutate(
+  .fitted = pred6$fit,
+  .lower = pred6$fit - 2 * pred6$se.fit,
+  .upper = pred6$fit + 2 * pred6$se.fit
+)
+pred7 <- predict(gam_fit, newdata = dat_grid_temp_lag2, type = "response", se.fit = TRUE)
+tempLag2_pred <- dat_grid_temp_lag2 |> mutate(
+  .fitted = pred7$fit,
+  .lower = pred7$fit - 2 * pred7$se.fit,
+  .upper = pred7$fit + 2 * pred7$se.fit
+)
+pred8 <- predict(gam_fit, newdata = dat_grid_pH_lag1, type = "response", se.fit = TRUE)
+pHLag1_pred <- dat_grid_pH_lag1 |> mutate(
+  .fitted = pred8$fit,
+  .lower = pred8$fit - 2 * pred8$se.fit,
+  .upper = pred8$fit + 2 * pred8$se.fit
+)
+pred9 <- predict(gam_fit, newdata = dat_grid_pH_lag2, type = "response", se.fit = TRUE)
+pHLag2_pred <- dat_grid_pH_lag2 |> mutate(
+  .fitted = pred9$fit,
+  .lower = pred9$fit - 2 * pred9$se.fit,
+  .upper = pred9$fit + 2 * pred9$se.fit
+)
+pred10 <- predict(gam_fit, newdata = dat_grid_cond_lag1, type = "response", se.fit = TRUE)
+condLag1_pred <- dat_grid_cond_lag1 |> mutate(
+  .fitted = pred10$fit,
+  .lower = pred10$fit - 2 * pred10$se.fit,
+  .upper = pred10$fit + 2 * pred10$se.fit
+)
+pred11 <- predict(gam_fit, newdata = dat_grid_cond_lag2, type = "response", se.fit = TRUE)
+condLag2_pred <- dat_grid_cond_lag2 |> mutate(
+  .fitted = pred11$fit,
+  .lower = pred11$fit - 2 * pred11$se.fit,
+  .upper = pred11$fit + 2 * pred11$se.fit
+)
+pred12 <- predict(gam_fit, newdata = dat_grid_turb_lag1, type = "response", se.fit = TRUE)
+turbLag1_pred <- dat_grid_turb_lag1 |> mutate(
+  .fitted = pred12$fit,
+  .lower = pred12$fit - 2 * pred12$se.fit,
+  .upper = pred12$fit + 2 * pred12$se.fit
+)
+pred13 <- predict(gam_fit, newdata = dat_grid_turb_lag2, type = "response", se.fit = TRUE)
+turbLag2_pred <- dat_grid_turb_lag2 |> mutate(
+  .fitted = pred13$fit,
+  .lower = pred13$fit - 2 * pred13$se.fit,
+  .upper = pred13$fit + 2 * pred13$se.fit
+)
+
+```
+
+```{r}
+summary(gam_fit)
+```
+
+```{r}
+ggplot(pred_df) +
+  geom_point(aes(x = FeCl_dose, y = pred), alpha = 0.3) +
+  geom_abline(slope = 1, 
+              intercept = coef(gam_fit)[1],
+              color = 'red',
+              lwd = 1)
+
+```
+```{r}
+ggplot(df_test) +
+  geom_line(aes(y = FeCl_dose), color = 'orange')
+
+```
+
+The GAM model shows poor performance when making prediction of the dose requirements for water treatment.

### 2025-12-10 Data Cleaning.Rmd

@@
 ## RANDOM FOREST
-rf_spec <- rand_forest(mtry = sqrt(.cols())) |>
+rf_spec <- rand_forest(mtry = sqrt(.cols()), trees=500, min_n=1) |>
   set_engine("randomForest", importance = TRUE) |>
@@
 # 34% accuracy for all predictors
-# 47% accuracy for just temperature and pH
+# 46.6% accuracy for just temperature and pH
 

### 2025-12-10 Presentation DSCI445.Rmd

@@
 
-Similar to the stock dataset we've seen in labs and homeworks, our time series dataset needs lagged variables.
+Similar to the stock dataset we've seen in labs and homeworks, our time series dataset needs **lagged variables**.
 
@@
 
-We also introduced a categorical predictor indicating whether the dose increased, decreased, or stayed the same compared to the day before.
+We also introduced a **categorical predictor** indicating whether the dose increased, decreased, or stayed the same compared to the day before.
 
@@
 
-**GOAL: Interpretable and Captures Nonlinear Behavior**
+**Priorities: Interpretability, Capturing of Nonlinear Behavior**
 
@@
 
-# Coagulant Dose Change Prediction: Modeling
+# Coagulant Dose Change Prediction: Multinomial, LDA/QDA
 
-#
+# Coagulant Dose Change Prediction: Random Forest
 
 # Summary of Dose Prediction Models
+
+__ performed the best out of models tested.
 
@@
 
-**Insert bar graph/brief conclusions**
+Overall ineffective, and prone to overfitting (especially random forest).  
+Model predicting no dose change would be more accurate.
+
+**Insert bar graph**
+
+# Conclusion and Possible Next Steps
+
+- Predicting chemical doses based on these raw water characteristics proved to be ineffective
+- Most likely reasons:
+   - Limited size of dataset
+   - Limited predictor set
+   - Operator-dependent
+   - Complexity of chemical relationships
+   - Other factors (example: reservoir turnover)
+- Operators make decisions based on variety of other factors: chemical waste production, filter performance, lab tests, etc.
+- Incorporate more varieties and time ranges of data in future modeling efforts

### 2025-12-10 Tree Models Reece C.Rmd

@@
 library(tidymodels)
-#library(yardstick)
 library(vip)
+library(knitr)
+library(kableExtra)
 
@@
   metrics(truth = FeCl_dose, estimate = .pred) # Change target variable here
+rmse = model_metrics$.estimate[1]
+r_squared = model_metrics$.estimate[2]
+mae = model_metrics$.estimate[3]
 
-print(model_metrics)
+metrics = data.frame(
+  Metric = c("RMSE", "R-Squared", "MAE"),
+  Value  = c(rmse, r_squared, mae)
+)
 
-ggplot(predictions, aes(x = FeCl_dose, y = .pred)) +
-  geom_point() +
-  geom_abline(intercept = 0, slope = 1) +
-  labs(title = "Random Forest Accuracy",
-       x = "Actual",
-       y = "Predicted") +
-  theme_bw()
+kable(metrics, booktabs = TRUE, digits = 3) %>%
+  kable_styling(position = "center", font_size = 12)
+model_metrics$.estimate
 
-vip(fit_rf)
+#ggplot(predictions, aes(x = FeCl_dose, y = .pred)) +
+#  geom_point() +
+#  geom_abline(intercept = 0, slope = 1) +
+#  labs(title = "Random Forest Accuracy",
+#       x = "Actual",
+#       y = "Predicted") +
+#  theme_bw()
 
+vip(fit_rf, num_features = 20) +
+  geom_col(color = "black", fill = "coral") +
+  labs(
+    title = "Feature Importance (Random Forest)",
+    x = "Importance",
+    y = "Feature"
+  ) +
+  theme_minimal(base_size = 14)
 ```

### 2025-12-11 Data Cleaning.Rmd
Updates to classification methods

@@
 # Model fit to take performance metrics of:
-rf_fit <- rf_fit3
+rf_fit <- rf_fit2
 
@@
 ## Accuracy
-rf_fit |>
+rf_acc <- rf_fit |>
   augment(new_data = df_test) |>
@@
 ## Test confusion matrix
-rf_fit |>
+rf_conf <- rf_fit |>
   augment(new_data = df_test) |>
@@
 
+rf_acc
+rf_conf
+
 ## Notes
 # 34% accuracy for all predictors
+# 39% accuracy for temp + pH + temp/pH lags
 # 46.6% accuracy for just temperature and pH

### 2025-12-11 Presentation DSCI445.Rmd
Updates to classification methods

@@
 library(grid)
+library(parsnip)
+library(discrim)
+```
+
+```{r}
+df_lag_incdec <- read.csv("RWAR_dosing_lag_incdec.csv") |>
+  mutate(Timestamp = ymd(Timestamp))
+df_lag_incdec$FeCl_change <- as.factor(df_lag_incdec$FeCl_change)
+df_lag_incdec$Polymer_change <- as.factor(df_lag_incdec$Polymer_change)
 ```
@@
 
-The drinking water treatment process takes water from a river, lake, reservoir, or other source, and purifies it to have it reach drinking water standards.
+**The drinking water treatment process takes water from a river, lake, reservoir, or other source, and purifies it to have it reach drinking water standards.**
 
@@
 
-**One role of an operator in the water treatment process is to
-look at a wide array of data and make decisions about how to adjust
-chemical doses.**
+**One role of an operator in the water treatment process is to look at a wide array of data and make decisions about how to adjust chemical doses.**
 
@@
 ```{r, fig.width=7, fig.height=2}
-df_lag_incdec <- read.csv("RWAR_dosing_lag_incdec.csv") |>
-  mutate(Timestamp = ymd(Timestamp))
 
@@
 
-# Summary of Models
+# Summary of Modeling
 
@@
 - Coagulant Dose Change Prediction
-   - Multinomial Classification
-   - LDA, QDA
-   - Random Forest
+   - Linear methods: Multinomial, LDA
+   - Nonlinear methods: QDA, Random Forest
 
@@
 
-- Performed best with multiple interactions between Conductivity, Temp, and pH
+- Performed best with multiple interactions between Conductivity, Temp, and pH.
 
@@
 
-- Linear Structure doesn't capture the relations in our predictors, and incorrect predictions could lead to disastorus results if too much or too little chemical is used.
+- Linear structure doesn't capture the relations in our predictors.
 
@@
 
-# Coagulant Dose Change Prediction: Multinomial, LDA/QDA
+# Coagulant Dose Change Prediction: LDA
+
+```{r}
+set.seed(445)
+
+## TRAIN/TEST DATA (2018-2019 and 2020)
+df_full <- df_lag_incdec |> select(-c(FeCl_dose, Polymer_dose))
+df_train <- df_full |> filter(year(Timestamp) <= 2019)
+df_test  <- df_full |> filter(year(Timestamp) == 2020)
+
+## LDA
+lda_spec <- discrim_linear()
+lda_fit1 <- lda_spec |>
+  fit(FeCl_change ~ ., data = df_train |> 
+        select(-c(Polymer_change, Timestamp)))
+lda_fit2 <- lda_spec |>
+  fit(FeCl_change ~ mean_pH + mean_Temp + pH_lag1 + pH_lag2 + Temp_lag1 + Temp_lag2, data = df_train)
+
+# select LDA to take metrics of
+lda_fit <- lda_fit2
+
+# performance metrics
+lda_fit_disp <- lda_fit |> extract_fit_engine()
+lda_conf <- lda_fit |> 
+  augment(new_data = df_test) |>
+  conf_mat(truth = FeCl_change, estimate = .pred_class)
+lda_acc <- lda_fit |>
+  augment(new_data = df_test) |>
+  accuracy(truth = FeCl_change, estimate = .pred_class)
+
+lda_table <- lda_conf$table |> 
+  as.data.frame() |>
+  rename(Predicted = Prediction, True = Truth, Count = Freq)
+lda_table |>
+  tidyr::pivot_wider(names_from = Predicted, values_from = Count) |>
+  knitr::kable("markdown", caption = "LDA Confusion Matrix")
+
+## Notes
+# 39% test accuracy on all predictors
+# 41% test accuracy on temp and pH
+
+```
+
+# Coagulant Dose Change Prediction: QDA
+
+```{r}
+set.seed(445)
+
+## QDA
+qda_spec <- discrim_quad()
+qda_fit1 <- qda_spec |>
+  fit(FeCl_change ~ ., data = df_train |> 
+        select(-c(Polymer_change, Timestamp)))
+qda_fit2 <- qda_spec |>
+  fit(FeCl_change ~ mean_pH + mean_Temp + pH_lag1 + pH_lag2 + Temp_lag1 + Temp_lag2, data = df_train)
+
+# select LDA to take metrics of
+qda_fit <- qda_fit2
+
+# performance metrics
+qda_fit_disp <- qda_fit |> extract_fit_engine()
+qda_conf <- qda_fit |> 
+  augment(new_data = df_test) |>
+  conf_mat(truth = FeCl_change, estimate = .pred_class)
+qda_acc <- qda_fit |>
+  augment(new_data = df_test) |>
+  accuracy(truth = FeCl_change, estimate = .pred_class)
+
+qda_table <- qda_conf$table |> 
+  as.data.frame() |>
+  rename(Predicted = Prediction, True = Truth, Count = Freq)
+qda_table |>
+  tidyr::pivot_wider(names_from = Predicted, values_from = Count) |>
+  knitr::kable("markdown", caption = "QDA Confusion Matrix")
+
+## Notes
+# 39.5% test accuracy on all predictors
+# 43.7% test accuracy on temp and pH
+
+```
+
+# Coagulant Dose Change Prediction: Multinomial Classification
+
+```{r}
+## MULTINOMIAL REGRESSION
+multi_model <- multinom_reg() |>
+  set_engine("nnet") |>
+  set_mode("classification")
+
+multi_recipe1 <- recipe(FeCl_change ~ mean_Temp + Temp_lag1 + Temp_lag2 + mean_pH + pH_lag1 + pH_lag2 + mean_Conductivity + Cond_lag1 + Cond_lag2, data = df_train) |>
+  step_normalize(all_numeric_predictors()) |>
+  step_dummy(all_nominal_predictors())
+
+multi_recipe <- multi_recipe1
+
+multi_workflow <- workflow() |>
+  add_recipe(multi_recipe) |>
+  add_model(multi_model)
+
+multi_fit <- fit(multi_workflow, data = df_train)
+
+df_pred_class <- predict(multi_fit, new_data = df_test, type = "class") |>
+  bind_cols(df_test |> select(FeCl_change))
+df_pred_probs <- predict(multi_fit, new_data = df_test, type = "prob") |>
+  bind_cols(df_test |> select(FeCl_change))
+
+# results
+multi_acc <- metrics(df_pred_class, truth = FeCl_change, estimate = .pred_class) |> 
+  filter(.metric == "accuracy")
+multi_conf <- conf_mat(df_pred_class, truth = FeCl_change, estimate = .pred_class)
+multi_fit_disp <- multi_fit |> extract_fit_engine()
+
+multi_table <- multi_conf$table |> 
+  as.data.frame() |>
+  rename(Predicted = Prediction, True = Truth, Count = Freq)
+multi_table |>
+  tidyr::pivot_wider(names_from = Predicted, values_from = Count) |>
+  knitr::kable("markdown", caption = "Multinomial Regression Confusion Matrix")
+
+## NOTES
+# turbidity is highly variable and does not help
+
+```
 
 # Coagulant Dose Change Prediction: Random Forest
+
+```{r}
+
+set.seed(445)
+
+## RANDOM FOREST
+rf_spec <- rand_forest(mtry = sqrt(.cols()), trees=500, min_n=1) |>
+  set_engine("randomForest", importance = TRUE) |>
+  set_mode("classification")
+
+# Different model fits
+rf_fit1 <- rf_spec |>
+  fit(FeCl_change ~ ., data = df_train |> 
+        select(-c(Polymer_change, Timestamp)))
+rf_fit2 <- rf_spec |>
+  fit(FeCl_change ~ mean_Temp + mean_pH + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2, 
+      data = df_train)
+rf_fit3 <- rf_spec |>
+  fit(FeCl_change ~ mean_pH + mean_Temp, 
+      data = df_train)
+
+# Model fit to take performance metrics of:
+rf_fit <- rf_fit2
+
+# Most important predictors
+#vip(rf_fit)
+
+## results
+rf_acc <- rf_fit |>
+  augment(new_data = df_test) |>
+  accuracy(truth = FeCl_change, estimate = .pred_class)
+rf_conf <- rf_fit |>
+  augment(new_data = df_test) |>
+  conf_mat(truth = FeCl_change, estimate = .pred_class)
+
+rf_table <- rf_conf$table |> 
+  as.data.frame() |>
+  rename(Predicted = Prediction, True = Truth, Count = Freq)
+rf_table |>
+  tidyr::pivot_wider(names_from = Predicted, values_from = Count) |>
+  knitr::kable("markdown", caption = "Random Forest Confusion Matrix")
+
+## Notes
+# 34% accuracy for all predictors
+# 39% accuracy for temp + pH + temp/pH lags
+# 46.6% accuracy for just temperature and pH
+
+```
 

### 2025-12-11 Presentation DSCI445.Rmd
Removed a slide of linear models section

@@
 
-# Performance
-
-- Performed best with multiple interactions between Conductivity, Temp, and pH.
-
-- Still not great predictions, many are overfit and underfit.
-
-- Linear structure doesn't capture the relations in our predictors.
-
 # Coagulant Dose Prediction: Decision Tree
@@
 
+- Experimented with predictor sets and found that pH and temperature (plus the associated lagged variables) were the main effective predictors.
+- Turbidity and conductivity did not improve prediction.
+
+LDA ACCURACY: 43.7%
+
 ```{r}
@@
 
+- Experimented with predictor sets and found that pH and temperature (plus the associated lagged variables) were the main effective predictors.
+- Turbidity and conductivity did not improve prediction.
+
+QDA ACCURACY: 35.7%
+
 ```{r}
@@
 
-# select LDA to take metrics of
+# select QDA to take metrics of
 qda_fit <- qda_fit2
@@
 # Coagulant Dose Change Prediction: Multinomial Classification
+
+- Experimented with predictor sets and found that pH and temperature (plus the associated lagged variables) were the main effective predictors.
+- Turbidity and conductivity did not improve prediction.
+
+MULTINOMIAL ACCURACY: 46.3%
 
@@
 # Coagulant Dose Change Prediction: Random Forest
+
+- Experimented with predictor sets and found that pH and temperature (plus the associated lagged variables) were the main effective predictors.
+- Turbidity and conductivity did not improve prediction.
+
+RANDOM FOREST ACCURACY: 39.2%
 

### 2025-12-11 Presentation DSCI445.Rmd
Final edits.

@@
 
-- Created 6 models with different predictors and combinations of interaction terms.
+- Created 6 models with different predictors and combinations of interaction terms
 
-- Used lag variables for temp, ph, conductivity, turbidity
+- Used lag variables for temp, pH, conductivity, turbidity
 
-- Dropped Turbidity as it was not a good predictor for FeCL dosing.
-  - Conductivity was also dropped eventually due to being similar to Turbidity
+- Dropped Turbidity as it was not a good predictor for FeCL dosing
 
-- Difficult to determine what factors of the dataset are driving the predictions.
+- Difficult to determine what factors of the dataset are driving the predictions
 
-- RMSE and MAE seem to be all over the place dependent on which predictors are selected and the interactions.
+- RMSE and MAE vary widely depending which predictors are selected and the interactions
 
@@
 predicted = predictions$.pred
-
-# Metrics and plots
 
@@
 
-__ performed the best out of models tested.
+```{r, fig.height=3.8, fig.width=7}
 
-**Insert bar graph/brief conclusions**
+# - The model with pH & temperature interactions with conductivity performed the best out of models tested.
+# - Tree-based models performed the worst, likely due to overfitting.
+# - All models diverged from the actual dose by an average of at least 1.4 mg/L.
+
+dosepred_df <- data.frame(Model = c("Random Forest", "Decision Tree", "Cond-pH + Cond-Temp", "Conductivity-Temp", "Temp and pH"),
+                          RMSE = c(2.811, 2.756, 1.672, 1.891, 2.350),
+                          MAE = c(2.283, 2.484, 1.421, 1.516, 1.941))
+
+# convert to long format
+dosepred_long <- dosepred_df |>
+  arrange(RMSE) |>
+  mutate(Model = factor(Model, levels = unique(Model))) |>
+  pivot_longer(
+    cols = c(RMSE, MAE),
+    names_to = "Metric",
+    values_to = "Value"
+  )
+
+# Plot generation
+ggplot(dosepred_long, aes(x = Model, y = Value, fill = Metric)) +
+  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
+  geom_text(
+    aes(label = sprintf("%.3f", Value)), 
+    position = position_dodge(width = 0.8),
+    vjust = -0.5, 
+    size = 2
+  ) +
+  labs(
+    title = "Dose Prediction Model Comparison (RMSE and MAE)",
+    x = "Model",
+    y = "Error Value"
+  ) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+```
 
@@
 
-**Insert bar graph**
+```{r, fig.width=7, fig.height=3}
 
-# Conclusion and Possible Next Steps
+dosechangepred_df <- data.frame(
+  Model = c("LDA", "QDA", "Multinomial", "Random Forest"),
+  Accuracy = c(lda_acc$.estimate, qda_acc$.estimate, 
+               multi_acc$.estimate, rf_acc$.estimate)
+) |> arrange(Accuracy)
+
+ggplot(dosechangepred_df, aes(x = reorder(Model,-Accuracy), y = Accuracy)) +
+  geom_col(fill = "steelblue") +
+  geom_text(aes(label = scales::percent(Accuracy, accuracy = 0.1)),
+            vjust = -0.3, size = 4) +
+  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
+                     limits = c(0, max(dosechangepred_df$Accuracy) * 1.1)) +
+  labs(x = "Model", y = "Accuracy") +
+  theme_minimal()
+
+```
+
+# Conclusion and Next Steps
 

### 2025-12-11 Presentation DSCI445.Rmd
deleted GAM from contents

@@
 - Coagulant Dose Prediction
-   - Linear Regression
+   - Linear Regression Models
    - Tree-based Models
-   - GAM
 - Coagulant Dose Change Prediction

### 2025-12-11 Presentation DSCI445.Rmd
Updated wording on a few sections of presentation.

--- 2025-12-11 Presentation DSCI445.Rmd
@@
 - Training and Testing
-   - Training: 2018-2019
-   - Testing: 2020
+   - Training 2018-2019, Testing 2020
    - Approximately 70/30 split
 - Coagulant Dose Prediction
-   - Linear Regression
-   - Tree-based Models
-   - GAM
+   - Linear Regression Models
+   - Tree-Based Models
 - Coagulant Dose Change Prediction
-   - Linear methods: Multinomial, LDA
-   - Nonlinear methods: QDA, Random Forest
+   - LDA, QDA
+   - Multinomial
+   - Random Forest
 
@@
 
-# RMSE & MAE Table
+# Coagulant Dose Prediction: Linear Regression
 

### 2025-12-11 Presentation DSCI445.Rmd
Final final updates

@@
 
-**One role of an operator in the water treatment process is to look at a wide array of data and make decisions about how to adjust chemical doses.**
+**An operator looks at a wide array of data and makes decisions about chemical dosing.**
+
+**OUR GOAL: Create a model that serves as an "operator," taking raw water data and making predictions of chemical doses.**
 
@@
 img_array = readPNG("Images/treatmentplot2.png")
-grid.raster(img_array)
-```
-
-# Background and Motivation
-
-**OUR GOAL: Create a model that serves the role of an "operator," taking raw water data and making predictions of chemical doses.**
-
-```{r}
-img_array = readPNG("Images/treatmentplot3.png")
 grid.raster(img_array)
@@
 
-- Raw Water Data
+- **Raw Water Data**
    - pH
@@
    - Conductivity
-   - Suspended Grain Size
-   - Alkalinity
-   - Hardness
-
-- Chemical Dosing Data
-   - Ferric Chloride Dose: primary additive
-   - Cationic Polymer Dose: secondary additive
-
-# Data Cleaning
-
-- Dropped Data  
-   - Suspended grain size was not used as it was highly variable
-   - Alkalinity and hardness were not used because of missing data chunks and differing time ranges
-
-- Processing
-   - Daily averages were taken for raw water data to match the dosing data's timestep
-   - Significant outliers (likely caused by equipment malfunction) were removed or interpolated
-   - Dosing and raw water data merged
+   - Suspended Grain Size (dropped)
+   - Alkalinity (dropped)
+   - Hardness (dropped)
+- **Chemical Dosing Data**
+   - Ferric Chloride Dose: primary
+   - Cationic Polymer Dose: secondary
+- Data Cleaning
+   - Removed significant outliers
+   - Took daily averages for raw water data
+   - Merged common dates between datasets
 
@@
 
-Temperature and pH seem to have the strongest relationship to dose, consistent with feedback from treatment plant operators.
-Other data were either highly variable or had minimum availability with large gaps.
-We will focus on the following to predict dose:
+Based on preliminary analyses, we will focus on the following dose predictors:
 
-- pH
-- Temperature
-- Conductivity (lots of noise, weaker)
-- Turbidity (lots of noise, weaker)
+- **pH** (strongest relationship)
+- **Temperature** (strongest relationship)
+- **Conductivity** (some noise, weaker)
+- **Turbidity** (lots of noise, weaker)
 
-Focus: **FeCl Dose**. Polymer dose is not changed significantly in plants.
+Outcome: **FeCl Dose**. (Polymer dose is not changed significantly in plants.)
 
@@
 
-Similar to the stock dataset we've seen in labs and homeworks, our time series dataset needs **lagged variables**.
+Introduced **lagged variables** like in the Stock dataset from the homework.
 
-- For each of the main four parameters:
+- For each of the main four variables:
    - Introduce two lagged variables
@@
 
-We also introduced a **categorical predictor** indicating whether the dose increased, decreased, or stayed the same compared to the day before.
+Introduced a **categorical predictor** indicating whether the dose increased, decreased, or stayed the same compared to the day before.
 
@@
    - Approximately 70/30 split
-- Coagulant Dose Prediction
+- **Coagulant Dose Prediction**
    - Linear Regression Models
    - Tree-Based Models
-- Coagulant Dose Change Prediction
-   - LDA, QDA
-   - Multinomial
-   - Random Forest
+- **Coagulant Dose Change Prediction**
+   - LDA, QDA Classification
+   - Multinomial Classification
+   - Random Forest Classification
 

### 2025-12-15 Paper Draft.Rmd
Moved content over from presentation sections. Set overall structure of report. Added table of contents. Did some editing and adding of explanations.

@@
 ---
-title: "Predictive Modeling of Drinking Water Treatment Process Chemical Doses"
+title: "Prediction of Drinking Water Treatment Process Chemical Doses"
 author: "Dawson Carney, Reece Carmody, Ethan Schilling, Joshua Tobey"
-date: "5 December 2025"
+date: "17 December 2025"
 output: pdf_document
+toc: TRUE
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 ```
@@
 ```{r, include=F}
+# Import required libraries
 library(dplyr)
@@
 library(knitr)
-library (patchwork)
+library(patchwork)
+library(tidyverse)
+library(tidymodels)
+library(rpart.plot)
+library(kableExtra)
+library(vip)
+library(png)
+library(grid)
+library(parsnip)
+library(discrim)
 ```
 
-## Introduction
+```{r}
+## Import primary cleaned dataset
+df_lag_incdec <- read.csv("RWAR_dosing_lag_incdec.csv") |>
+  mutate(Timestamp = ymd(Timestamp))
+df_lag_incdec$FeCl_change <- as.factor(df_lag_incdec$FeCl_change)
+df_lag_incdec$Polymer_change <- as.factor(df_lag_incdec$Polymer_change)
+```
 
-### Background and Motivation
+# Introduction
+
+## Background and Motivation
 
@@
 look at a wide array of these factors and make decisions about how to adjust
-chemical doses.***
+chemical doses.*** The treatment process and operator role are summarized in the figure below.
+
+```{r}
+
+img_array = readPNG("Images/treatmentplot2.png")
+grid.raster(img_array)
+
+```
 
@@
 
-### Project Purpose
+## Project Purpose
 ***The goal of this project is to produce a model that serves the role of a 
@@
 
-## Data Overview and Cleaning
+# Data Overview and Cleaning
 
-### Data Overview
+## Data Overview
 
@@
 
-### Data Cleaning
+## Data Cleaning
 
@@
 
-* pH: No significant outliers that we could see, so we simply day-averaged.  
-* Temperature: There was one region of approximately zero temperature (unrealistic
+* **pH**: No significant outliers that we could see, so we simply day-averaged.
+* **Temperature**: There was one region of approximately zero temperature (unrealistic
 for the reservoir). We got rid of these values and interpolated to the nearest non-zero temperature.
-* Conductivity: Values less than 0.2 and greater than 2 were removed, since they 
+* **Conductivity**: Values less than 0.2 and greater than 2 were removed, since they 
 are unreasonable based on this dataset and the "usual" values.
-* Turbidity: A few exceptionally high values (relative to the usual data values)
+* **Turbidity**: A few exceptionally high values (relative to the usual data values)
 were removed before averaging.
-* Grain Size Information: This data was highly variable, and simply looks like "noise",
+* **Grain Size Information**: This data was highly variable, and simply looks like "noise",
 so did not seem helpful for prediction. We thus did not clean it.
-* Alkalinity and Hardness: These data have different time availability than the other
+* **Alkalinity and Hardness**: These data have different time availability than the other
 datasets but will be inspected. Removed zero-values.
@@
 
-### Plots of Key Treatment Plant Variables
+### Alkalinity and Hardness Data
 
@@
 ```
-See below for dosing compared to key raw water characteristics.
 
-```{r,echo=F}
+### Primary Raw Water Characteristic Data
 
-df_lag_incdec <- read.csv("RWAR_dosing_lag_incdec.csv") |>
-  mutate(Timestamp = ymd(Timestamp))
+```{r, fig.width=7, fig.height=4}
 
-plot_pH <- ggplot(df_lag_incdec, aes(x = Timestamp,
-                           y = mean_pH,
-                           color = mean_pH)) +
-  geom_line() + 
-  scale_color_gradient(low = 'blue', high = 'red') +
+df_lag_incdec_long <- df_lag_incdec |>
+  pivot_longer(
+    cols = starts_with("mean_"),
+    names_to = "Characteristics",
+    values_to = "Value"
+  )
+
+ggplot(df_lag_incdec_long, aes(x = Timestamp, y = Value)) +
+  geom_point(size = 0.5, color="blue") +
+  facet_wrap(~Characteristics, scales = "free_y") +
   theme_minimal() +
-  ggtitle("Average pH")
+  labs(
+    x = "Date",
+    y = "Value",
+    title = "Raw Water Characteristics"
+  )
+
+```
+
+### Coagulant Dose Data
+
+```{r, fig.width=7, fig.height=2}
 
@@
 
+print(plot_polymer)
+
 plot_FeCl <- ggplot(df_lag_incdec, aes(x = Timestamp,
@@
 
-plot_temp <- ggplot(df_lag_incdec, aes(x = Timestamp,
-                           y = mean_Temp,
-                           color = mean_Temp)) +
-  geom_line() + 
-  scale_color_gradient(low = 'blue', high = 'red') +
-  theme_minimal() +
-  ggtitle("Water Temperature")
+print(plot_FeCl)
+```
 
-plot_cond <- ggplot(df_lag_incdec, aes(x = Timestamp,
-                           y = mean_Conductivity,
-                           color = mean_Conductivity)) +
-  geom_line() + 
-  scale_color_gradient(low = 'purple', high = 'yellow') +
-  theme_minimal() +
-  ggtitle("Water Conductivity")
+## Predictor Selection
 
-plot_turb <- ggplot(df_lag_incdec, aes(x = Timestamp,
-                           y = mean_Turbidity,
-                           color = mean_Turbidity)) +
-  geom_line() + 
-  scale_color_gradient(low = 'black', high = 'orange') +
-  theme_minimal() +
-  ggtitle("Water Turbidity")
+Based on preliminary analyses, we will focus on the following dose predictors:
 
-plot_polymer/plot_FeCl
-plot_temp / plot_pH
-plot_cond / plot_turb
-```
+- **pH** (strongest relationship)
+- **Temperature** (strongest relationship)
+- **Conductivity** (some noise, weaker)
+- **Turbidity** (lots of noise, weaker)
+
+Overall, it seems like temperature and pH have the strongest 
+relationships to chemical dosing. This is consistent with feedback from treatment
+plant operators. It is not immediately clear which relationships could
+make effective predictors. Turbidity is highly variable and likely does 
+not have a strong relationship.
+
+Outcome: **FeCl Dose**. (Polymer dose is not changed significantly in plants.)
+
+## Added Variables
+
+We introduced **lagged variables** like in the Stock dataset from the homework.
+
+To capture time-dependence of this dataset, we will introduce lagged variables. 
+For each of the main water parameters, we will introduce two lagged variables, giving
+the water characteristics for the day before and two days before.
+
+We also introduced a **categorical predictor** indicating whether the dose increased, decreased, or stayed the same compared to the day before.
+
+# Modeling
+
+## Summary of Modeling Efforts
+
+### Modeling Priorities
+The main priorities in this modeling effort are *interpretability*, and effective capturing of *nonlinear behavior*.
+Interpretability is important because to trust a model, a treatment plant operator will want to understand the inner workings of how it is making its predictions.
+Nonlinear behavior is important because the chemical process involved in the treatment process are highly complex so a simple linear model will likely not capture these chemical relationships effectively.
+We recognize that these two goals can be at odds, due to the harder-to-interpret nature of highly nonlinear models, so we will attempt to choose models that balance these two objectives.
+
+### Training and Testing Datasets
+- Training and Testing
+   - Training 2018-2019, Testing 2020
+   - Approximately 70/30 split
+
+### Models Used
+The following models will be used for the two main tasks at hand.
+Models that work well with numerical outcomes will be used for coagulant dose prediction.
+Models that are effective for classification will be used for coagulant dose change prediction, since this is a three-category classification problem.
+
+- **Coagulant Dose Prediction**
+   - Linear Regression Models
+   - Tree-Based Models
+   - Generalized Additive Model
+- **Coagulant Dose Change Prediction**
+   - LDA and QDA Classification
+   - Multinomial Classification
+   - Random Forest Classification
+
+## Dose Prediction Models
+
+### Overview of Dose Prediction
 
@@
 
+```
+
+### Linear Regression
+
+**Ethan's section to fill in**
+
+**from presentation:**
+
+- Created 6 models with different predictors and combinations of interaction terms
+
+- Used lag variables for temp, pH, conductivity, turbidity
+
+- Dropped Turbidity as it was not a good predictor for FeCL dosing
+
+- Difficult to determine what factors of the dataset are driving the predictions
+
+- RMSE and MAE vary widely depending which predictors are selected and the interactions
+
+```{r, out.width="80%", out.height="88%"}
+knitr::include_graphics("Images/linear_model_table.png")
+```
+
+### Decision Tree
+The benefit of a tree-based model in this application is that it is easily interpretable.
+It could serve as a decision-making tool for a treatment plant operator:
+
+* The operator sees a change in raw water conditions
+* Obtains a recommended dose based on the decision tree
+* Easily interprets the "why" of the dosing recommendation by following the logic of the decision tree
+* Understands which variables were most important in deciding on a dose
+
+```{r, fig.width=5, fig.height=2.5}
+set.seed(445)
+
+# Decision Tree
+
+# For now just predicting numerical amounts, not classifying
+data = read.csv("RWAR_dosing_lag_incdec.csv")
+data = as.data.frame(data)
+data$FeCl_change = NULL
+data$Polymer_change = NULL
+
+# Set up training & testing data
+
+# Can change these if you want to modify the tree/data
+tree_depth = 5 # tree depth
+min_n = 50 # minimum observations needed to further split
+
+train_data = data %>%
+  filter(year(Timestamp) <= 2019)
+test_data = data %>%
+  filter(year(Timestamp) >= 2020)
+
+# Create the tree
+
+recipe = recipe(FeCl_dose ~ ., data = train_data) %>% # Change target variable here
+  step_rm(Timestamp) %>%
+  step_rm(Polymer_dose) # Remove the other target variable here
+
+tree_spec = decision_tree(tree_depth = tree_depth, min_n = min_n) %>%
+  set_engine("rpart") %>%
+  set_mode("regression")
+
+workflow = workflow() %>%
+  add_recipe(recipe) %>%
+  add_model(tree_spec)
+
+fit_tree = workflow %>%
+  fit(data = train_data)
+
+# Plot the tree
+
+fit_tree %>%
+  extract_fit_engine() %>%
+  rpart.plot(roundint = FALSE)
+
+# Assess accuracy
+
+predictions = fit_tree %>%
+  predict(test_data) %>%
+  bind_cols(test_data)
+
+actual = predictions$`FeCl_dose` # Change target variable here
+predicted = predictions$.pred
+
+rmse = sqrt(mean((actual - predicted)^2))
+r_squared = cor(actual, predicted)^2
+mae = mean(abs(actual-predicted))
+
+metrics = data.frame(
+  Metric = c("RMSE", "R-Squared", "MAE"),
+  Value  = c(rmse, r_squared, mae)
+)
+
+kable(metrics, booktabs = TRUE, digits = 3) %>%
+  kable_styling(position = "center", font_size = 12)
+
+```
+
+### Random Forest
+
+(talk about why random forest was used, parameters and predictors used, process of building, etc.)
+
+```{r, fig.width=5, fig.height=2.1}
+set.seed(445)
+
+# Random Forest
+
+# For now just predicting numerical amounts, not classifying
+data = read.csv("RWAR_dosing_lag_incdec.csv")
+data = as.data.frame(data)
+data$FeCl_change = NULL
+data$Polymer_change = NULL
+
+# Set up training & testing data
+
+# Can change these if you want to modify the tree/data
+trees = 500 # number of trees in the forest
+min_n = 50 # minimum observations needed to further split
+
+train_data = data %>%
+  filter(year(Timestamp) <= 2019)
+test_data = data %>%
+  filter(year(Timestamp) >= 2020)
+
+# Create the tree
+
+recipe = recipe(FeCl_dose ~ ., data = train_data) %>% # Change target variable here
+  step_rm(Timestamp) %>%
+  step_rm(Polymer_dose) # Remove the other target variable here
+
+rf_spec = rand_forest(trees = trees, min_n = min_n) %>%
+  set_engine("ranger", importance = "impurity") %>%
+  set_mode("regression")
+
+workflow = workflow() %>%
+  add_recipe(recipe) %>%
+  add_model(rf_spec)
+
+fit_rf = workflow %>%
+  fit(data = train_data)
+
+# Assess accuracy
+
+predictions = fit_rf %>%
+  predict(test_data) %>%
+  bind_cols(test_data)
+
+actual = predictions$`FeCl_dose` # Change target variable here
+predicted = predictions$.pred
+
+# Metrics and plots
+
+model_metrics = predictions %>%
+  metrics(truth = FeCl_dose, estimate = .pred) # Change target variable here
+rmse = model_metrics$.estimate[1]
+r_squared = model_metrics$.estimate[2]
+mae = model_metrics$.estimate[3]
+
+metrics = data.frame(
+  Metric = c("RMSE", "R-Squared", "MAE"),
+  Value  = c(rmse, r_squared, mae)
+)
+
+vip(fit_rf, num_features = 20) +
+  geom_col(color = "black", fill = "coral") +
+  labs(
+    title = "Feature Importance (Random Forest)",
+    x = "Importance",
+    y = "Feature"
+  ) +
+  theme_minimal(base_size = 9)
+
+kable(metrics, booktabs = TRUE, digits = 3) %>%
+  kable_styling(position = "center", font_size = 12)
+
+```
+
+### Generalized Additive Model (GAM)
+
+**Josh's section to insert his GAM**
+
+### Summary of Dose Prediction Models
+
+```{r, fig.height=3.8, fig.width=7}
+
+# - The model with pH & temperature interactions with conductivity performed the best out of models tested.
+# - Tree-based models performed the worst, likely due to overfitting.
+# - All models diverged from the actual dose by an average of at least 1.4 mg/L.
+
+dosepred_df <- data.frame(Model = c("Random Forest", "Decision Tree", "Cond-pH + Cond-Temp", "Conductivity-Temp", "Temp and pH"),
+                          RMSE = c(2.811, 2.756, 1.672, 1.891, 2.350),
+                          MAE = c(2.283, 2.484, 1.421, 1.516, 1.941))
+
+# convert to long format
+dosepred_long <- dosepred_df |>
+  arrange(RMSE) |>
+  mutate(Model = factor(Model, levels = unique(Model))) |>
+  pivot_longer(
+    cols = c(RMSE, MAE),
+    names_to = "Metric",
+    values_to = "Value"
+  )
+
+# Plot generation
+ggplot(dosepred_long, aes(x = Model, y = Value, fill = Metric)) +
+  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
+  geom_text(
+    aes(label = sprintf("%.3f", Value)), 
+    position = position_dodge(width = 0.8),
+    vjust = -0.5, 
+    size = 2
+  ) +
+  labs(
+    title = "Dose Prediction Model Comparison (RMSE and MAE)",
+    x = "Model",
+    y = "Error Value"
+  ) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+```
+
+## Dose Change Prediction Models
+
+### Overview of Dose Change Prediction
+
+The goal of dose change prediction models is to predict whether coagulant dose increased, decreased, or stayed the same based on raw water characteristics.
+
+As mentioned previously, we added two variables called `FeCl_change` and `Polymer_change` that indicate the chemical dose's direction of change.
+
+This will allow classification methods to be applied. Additionally, if effective,
+this could provide a more helpful tool for operators. Instead of attempting to 
+predict a specific dose, it could give a general direction for an operator to 
+attempt a dose change. This is how plants work in practice: an operator uses 
+some tool (a water test, a reading from a device, etc.), and uses that to make 
+a dose change. They then check important treatment metrics and make any needed
+further adjustments.
+
+The following is a helpful visualization of how dose change varies over time with each of the key raw water characteristics.
+
+```{r, fig.width=7, fig.height=3}
+
 ggplot(df_lag_incdec_long, aes(x = Timestamp, y = Value, color = FeCl_change)) +
@@
     shape = "FeCl_change",
-    title = "Water Quality Characteristics Colored by Dose Change"
+    title = "Raw Water Characteristics Colored by Dose Change"
   )
@@
 
-### Predictor Selection
+Based on this plot, [in essence, we're screwed, but say it diplomatically]
 
-Overall, it seems like temperature and pH have the strongest 
-relationships to chemical dosing. This is consistent with feedback from treatment
-plant operators. It is not immediately clear which relationships could
-make effective predictors. Turbidity is highly variable and likely does 
-not have a strong relationship. Conductivity is similar.
+### LDA
 
-### Addition of Lagged Variables
+- Experimented with predictor sets and found that pH and temperature (plus the associated lagged variables) were the main effective predictors.
+- Turbidity and conductivity did not improve prediction.
 
-To capture time-dependence of this dataset, we will introduce lagged variables. 
-For each of the main water parameters, we will introduce two lagged variables, giving
-the water characteristics for the day before and two days before.
-```{r, echo=F}
-data <- read.csv("RWAR_dosing_cleaned.csv")
-lagged_data <- data |>
-  mutate(Temp_lag1 = lag(mean_Temp, 1), 
-         Temp_lag2 = lag(mean_Temp, 2),
-         pH_lag1 = lag(mean_pH, 1),
-         pH_lag2 = lag(mean_pH, 2),
-         Cond_lag1 = lag(mean_Conductivity, 1),
-         Cond_lag2 = lag(mean_Conductivity, 2),
-         Turb_lag1 = lag(mean_Turbidity, 1),
-         Turb_lag2 = lag(mean_Turbidity, 2))
-# Temp
-lagged_data$Temp_lag1[1] <- lagged_data$Temp_lag1[2]
-lagged_data$Temp_lag2[1:2] <- lagged_data$Temp_lag2[3]
-# pH
-lagged_data$pH_lag1[1] <- lagged_data$pH_lag1[2]
-lagged_data$pH_lag2[1:2] <- lagged_data$pH_lag2[3]
-# Conductivity
-lagged_data$Cond_lag1[1] <- lagged_data$Cond_lag1[2]
-lagged_data$Cond_lag2[1:2] <- lagged_data$Cond_lag2[3]
-# Turbidity
-lagged_data$Turb_lag1[1] <- lagged_data$Turb_lag1[2]
-lagged_data$Turb_lag2[1:2] <- lagged_data$Turb_lag2[3]
+LDA ACCURACY: 43.7%
+
+```{r}
+set.seed(445)
+
+## TRAIN/TEST DATA (2018-2019 and 2020)
+df_full <- df_lag_incdec |> select(-c(FeCl_dose, Polymer_dose))
+df_train <- df_full |> filter(year(Timestamp) <= 2019)
+df_test  <- df_full |> filter(year(Timestamp) == 2020)
+
+## LDA
+lda_spec <- discrim_linear()
+lda_fit1 <- lda_spec |>
+  fit(FeCl_change ~ ., data = df_train |> 
+        select(-c(Polymer_change, Timestamp)))
+lda_fit2 <- lda_spec |>
+  fit(FeCl_change ~ mean_pH + mean_Temp + pH_lag1 + pH_lag2 + Temp_lag1 + Temp_lag2, data = df_train)
+
+# select LDA to take metrics of
+lda_fit <- lda_fit2
+
+# performance metrics
+lda_fit_disp <- lda_fit |> extract_fit_engine()
+lda_conf <- lda_fit |> 
+  augment(new_data = df_test) |>
+  conf_mat(truth = FeCl_change, estimate = .pred_class)
+lda_acc <- lda_fit |>
+  augment(new_data = df_test) |>
+  accuracy(truth = FeCl_change, estimate = .pred_class)
+
+lda_table <- lda_conf$table |> 
+  as.data.frame() |>
+  rename(Predicted = Prediction, True = Truth, Count = Freq)
+lda_table |>
+  tidyr::pivot_wider(names_from = Predicted, values_from = Count) |>
+  knitr::kable("markdown", caption = "LDA Confusion Matrix")
+
+## Notes
+# 39% test accuracy on all predictors
+# 41% test accuracy on temp and pH
+
 ```
 
-## Modeling
+### QDA
 
-### Linear Regression
+- Experimented with predictor sets and found that pH and temperature (plus the associated lagged variables) were the main effective predictors.
+- Turbidity and conductivity did not improve prediction.
+
+QDA ACCURACY: 35.7%
+
+```{r}
+set.seed(445)
+
+## QDA
+qda_spec <- discrim_quad()
+qda_fit1 <- qda_spec |>
+  fit(FeCl_change ~ ., data = df_train |> 
+        select(-c(Polymer_change, Timestamp)))
+qda_fit2 <- qda_spec |>
+  fit(FeCl_change ~ mean_pH + mean_Temp + pH_lag1 + pH_lag2 + Temp_lag1 + Temp_lag2, data = df_train)
+
+# select QDA to take metrics of
+qda_fit <- qda_fit2
+
+# performance metrics
+qda_fit_disp <- qda_fit |> extract_fit_engine()
+qda_conf <- qda_fit |> 
+  augment(new_data = df_test) |>
+  conf_mat(truth = FeCl_change, estimate = .pred_class)
+qda_acc <- qda_fit |>
+  augment(new_data = df_test) |>
+  accuracy(truth = FeCl_change, estimate = .pred_class)
+
+qda_table <- qda_conf$table |> 
+  as.data.frame() |>
+  rename(Predicted = Prediction, True = Truth, Count = Freq)
+qda_table |>
+  tidyr::pivot_wider(names_from = Predicted, values_from = Count) |>
+  knitr::kable("markdown", caption = "QDA Confusion Matrix")
+
+## Notes
+# 39.5% test accuracy on all predictors
+# 43.7% test accuracy on temp and pH
+
+```
+
+### Multinomial Classification
+
+- Experimented with predictor sets and found that pH and temperature (plus the associated lagged variables) were the main effective predictors.
+- Turbidity and conductivity did not improve prediction.
+
+MULTINOMIAL ACCURACY: 46.3%
+
+```{r}
+## MULTINOMIAL REGRESSION
+multi_model <- multinom_reg() |>
+  set_engine("nnet") |>
+  set_mode("classification")
+
+multi_recipe1 <- recipe(FeCl_change ~ mean_Temp + Temp_lag1 + Temp_lag2 + mean_pH + pH_lag1 + pH_lag2 + mean_Conductivity + Cond_lag1 + Cond_lag2, data = df_train) |>
+  step_normalize(all_numeric_predictors()) |>
+  step_dummy(all_nominal_predictors())
+
+multi_recipe <- multi_recipe1
+
+multi_workflow <- workflow() |>
+  add_recipe(multi_recipe) |>
+  add_model(multi_model)
+
+multi_fit <- fit(multi_workflow, data = df_train)
+
+df_pred_class <- predict(multi_fit, new_data = df_test, type = "class") |>
+  bind_cols(df_test |> select(FeCl_change))
+df_pred_probs <- predict(multi_fit, new_data = df_test, type = "prob") |>
+  bind_cols(df_test |> select(FeCl_change))
+
+# results
+multi_acc <- metrics(df_pred_class, truth = FeCl_change, estimate = .pred_class) |> 
+  filter(.metric == "accuracy")
+multi_conf <- conf_mat(df_pred_class, truth = FeCl_change, estimate = .pred_class)
+multi_fit_disp <- multi_fit |> extract_fit_engine()
+
+multi_table <- multi_conf$table |> 
+  as.data.frame() |>
+  rename(Predicted = Prediction, True = Truth, Count = Freq)
+multi_table |>
+  tidyr::pivot_wider(names_from = Predicted, values_from = Count) |>
+  knitr::kable("markdown", caption = "Multinomial Regression Confusion Matrix")
+
+## NOTES
+# turbidity is highly variable and does not help
+
+```
+
+### Random Forest
+
+- Experimented with predictor sets and found that pH and temperature (plus the associated lagged variables) were the main effective predictors.
+- Turbidity and conductivity did not improve prediction.
+
+RANDOM FOREST ACCURACY: 39.2%
 
@@
 
-lm_fit <- lm(Ferric.chloride..mg.L ~ ., data=lagged_data |> select(-c(Timestamp, Cationic.Polimer..mg.L)))
-summary(lm_fit)
+set.seed(445)
+
+## RANDOM FOREST
+rf_spec <- rand_forest(mtry = sqrt(.cols()), trees=500, min_n=1) |>
+  set_engine("randomForest", importance = TRUE) |>
+  set_mode("classification")
+
+# Different model fits
+rf_fit1 <- rf_spec |>
+  fit(FeCl_change ~ ., data = df_train |> 
+        select(-c(Polymer_change, Timestamp)))
+rf_fit2 <- rf_spec |>
+  fit(FeCl_change ~ mean_Temp + mean_pH + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2, 
+      data = df_train)
+rf_fit3 <- rf_spec |>
+  fit(FeCl_change ~ mean_pH + mean_Temp, 
+      data = df_train)
+
+# Model fit to take performance metrics of:
+rf_fit <- rf_fit2
+
+# Most important predictors
+#vip(rf_fit)
+
+## results
+rf_acc <- rf_fit |>
+  augment(new_data = df_test) |>
+  accuracy(truth = FeCl_change, estimate = .pred_class)
+rf_conf <- rf_fit |>
+  augment(new_data = df_test) |>
+  conf_mat(truth = FeCl_change, estimate = .pred_class)
+
+rf_table <- rf_conf$table |> 
+  as.data.frame() |>
+  rename(Predicted = Prediction, True = Truth, Count = Freq)
+rf_table |>
+  tidyr::pivot_wider(names_from = Predicted, values_from = Count) |>
+  knitr::kable("markdown", caption = "Random Forest Confusion Matrix")
+
+## Notes
+# 34% accuracy for all predictors
+# 39% accuracy for temp + pH + temp/pH lags
+# 46.6% accuracy for just temperature and pH
 
 ```
- Interesting initial result: the pH and temperature from two days before are
- the only predictors with low p-values. This makes sense based on the "lag time"
- from water quality change to dosing change.
 
- We will now try models incorporating only some of the predictors.
+### Summary of Dose Change Prediction Models
 
-### [Other Models]
+Overall ineffective, and prone to overfitting (especially random forest).  
+Model predicting no dose change would be more accurate.
 
-### Tree-Based Models
-The benefit of a tree-based model in this application is that it is easily interpretable.
-It could serve as a decision-making tool for a treatment plant operator:
+```{r, fig.width=7, fig.height=3}
 
-* The operator sees a change in raw water conditions
-* Obtains a recommended dose based on the decision tree
-* Easily interprets the "why" of the dosing recommendation by following the logic of the decision tree
-* Understands which variables were most important in deciding on a dose
+dosechangepred_df <- data.frame(
+  Model = c("LDA", "QDA", "Multinomial", "Random Forest"),
+  Accuracy = c(lda_acc$.estimate, qda_acc$.estimate, 
+               multi_acc$.estimate, rf_acc$.estimate)
+) |> arrange(Accuracy)
 
-### Classification of Dose Change
-
-We added two variables called `dose_change_FeCl` and `dose_change_Poly` that indicate
-whether the chemical dose was increased, decreased, or stayed the same. Additionally, we
-added variables indicating change in temperature, pH, turbidity, and conductivity from
-the previous day.
-
-This will allow classification methods to be applied. Additionally, if effective,
-this could provide a more helpful tool for operators. Instead of attempting to 
-predict a specific dose, it could give a general direction for an operator to 
-attempt a dose change. This is how plants work in practice: an operator uses 
-some tool (a water test, a reading from a device, etc.), and uses that to make 
-a dose change. They then check important treatment metrics and make any needed
-further adjustments.
-
-```{r}
+ggplot(dosechangepred_df, aes(x = reorder(Model,-Accuracy), y = Accuracy)) +
+  geom_col(fill = "steelblue") +
+  geom_text(aes(label = scales::percent(Accuracy, accuracy = 0.1)),
+            vjust = -0.3, size = 4) +
+  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
+                     limits = c(0, max(dosechangepred_df$Accuracy) * 1.1)) +
+  labs(x = "Model", y = "Accuracy") +
+  theme_minimal()
 
 ```
+
+# Overview of Model Results
+
+(talk about how the models compared, more just synthesizing than making commentary. save commentary for conclusion)
+
+# Conclusion
+
+- Predicting chemical doses based on these raw water characteristics proved to be ineffective
+- Most likely reasons:
+   - Limited size of dataset
+   - Limited predictor set
+   - Operator-dependent
+   - Complexity of chemical relationships
+   - Other factors (example: reservoir turnover)
+- Operators make decisions based on variety of other factors: chemical waste production, filter performance, lab tests, etc.
+- Incorporate more varieties and time ranges of data in future modeling efforts

### 2025-12-16 Paper Draft.Rmd
Added a description for the random forest

@@
 
-(talk about why random forest was used, parameters and predictors used, process of building, etc.)
+Random forest was used was the logical next step to try to better capture the nonlinear patterns in our data.
+Multiple different tree amounts were attempted, but ultimately they all performed similarly.
+Although the random forest was unable to model the data, we were able to determine which features had the most importance.
+A graph was created to visualize the feature importance for this dataset.
+Notably, pH and the pH lag variables were the most important.
+Temperature and the temperature lag variables followed, which is similar to what we concluded from other models and our exploratory analysis.
 

### 2025-12-16 Paper Draft.Rmd
Added linear regression sections and a few libraries at the top

@@
 library(discrim)
+library(gt)
+library(Metrics)
 ```
@@
 
-**Ethan's section to fill in**
+To model FeCl dosing using linear regression, we fit six different linear regression models that varied in their choice of predictors and inclusion of differing interaction terms.
+The models incorporated current and lagged values of temperature, pH, conductivity, and turbidity, allowing us to account for delayed effects in the system from previous days.
+Several interactions were tested, including temperature-pH and temperature-conductivity, to capture potential nonlinear relationships between water quality variables.
+Turbidity was ultimately dropped from the final models, as it consistently preformed poorly as a predictor for FeCl dosing.
+```{r}
 
-**from presentation:**
+RWAR_dosing <- df_lag_incdec %>% 
+  mutate(Timestamp = ymd(Timestamp),
+         Year = year(Timestamp))
 
-- Created 6 models with different predictors and combinations of interaction terms
+train <- RWAR_dosing %>% filter(Year %in% c(2018, 2019))
+test  <- RWAR_dosing %>% filter(Year == 2020)
 
-- Used lag variables for temp, pH, conductivity, turbidity
+# All predictors
 
-- Dropped Turbidity as it was not a good predictor for FeCL dosing
+model0 <- lm(FeCl_dose ~ mean_Temp + mean_pH + mean_Conductivity + mean_Turbidity + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2 + Turb_lag1 + Turb_lag2, data = train)
 
-- Difficult to determine what factors of the dataset are driving the predictions
+test$pred_model0 <- predict(model0, newdata = test)
 
-- RMSE and MAE vary widely depending which predictors are selected and the interactions
+# 1 Interaction Term
+
+model1 = lm(FeCl_dose ~ mean_Temp * mean_pH + mean_Turbidity + Temp_lag1  + Temp_lag2 + Turb_lag1 +     Turb_lag2,data = train)
+
+test$pred_model1 <- predict(model1, newdata = test)
+
+# 2 Interaction Terms
+
+model2 = lm(FeCl_dose ~ mean_Temp * mean_pH + mean_Conductivity * mean_Turbidity +
+                Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2, data = train)
+
+test$pred_model2 <- predict(model2, newdata = test)
+
+model3 = lm(FeCl_dose ~ mean_Temp * mean_Conductivity + Temp_lag1 + Temp_lag2 + Cond_lag1 + Cond_lag2,                 data = train)
+
+test$pred_model3 <- predict(model3, newdata = test)
+
+# 2 Interactions: Cond and PH and Temp and Cond
+
+model4 = lm(FeCl_dose ~ mean_Conductivity * mean_pH + mean_Temp * mean_Conductivity +
+                Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2,
+              data = train)
+
+test$pred_model4 <- predict(model4, newdata = test)
+
+model5 = lm(FeCl_dose ~ mean_Temp + mean_pH + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2,
+              data = train)
+
+test$pred_model5 <- predict(model5, newdata = test)
+
+```
 
 ```{r, out.width="80%", out.height="88%"}
-knitr::include_graphics("Images/linear_model_table.png")
+# Creating Table and Top Model Graphs
+
+models <- list(
+
+  model0 = lm(FeCl_dose ~ mean_Temp + mean_pH + mean_Conductivity + mean_Turbidity +
+                Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2 +
+                Turb_lag1 + Turb_lag2, data = train),
+
+  model1 = lm(FeCl_dose ~ mean_Temp * mean_pH + mean_Turbidity + Temp_lag1  + Temp_lag2 + Turb_lag1 +
+                Turb_lag2,data = train),
+
+  model2 = lm(FeCl_dose ~ mean_Temp * mean_pH + mean_Conductivity * mean_Turbidity +
+                Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2, data = train),
+
+  model3 = lm(FeCl_dose ~ mean_Temp * mean_Conductivity + Temp_lag1 + Temp_lag2 + Cond_lag1 + Cond_lag2,                 data = train),
+
+  model4 = lm(FeCl_dose ~ mean_Conductivity * mean_pH + mean_Temp * mean_Conductivity +
+                Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2,
+              data = train),
+
+  model5 = lm(FeCl_dose ~ mean_Temp + mean_pH + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2,
+              data = train)
+)
+
+descriptions <- c(
+  "Main effects only",
+  "Temp × pH interaction",
+  "Temp × pH + Cond × Turb interactions",
+  "Temp × Cond interaction",
+  "Cond × pH + Temp × Cond interactions",
+  "Only Temp and pH"
+)
+
+performance <- data.frame(Model = character(),
+                          RMSE = numeric(),
+                          MAE = numeric(),
+                          Description = character(),
+                          stringsAsFactors = FALSE)
+
+for(i in 1:length(models)) {
+  pred <- predict(models[[i]], newdata = test)
+  performance <- rbind(performance,
+                       data.frame(Model = paste0("Model ", i-1),
+                                  RMSE = round(Metrics::rmse(test$FeCl_dose, pred), 3),
+                                  MAE  = round(Metrics::mae(test$FeCl_dose, pred), 3),
+                                  Description = descriptions[i],
+                                  stringsAsFactors = FALSE))
+
+}
+  performance <- performance %>% arrange(RMSE)
+
+best_rmse <- min(performance$RMSE)
+
+kable_table <- performance %>%
+  kable("html", caption = "Model Performance Summary") %>%
+  kable_styling(full_width = F, position = "center") %>%
+  row_spec(which(performance$RMSE == best_rmse), background = "#cfe2f3")
+
+kable_table
+
+p3 <- ggplot(test, aes(x = FeCl_dose, y = pred_model3)) +
+  geom_point(color = "blue") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
+  labs(title = "Model 3: Predicted vs Actual FeCl_dose",
+       x = "Actual FeCl_dose",
+       y = "Predicted FeCl_dose") +
+  theme_minimal()
+
+p4 <- ggplot(test, aes(x = FeCl_dose, y = pred_model4)) +
+  geom_point(color = "green") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
+  labs(title = "Model 4: Predicted vs Actual FeCl_dose",
+       x = "Actual FeCl_dose",
+       y = "Predicted FeCl_dose") +
+  theme_minimal()
+
+p3
+p4
 ```
+
+ Model performance was evaluated using RMSE and MAE on a 2020 test set, and varied substantially depending on the selected predictors and interactions.
+There is evidence of other factors influencing the dosing as many of the predictors in our data set proved to be not be as influential as we originally had suspected.
+This is reflected through the model performance in which the model with temperature-pH and temperature-conductivity being the best, but its predictions still deviated considerably from the actual FeCl dosing values.
+As seen in the plots and table above, none of the models performed particularly well, and there was a ton of variance within each model and how well it predicted.
 

### 2025-12-16 Paper Draft.Rmd
Looked very weird on the github hopefully i fixed it

@@
 ### Linear Regression
-
 To model FeCl dosing using linear regression, we fit six different linear regression models that varied in their choice of predictors and inclusion of differing interaction terms.
@@
 Turbidity was ultimately dropped from the final models, as it consistently preformed poorly as a predictor for FeCl dosing.
+
 ```{r}

### 2025-12-17 Paper Draft.Rmd
Incorporated Josh's GAM. Wrote conclusion from bullet points that I presented in slide. Re-formatted table in Ethan's section. Added descriptions to all classification sections and some of the dose prediction sections. Added description to exploratory analysis section.

@@
 library(Metrics)
+library(mgcv)
 ```
@@
 based on other information.
+We will attempt to create both a predictive model for predicting the chemical dose, and a model to predict dose change (increase, decrease, stay same), both based on raw water characteristics.
 
@@
     + pH
-    + Temperature (of the water)
-    + Turbidity
-    + Suspended Grain Size Information
-    + Alkalinity
-    + Hardness
+    + Temperature
+    + Turbidity ("cloudiness" of the water)
+    + Suspended Grain Size Distributions
+    + Alkalinity (resistance of water to changes in pH)
+    + Hardness (mineral content of water)
 
@@
 
-* **pH**: No significant outliers that we could see, so we simply day-averaged.
+* **pH**: No significant outliers were present that we could see, so we simply day-averaged.
 * **Temperature**: There was one region of approximately zero temperature (unrealistic
@@
 
-Some of this cleaned data is presented below.
-
 ### Alkalinity and Hardness Data
 
-Alkalinity and hardness tend to stay fairly steady over time.
+While alkalinity and hardness display seasonal trends that could be helpful for prediction, the available time ranges of the data did not align with other datasets (namely, dosing and other raw water characteristics), so they will not be used in this analysis.
+However, they are pictured below.
+Note the gap in data availability from mid-2017 to mid-2018.
 
@@
 
+The main four raw water characteristics used in this analysis are pictured below.
+Turbidity displays the highest levels of fluctuation and noise, followed by conductivity.
+pH and temperature tend to display fairly consistent annual cycles without significant noise.
+According to the treatment operator who provided the data, temperature and pH are important for the treatment process, so less noise could be advantageous for modeling.
+
 ```{r, fig.width=7, fig.height=4}
@@
 ### Coagulant Dose Data
+
+The following are the dosing over time of Ferric Chloride (primary coagulant) and the polymer coagulant.
+Note the long periods of no dosing change, some periods of sharp dips and jumps, and periods of steady increase and decrease.
+This is how operators alter coagulant doses day-to-day.
+This gives a picture of some of the challenges that may be present in trying to predict this coagulant dosage based on water characteristics.
 
@@
 
-Based on preliminary analyses, we will focus on the following dose predictors:
+Based on preliminary analyses, we will focus on the following dose predictors, which have the greatest time availability and fewest data gaps out of our available raw water data:
 
@@
 
-Outcome: **FeCl Dose**. (Polymer dose is not changed significantly in plants.)
+We will focus on **FeCl Dose** as the outcome variable for our modeling efforts.
+Polymer dose is not changed significantly in plants, as it has less direct impact on the coagulation performance and serves as a secondary "helper" in particle formation to the ferric chloride.
 
@@
 
-We introduced **lagged variables** like in the Stock dataset from the homework.
-
-To capture time-dependence of this dataset, we will introduce lagged variables. 
-For each of the main water parameters, we will introduce two lagged variables, giving
-the water characteristics for the day before and two days before.
+To capture some of the time-dependence likely present in our dataset, we introduced **lagged variables** for each of our main raw water characteristics.
+We introduced two lagged variables, representing the water characteristics for the day before and two days before.
 
 We also introduced a **categorical predictor** indicating whether the dose increased, decreased, or stayed the same compared to the day before.
+This will be used for dose change prediction modeling (classification), which is detailed below.
 
@@
 ### Training and Testing Datasets
-- Training and Testing
-   - Training 2018-2019, Testing 2020
-   - Approximately 70/30 split
+Our total dataset spans 2018-2020, so we will use data from 2018-2019 as the training dataset, and 2020 as the testing dataset.
+This accomplishes approximately a 70/30 training/testing split for our available data.
+
+We recognize that there may be limitations to this train/test split, since 2020 was both the height of the COVID-19 pandemic, and a year of above-average numbers of forest fires.
+The former of these could have impacted the funding and staffing of the water treatment plant, along with consumptive use patterns in the regions serviced by the plant.
+The fires could have directly impacted water quality with more dissolved ash in the water supply.
+
+These could have unforeseen impacts, but it seems upon preliminary assessment that these factors should not impact plant performance significantly.
+Water treatment is critical infrastructure, so should not be substantially impacted by shutdowns.
+Additionally, it is already only staffed by a few people at a time.
+Shifting consumptive use patterns would not impact reservoir water quality.
+
+Reservoirs also tend to be fairly stable water sources, in terms of quality, so the additional ash that may have come from the fires is likely negligible.
+We observed some sparse datasets available, including dissolved organic carbon and total organic carbon measurements, and 2020 was not significantly different from 2018 or 2019 in this way, which could be a helpful indicator.
+
+With this in mind, this split will be maintained. For future efforts, with more data, different splits will be tested, but with this available data, this seems like the best approach.
 
@@
 Models that work well with numerical outcomes will be used for coagulant dose prediction.
-Models that are effective for classification will be used for coagulant dose change prediction, since this is a three-category classification problem.
+Models that are effective for classification will be used for coagulant dose change prediction, since this is a three-category classification problem (increase, decrease, stay same).
 
@@
    - LDA and QDA Classification
-   - Multinomial Classification
+   - Multinomial Logistic Regression
    - Random Forest Classification
+
+Test MAE and RMSE will be used to assess dose prediction models, with best model chosen based on RMSE. Test accuracy will be used to assess dose change prediction models.
 
@@
 
-```{r}
+**The goal of the dose prediction modeling is to predict FeCl dosage from the raw water characteristics.** The following plot helps illustrate the data that will be used for the dose prediction model.
+This plot does not reveal any clear trends in dosage relative to any of the water treatment variables (dose varying with a water quality metric), indicating this could be a challenging modeling effort.
+
+```{r, fig.width=7, fig.height=4}
 
@@
 Turbidity was ultimately dropped from the final models, as it consistently preformed poorly as a predictor for FeCl dosing.
+
+Displayed are both a summary of the fitted regression models, and an example plot of the actual versus predicted dosing, presented for the best-performing linear model, Model 3.
+This better illustrates how far the predictions are from the actual dosing.
 
@@
 
-```
-
-```{r, out.width="80%", out.height="88%"}
-# Creating Table and Top Model Graphs
+## Create graphs and get model results
 
@@
 
-kable_table <- performance %>%
-  kable("html", caption = "Model Performance Summary") %>%
-  kable_styling(full_width = F, position = "center") %>%
-  row_spec(which(performance$RMSE == best_rmse), background = "#cfe2f3")
-
-kable_table
+kable_table_pdf <- performance %>%
+  # Use format = "latex" for PDF output
+  kable("latex",
+        booktabs = TRUE)
 
@@
 
+kable_table_pdf
+
 p3
-p4
+
 ```
@@
 
-```{r, fig.width=5, fig.height=2.5}
+A decision tree was fit with a depth of 5, and 50 observations required to split.
+These parameters were varied but it did not significantly improve prediction.
+The model split first based on pH, which was mentioned by treatment plant operators as an important predictor.
+However, the next splits were on lagged conductivity and turbidity measurements.
+This was surprising, and could likely be due to the noise present in the data.
+(Possible result of overfitting.) This model performed more poorly than the linear models.
+See below for model performance metrics and a visualization of the final decision tree.
+
+```{r, fig.width=8, fig.height=4}
 set.seed(445)
@@
 Random forest was used was the logical next step to try to better capture the nonlinear patterns in our data.
-Multiple different tree amounts were attempted, but ultimately they all performed similarly.
+Multiple different tree amounts were attempted, but ultimately they all performed similarly, and worse than the linear models.
 Although the random forest was unable to model the data, we were able to determine which features had the most importance.
@@
 
-**Josh's section to insert his GAM**
+Next, a GAM was fitted to the modeled data.
+This model uses splines to model nonlinear relationships between predictor and response.
+The hope would be that this flexible model form could help capture the nonlinearity present in these chemical processes, for effective prediction.
+See below for a plot of how the GAM performed on the test dataset.
+
+```{r}
+
+train <- df_lag_incdec |> filter(year(Timestamp) <= 2019)
+test  <- df_lag_incdec |> filter(year(Timestamp) == 2020)
+
+gam_spec <- gen_additive_mod() |> 
+  set_engine("mgcv") |> 
+  set_mode("regression")
+gam_rec <- recipe(FeCl_dose ~ Polymer_dose + mean_Temp + mean_pH + mean_Conductivity +
+                 mean_Turbidity + Temp_lag1 + Temp_lag2 +
+                 pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2 +
+                 Turb_lag1 + Turb_lag2, data = train)
+
+gam_wf <- workflow() |>
+  add_model(gam_spec) |>
+  add_recipe(gam_rec)
+
+gam_fit <- gam(
+  FeCl_dose ~ s(Polymer_dose) + s(mean_Temp) + s(mean_pH) + s(mean_Conductivity) +
+               s(mean_Turbidity) + s(Temp_lag1) + s(Temp_lag2) +
+               s(pH_lag1) + s(pH_lag2) + s(Cond_lag1) + s(Cond_lag2) +
+               s(Turb_lag1) + s(Turb_lag2),
+  data = train
+)
+
+# gam_fit |> tidy()
+# gam_fit |> glance()
+
+pred_df <- test %>% 
+  mutate(pred = predict(gam_fit, newdata = test))
+
+gam_rmse <- yardstick::rmse(pred_df, truth = FeCl_dose, estimate = pred)
+gam_mae <- yardstick::mae(pred_df, truth = FeCl_dose, estimate = pred)
+
+ggplot(pred_df, aes(x = Timestamp)) +
+  geom_line(aes(y = FeCl_dose, color = "Actual")) + 
+  geom_line(aes(y = pred, color = "Predicted")) + 
+  scale_color_manual(
+    name = "FeCl Dose",
+    values = c("Actual" = "orange", "Predicted" = "skyblue")) +
+  labs(
+    y = "FeCl Dose (mg/L)" # Add a proper Y-axis label
+  ) +
+  theme_minimal() +
+  ggtitle("FeCl Dosing Predictions by GAM on Test Dataset")
+
+```
+
+Overall, the GAM was able to vaguely follow trends in dosing for the testing dataset, but did not perform overly well, with MAE and RMSE of 1.8 mg/L and 2.3 mg/L, respectively.
+The model flexibility did not turn out to yield the hoped-for results, though it performed similarly to the standard linear models.
+It did out-perform both of the tree-based models.
 
@@
 
-```{r, fig.height=3.8, fig.width=7}
+Overall, none of the dose prediction models performed particularly well or were able to follow the changes in chemical dose present in the modeling.
+The top three models by RMSE were the models involving interactions between conductivity and temperature/pH, and the GAM.
+However, the best-performing models still differed from the actual dosing by an average of at least 1.5 mg/L based on MAE, which is a significant dosing deviation that could hinder proper functioning of a treatment plant if implemented.
+
+```{r, fig.height=4.2, fig.width=7}
 
@@
 
-dosepred_df <- data.frame(Model = c("Random Forest", "Decision Tree", "Cond-pH + Cond-Temp", "Conductivity-Temp", "Temp and pH"),
-                          RMSE = c(2.811, 2.756, 1.672, 1.891, 2.350),
-                          MAE = c(2.283, 2.484, 1.421, 1.516, 1.941))
+dosepred_df <- data.frame(Model = c("Random Forest", "Decision Tree", "Cond-pH + Cond-Temp", "Conductivity-Temp", "Temp and pH", "GAM"),
+                          RMSE = c(2.811, 2.756, 1.672, 1.891, 2.350, gam_rmse$.estimate),
+                          MAE = c(2.283, 2.484, 1.421, 1.516, 1.941, gam_mae$.estimate))
 
@@
 
+Next, we will move on to dose change prediction, and see how its performance compares to this model estimation of exact doses.
+
 ## Dose Change Prediction Models
@@
 
-The goal of dose change prediction models is to predict whether coagulant dose increased, decreased, or stayed the same based on raw water characteristics.
+**The goal of dose change prediction models is to predict whether coagulant dose increased, decreased, or stayed the same based on raw water characteristics.**
 
@@
 The following is a helpful visualization of how dose change varies over time with each of the key raw water characteristics.
+As with dose prediction, there are not clear patterns of when dose is increasing, decreasing, or staying the same relative to water quality metrics.
+This will most likely make the data difficult to separate via classification methods.
 
-```{r, fig.width=7, fig.height=3}
+```{r, fig.width=7, fig.height=4}
 
@@
 
-Based on this plot, [in essence, we're screwed, but say it diplomatically]
+For each of the following classification models, different combinations of predictor variable sets were experimented with.
+It was found that pH and temperature, with their associated lag variables, produced the highest test accuracies.
+Turbidity and conductivity did not improve test accuracy, so were not included in the final model that is represented by the confusion matrices and accuracies below.
 
-### LDA
+### Linear Discriminant Analysis (LDA)
 
-- Experimented with predictor sets and found that pH and temperature (plus the associated lagged variables) were the main effective predictors.
-- Turbidity and conductivity did not improve prediction.
+LDA uses linear combinations of features to perform dimensionality reduction and create better separation between classes of data.
+Thus, we hope that for this hard-to-separate data, the dimension reduction could assist, though there are not many dimensions to start with, so we could foresee it still being highly limited.
 
-LDA ACCURACY: 43.7%
+TEST ACCURACY: 43.7%
 
@@
   augment(new_data = df_test) |>
-  accuracy(truth = FeCl_change, estimate = .pred_class)
+  yardstick::accuracy(truth = FeCl_change, estimate = .pred_class)
 
@@
 
-### QDA
+### Quadratic Discriminant Analysis (QDA)
 
-- Experimented with predictor sets and found that pH and temperature (plus the associated lagged variables) were the main effective predictors.
-- Turbidity and conductivity did not improve prediction.
+QDA is similar to LDA but incorporates quadratic terms, not just linear terms. This could potentially help address the nonlinearity in our dataset.
 
-QDA ACCURACY: 35.7%
+However, based on test accuracy and the confusion matrix, the QDA model performed just as poorly as the LDA, and slightly worse. See results below.
+
+TEST ACCURACY: 35.7%
 
@@
 
-### Multinomial Classification
+### Multinomial Logistic Regression
 
-- Experimented with predictor sets and found that pH and temperature (plus the associated lagged variables) were the main effective predictors.
-- Turbidity and conductivity did not improve prediction.
+Multinomial logistic regression is an extension of logistic regression that allows for multi-category classification, by setting one category as a "reference category" and calculating probabilities of other categories relative to that category.
+Overall, this model tended to predict the best on the test dataset.
+Looking at the confusion matrix, it tended to lean towards predicting "same" when it made an error, which is more acceptable than choosing the opposite direction of dose change, though it made that error as well.
+However, it is still not an exceptional model by any means.
+See below for confusion matrix and accuracy.
 
-MULTINOMIAL ACCURACY: 46.3%
+TEST ACCURACY: 46.0%
 
@@
 
-multi_recipe1 <- recipe(FeCl_change ~ mean_Temp + Temp_lag1 + Temp_lag2 + mean_pH + pH_lag1 + pH_lag2 + mean_Conductivity + Cond_lag1 + Cond_lag2, data = df_train) |>
+multi_recipe1 <- recipe(FeCl_change ~ mean_Temp + Temp_lag1 + Temp_lag2 + mean_pH + pH_lag1 + pH_lag2, data = df_train) |>
   step_normalize(all_numeric_predictors()) |>
@@
 
-## NOTES
-# turbidity is highly variable and does not help
-
 ```
@@
 
-- Experimented with predictor sets and found that pH and temperature (plus the associated lagged variables) were the main effective predictors.
-- Turbidity and conductivity did not improve prediction.
+Finally, a Random Forest model was fit to this data.
+The standard number $\sqrt{P}$ predictors were sampled at each split, and 500 trees were grown.
+The number of trees was varied but it did not change prediction accuracy.
 
-RANDOM FOREST ACCURACY: 39.2%
+This model performed second-worst out of the classification models.
+We believe that this is mainly due to overfitting.
+Random forest models tend to overfit on noise-heavy data, so this model performed perfectly on the training dataset and very poorly on the test dataset.
+See accuracy and confusion matrix, below.
+
+TEST ACCURACY: 39.2%
 
@@
   augment(new_data = df_test) |>
-  accuracy(truth = FeCl_change, estimate = .pred_class)
+  yardstick::accuracy(truth = FeCl_change, estimate = .pred_class)
 rf_conf <- rf_fit |>
@@
 
-## Notes
-# 34% accuracy for all predictors
-# 39% accuracy for temp + pH + temp/pH lags
-# 46.6% accuracy for just temperature and pH
-
 ```
@@
 
-Overall ineffective, and prone to overfitting (especially random forest).  
-Model predicting no dose change would be more accurate.
+These models were ineffective at predicting the direction of dose change.
+This aligns with initial expectations, since the data was shown to be extremely scattered and difficult to separate in the time series plots shown at the start of this section.
+All models performed very poorly, so it is hard to compare which models may have performed "best." At the surface, multinomial performed best, likely due to it using linear combinations of predictors (which aligns with the most effective dose prediction methods).
+Random forest was the most prone to overfitting and worst-performing, which aligns with its tendency to overfit on noise-heavy data.
+Model accuracies are summarized below.
 
-```{r, fig.width=7, fig.height=3}
+```{r, fig.width=7, fig.height=3.2}
 
@@
   labs(x = "Model", y = "Accuracy") +
-  theme_minimal()
+  theme_minimal() +
+  ggtitle("Dose Change Classification Model Comparison")
 
@@
 
-# Overview of Model Results
-
-(talk about how the models compared, more just synthesizing than making commentary. save commentary for conclusion)
-
 # Conclusion
 
-- Predicting chemical doses based on these raw water characteristics proved to be ineffective
-- Most likely reasons:
-   - Limited size of dataset
-   - Limited predictor set
-   - Operator-dependent
-   - Complexity of chemical relationships
-   - Other factors (example: reservoir turnover)
-- Operators make decisions based on variety of other factors: chemical waste production, filter performance, lab tests, etc.
-- Incorporate more varieties and time ranges of data in future modeling efforts
+Overall, predicting chemical doses based on these raw water characteristics proved to be ineffective.
+
+The dose prediction models produced high RMSE's, either being unable to capture the complexity of the dosing, or overfitting to the noise-filled signals of the input data.
+The dose change prediction models, as predicted by observation of the timeseries plots, yielded extremely low accuracies, due to this being a dataset not separable by any one of these four input water quality characteristics.
+
+Some of the most likely reasons for this are as follows:
+
+* **Limited dataset size**: our dataset was limited in time, forcing us to use a small training and testing dataset that limited our ability to choose other years for testing.
+This likely produced errors specific to the year that we chose for testing, which could be averted with more data.
+* **Limited predictor set**: These four water quality variables were not effective for predicting chemical doses.
+However, there are other water quality parameters, and other treatment plant performance characteristics such as filter performance, that are used to make dose decisions.
+Access to these data could significantly improve model performance.
+* **Operator-dependence**: Whether doses even change at all is largely dependent on operators.
+This plant is one of the more advanced in the nation, but some plants leave their dose the same for years at a time.
+So, our model is in part simulating human behavior, which is much more difficult than a basic model can capture.
+* **Complexity of chemical relationships**: The drinking water treatment process is an enormous chemical reaction with a large number of interconnected processes and chemical interactions, so our models likely failed in part to not being able to fully capture these relationships.
+* **Other factors**: Other miscellaneous factors can impact water quality, such as reservoir turnover, which is when a reservoir's water level drops, and water quality changes drastically over the course of a few days at a specific time of year.
+This is hard to predict, and our model could not capture this.
+
+Ultimately, water treatment is both a science and an art, and operators make decisions based on variety of inputs: chemical waste production, filter performance, or lab tests, to name a few.
+Future modeling efforts should incorporate more varieties and time ranges of data to attempt to capture some of these complexities.

### 2025-12-17 FinalPaper_DSCI445_Group4.Rmd
Final overview of report and minor tweaks

--- 2025-12-17 Paper Draft.Rmd
@@
 ---
-
+CAUTION: Please uncomment any of the install.packages that aren't already installed.
 ```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo = FALSE)
+#install.packages("zoo")
+#install.packages("purrr")
+#install.packages("kableExtra")
+#install.packages("vip")
+#install.packages("png")
+#install.packages("grid")
+#install.packages("discrim")
+#install.packages("gt")
+#install.packages("Metrics")
+#install.packages("ranger")
+#install.packages("randomForest")
+
 ```
@@
 library(mgcv)
+library(ranger)
+library(randomForest)
 ```
@@
 test  <- RWAR_dosing %>% filter(Year == 2020)
-
-# All predictors
-
+```
+All predictors:
+```{r}
 model0 <- lm(FeCl_dose ~ mean_Temp + mean_pH + mean_Conductivity + mean_Turbidity + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2 + Turb_lag1 + Turb_lag2, data = train)
@@
 test$pred_model5 <- predict(model5, newdata = test)
-
-## Create graphs and get model results
-
+```
+
+Generate model results
+```{r}
 models <- list(
@@
   performance <- performance %>% arrange(RMSE)
-
+performance
+```
+Selecting the best RMSE:
+```{r}
 best_rmse <- min(performance$RMSE)
-
+best_rmse
+```
+Display Plots of models' performance on test data:
+```{r}
 kable_table_pdf <- performance %>%
@@
   theme_minimal()
-
+p3
+```
+```{r}
 p4 <- ggplot(test, aes(x = FeCl_dose, y = pred_model4)) +
@@
 
-p3
+p4
 
@@
 
-Random forest was used was the logical next step to try to better capture the nonlinear patterns in our data.
+Random forest was used and was the next logical step to try to better capture any nonlinear patterns in our data.
 Multiple different tree amounts were attempted, but ultimately they all performed similarly, and worse than the linear models.
@@
 
-# gam_fit |> tidy()
-# gam_fit |> glance()
-
+gam_fit |> tidy()
+gam_fit |> glance()
+```
+```{r}
 pred_df <- test %>% 
@@
 It did out-perform both of the tree-based models.
+
+There does seem to be a faint trend between the model performance and the observations in a postive slope direction.
+This model may perform better if we had more years at our disposal to train the GAM on, but how much better is highly speculative because of the variation of predictions from the model in contrast to the observations.
+More data could be what this model needs.
 

### 2025-12-17 FinalPaper_DSCI445_Group4.Rmd
Edits made to output redundancies

@@
 ---
-CAUTION: Please uncomment any of the install.packages that aren't already installed.
 ```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo = FALSE)
-#install.packages("zoo")
-#install.packages("purrr")
-#install.packages("kableExtra")
-#install.packages("vip")
-#install.packages("png")
-#install.packages("grid")
-#install.packages("discrim")
-#install.packages("gt")
-#install.packages("Metrics")
-#install.packages("ranger")
-#install.packages("randomForest")
 
@@
   performance <- performance %>% arrange(RMSE)
-performance
 ```
@@
 )
-
-gam_fit |> tidy()
-gam_fit |> glance()
 ```

### 2025-12-17 FinalPaper_DSCI445_Group4.Rmd
final changes to paper

@@
 ```
-All predictors:
+
 ```{r}
@@
 
-Generate model results
 ```{r}
@@
 ```
-Selecting the best RMSE:
+
 ```{r}
 best_rmse <- min(performance$RMSE)
-best_rmse
-```
-Display Plots of models' performance on test data:
+```
+
 ```{r}
